### 2025-04-04 manuscript/paper.qmd
updated manuscript

---
title: "Introduction for a BART-Based Model to Predict Chlorophyll-a from Satellite Reflectance"
author: Erdem Karaköylü
date: "2025-04-03"
format:
  pdf:
    documentclass: agujournal2019
    template: templates/agujournaltemplate.tex
    keep-tex: true
bibliography: references.bib
---

### Introduction

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color missions have long relied on empirical spectral algorithms to estimate near-surface chlorophyll-a (Chl-a) concentrations from water‐leaving radiance. NASA’s standard approach since the SeaWiFS era has been the family of “OCx” band‐ratio algorithms, which use blue-to-green reflectance ratios fit to in situ Chl data via polynomial regression [@OReilly2000]. These O’Reilly et al. formulations (e.g., OC4 for SeaWiFS, OC3 for MODIS) became the default global chlorophyll algorithms for multiple sensors due to their relative simplicity and consistency in Case-1 waters. In 2014, NASA introduced the Ocean Color Index (OCI) algorithm, a hybrid empirical model that blends a three‐band reflectance difference method for low‐Chl regimes with the traditional OCx ratio approach at higher Chl concentrations [@NASA2014; @NASA_OC4]. This transition from pure band ratios to a band-difference (baseline subtraction) technique was motivated by the need for improved cross-sensor consistency and better performance in oligotrophic waters. Notably, the color‐index method (CI; Hu et al., 2012) was found to be far less sensitive to instrument noise and atmospheric correction errors than the band‐ratio algorithms. As a result, the OCI approach yields smoother, more stable chlorophyll retrievals in low‐Chl oceans, reducing spurious spatial variability and achieving closer agreement between sensors (SeaWiFS, MODIS, VIIRS) [@hu2019improving].

#### Limitations of Existing Approaches
Despite their empirical success, the existing chlorophyll algorithms have fundamental statistical limitations. The regression procedure underpinning OCx algorithms effectively optimizes the relationship in a forward sense – maximizing the likelihood of the observed reflectance given a chlorophyll value $p(Rrs | Chl)$ – rather than solving the inverse problem of interest $p(Chl | Rrs)$. In practice, the coefficients are tuned to reproduce the sample data distribution, which can lead to overfitting the calibration dataset and biased retrievals when applied broadly. Moreover, uncertainty quantification in these algorithms is ad hoc at best; they provide a point estimate of Chl-a with empirically derived error bounds, but no rigorous probabilistic measure of confidence for each prediction. Another drawback is the use of high-order polynomial fits in the OCx formulations that, while flexible, reduce the interpretability of the model’s spectral response. For example, the standard OC4 algorithm expresses log-transformed Chl-a as a fourth-degree polynomial function of a blue/green reflectance ratio [@OReilly2000]. Such complex polynomial relationships make it difficult to discern how any single band influences the chlorophyll estimate, obscuring physical insights. In summary, current empirical algorithms are not formulated for optimal inference and transparency: they are calibrated to mimic data rather than to robustly infer chlorophyll from reflectance, and they lack a formal mechanism to provide per-observation uncertainty or mechanistic interpretation of the spectral inputs.

#### Motivation for Bayesian Tree-Based Models
To address these limitations, we explore Bayesian tree-based modeling – specifically, Bayesian Additive Regression Trees (BART) – as a novel approach for ocean color chlorophyll retrieval. BART offers several advantages for this application:

Direct Bayesian inference: As a Bayesian ensemble method, BART directly targets the posterior distribution $p(Chl | Rrs)$ rather than a purely empirical fit. The model inherently provides a probabilistic mapping from reflectance to chlorophyll. Rather than producing a single best-fit value, BART returns a posterior mean and credible interval for each prediction, naturally quantifying predictive uncertainty.

Uncertainty quantification: The BART framework includes uncertainty in its predictions by design. Every estimated chlorophyll value comes with a confidence range (e.g., 95% credible interval) reflecting model uncertainty and observational noise. This is a key improvement over deterministic algorithms, enabling more reliable use of the data for scientific and operational purposes where understanding uncertainty is crucial e.g. for forecasting trends.

Interpretability: Being a sum-of-trees model, BART can be interrogated to understand the influence of each input variable. Tools such as partial dependence plots allow us to visualize the relationship between each spectral band and the predicted chlorophyll, marginalizing over other bands. In contrast to high-order polynomial coefficients, these tree-based partial dependencies are intuitively interpretable. For instance, one can observe how changes in a particular waveband’s reflectance (holding others average) drive the chlorophyll estimate, revealing any thresholds or non-linear responses.

BART’s modeling approach is non-parametric and highly flexible, which means it can capture complex, non-linear relationships between multi-spectral Rrs inputs and chlorophyll without imposing a rigid functional form. At the same time, its Bayesian regularization – which shrinks the contribution of each individual tree – helps prevent overfitting even with a rich model structure. This balance between flexibility and regularization is well-suited to global ocean color data, where the true Rrs–Chl relationship is known to be non-linear and context-dependent, yet a model must generalize across diverse water types.

The interpretability of the BART model is illustrated by partial dependence analysis on the six input Rrs bands (centered approximately at 412, 443, 490, 510, 555, and 670 nm). The partial dependence plots reveal nuanced spectral–chlorophyll relationships and clear transitions in the dominant wavelength influence across different chlorophyll regimes. For example, the shorter blue wavelengths (412–443 nm) show a strong inverse influence on predicted $log(Chl_a)$ (higher blue reflectance corresponds to lower chlorophyll) that gradually saturates in the most oligotrophic waters, consistent with the leveling off of blue-to-green ratios at very low Chl. The green band (~555 nm) exhibits a more complex, non-monotonic effect: at low reflectance (clear water) its influence on the Chl estimate is minimal, but it becomes increasingly important through intermediate reflectance ranges – reflecting the band-difference signal exploited by the CI algorithm – and then diminishes for very high chlorophyll where green reflectance tends to flatten. In contrast, the red band (~670 nm) has virtually no effect on the model’s predictions until a threshold is reached at elevated chlorophyll concentrations, after which its influence steeply increases. This behavior aligns with optical expectations, since red reflectance is negligible in low-Chl waters but rises sharply once phytoplankton absorption in the blue saturates and biomass is high. These partial dependence results indicate that the BART model automatically learns the piecewise spectral logic that oceanographers often handle via separate algorithms (blue/green ratios for low-to-moderate Chl, red bands for high Chl). The ability to capture such regime-dependent responses in a single unified model, and to visualize them, underscores the interpretability and scientific insight provided by the BART approach.

Statement of Contribution
In this study, we develop and demonstrate a new global chlorophyll retrieval model based on Bayesian Additive Regression Trees implemented in PyMC. We train the BART model on a large, standardized dataset of satellite remote-sensing reflectance (Rrs) spectra matched with in situ chlorophyll measurements, using log₁₀-transformed Chl-a as the response to stabilize variance. The resulting model is applied globally to produce chlorophyll-a estimates from multi-spectral satellite data, with associated uncertainty estimates for each prediction. We show that this Bayesian tree-based model can serve as a general-purpose ocean color algorithm that is sensor-agnostic (provided reflectances are harmonized to common wavebands), interpretable, and uncertainty-aware. Unlike conventional empirical algorithms, the BART approach allows users to examine the inferred Rrs–Chl relationships and trust the model’s performance across regimes, while also quantifying confidence in each retrieval. This work thus contributes a novel methodological advance to satellite ocean color science: a unified chlorophyll retrieval model that marries the strengths of empirical algorithms (global applicability and simplicity) with the benefits of modern Bayesian machine learning (flexibility, interpretability, and rigorous uncertainty quantification). Our introduction of BART for global chlorophyll prediction opens the door for more robust monitoring of ocean biogeochemistry and improved integration of ocean color data into scientific and management applications.

### 2025-04-08 manuscript/index.qmd
improved introduction

--- 2025-04-04 manuscript/paper.qmd
@@
 ---
-title: "Introduction for a BART-Based Model to Predict Chlorophyll-a from Satellite Reflectance"
-author: Erdem Karaköylü
-date: "2025-04-03"
-format:
-  pdf:
-    documentclass: agujournal2019
-    template: templates/agujournaltemplate.tex
-    keep-tex: true
+title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
+author:  
+  - name: Erdem M. Karaköylü
+    orcid: 0000-0002-6156-1720
+    corresponding: true
+    email: erdemk@protonmail.com
+    roles:
+      - Investigation
+      - Software
+      - Visualization
+    affiliations:
+      - Consultant
+  - name: Susanne E. Craig
+    orcid: 0000-0002-8963-0951
+    corresponding: true
+    email: susanne.e.craig@nasa.gov
+    roles:
+      - Investigation
+      - Project administration
+    affiliations:
+      - NASA
+abstract: Placeholder
+date: last-modified
 bibliography: references.bib
@@
 #### Historical Context of Chlorophyll Algorithms
-Satellite ocean color missions have long relied on empirical spectral algorithms to estimate near-surface chlorophyll-a (Chl-a) concentrations from water‐leaving radiance.
-NASA’s standard approach since the SeaWiFS era has been the family of “OCx” band‐ratio algorithms, which use blue-to-green reflectance ratios fit to in situ Chl data via polynomial regression [@OReilly2000].
-These O’Reilly et al.
-formulations (e.g., OC4 for SeaWiFS, OC3 for MODIS) became the default global chlorophyll algorithms for multiple sensors due to their relative simplicity and consistency in Case-1 waters.
-In 2014, NASA introduced the Ocean Color Index (OCI) algorithm, a hybrid empirical model that blends a three‐band reflectance difference method for low‐Chl regimes with the traditional OCx ratio approach at higher Chl concentrations [@NASA2014; @NASA_OC4].
-This transition from pure band ratios to a band-difference (baseline subtraction) technique was motivated by the need for improved cross-sensor consistency and better performance in oligotrophic waters.
-Notably, the color‐index method (CI; Hu et al., 2012) was found to be far less sensitive to instrument noise and atmospheric correction errors than the band‐ratio algorithms.
-As a result, the OCI approach yields smoother, more stable chlorophyll retrievals in low‐Chl oceans, reducing spurious spatial variability and achieving closer agreement between sensors (SeaWiFS, MODIS, VIIRS) [@hu2019improving].
+Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a—a key indicator of phytoplankton biomass and ocean productivity.
+Early empirical algorithms, notably developed by O’Reilly et al.
+[@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models.
+These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ chlorophyll‑a, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial.
+These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters.
+However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.
+
+Subsequent refinements were introduced to address these deficiencies.
+For example, Hu et al.
+[@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving].
+The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations.
+O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, improving the dynamic range at low chlorophyll concentrations.
+In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor.
+With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.
 
 #### Limitations of Existing Approaches
-Despite their empirical success, the existing chlorophyll algorithms have fundamental statistical limitations.
-The regression procedure underpinning OCx algorithms effectively optimizes the relationship in a forward sense – maximizing the likelihood of the observed reflectance given a chlorophyll value $p(Rrs | Chl)$ – rather than solving the inverse problem of interest $p(Chl | Rrs)$.
-In practice, the coefficients are tuned to reproduce the sample data distribution, which can lead to overfitting the calibration dataset and biased retrievals when applied broadly.
-Moreover, uncertainty quantification in these algorithms is ad hoc at best; they provide a point estimate of Chl-a with empirically derived error bounds, but no rigorous probabilistic measure of confidence for each prediction.
-Another drawback is the use of high-order polynomial fits in the OCx formulations that, while flexible, reduce the interpretability of the model’s spectral response.
-For example, the standard OC4 algorithm expresses log-transformed Chl-a as a fourth-degree polynomial function of a blue/green reflectance ratio [@OReilly2000].
-Such complex polynomial relationships make it difficult to discern how any single band influences the chlorophyll estimate, obscuring physical insights.
-In summary, current empirical algorithms are not formulated for optimal inference and transparency: they are calibrated to mimic data rather than to robustly infer chlorophyll from reflectance, and they lack a formal mechanism to provide per-observation uncertainty or mechanistic interpretation of the spectral inputs.
+Unfortunately, the development process for traditional ocean color algorithms is founded on a statistical fallacy.
+Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena.
+These samples' probabilities are  considered the observed frequencies that capture the the true - albeit noised up and potentially biased - behavior of the system.
+Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $$D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $$M$ is true.
+The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$.
+Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model.
+In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek.
+The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
+
+Bayesianism, Frequentism's archrival, provides the framoework to establish the plausibility of $M$ using observations through the reinterpretation of how probability should be construed.
+As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a statement of our knowledge rather than as an inherent property of the data.
+the Bayesian perspective treats the data, once observed, as fixed and the model parameters as latent random variables the uncertainty of which  can—and should—be explicitly modeled via prior distributions.
+Prior distributions are probabilistic constructs that help workers encode prior knowledge.
+Priors can represents knowledge elicited from subject matter experts, assumptions held by the modeler or a complete absence thereof.
+
+This conceptual framework dissolves the false dichotomy between empirical and semi‑analytical approaches.
+Both are, in essence, different hypotheses or approximations about how ocean color data are generated; they merely provide alternative expressions of the likelihood
+$p(Chl | Rrs)$ to be updated in light of new data.
+By adopting a Bayesian paradigm, our approach integrates prior knowledge with observed data to yield a full posterior predictive distribution—offering not only improved predictive performance but also principled uncertainty quantification.
+In doing so, it circumvents the need for the proliferation of narrowly tuned, sensor-specific variants (65 versions for 25 sensors, as documented by O’Reilly and Werdell [@OREILLY2019]) that are symptomatic of ad hoc calibration processes.
+Detailed methodological comparisons and additional model refinements are provided in the Supplementary Material.
 

### 2025-04-08 manuscript/manuscript.qmd
upated manuscript

--- 2025-04-08 manuscript/index.qmd
@@
 title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
+format:
+    agu-pdf:
+        keep-tex: true
+    agu-html: default
 author:  
@@
     orcid: 0000-0002-8963-0951
-    corresponding: true
+    corresponding: false
     email: susanne.e.craig@nasa.gov
@@
 abstract: Placeholder
+plain-language-summary: Placeholder
+keywords: []
+key-points:
+  - Key Points convey the main points and conclusions of the article. 
+  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
+  - Key Points are not included in the word count.
+bibliography: references.bib
+citation:
+  container-title: Geophysical Research Letters
+keep-tex: true
 date: last-modified
-bibliography: references.bib
 ---
@@
 #### Limitations of Existing Approaches
-Unfortunately, the development process for traditional ocean color algorithms is founded on a statistical fallacy.
-Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena.
-These samples' probabilities are  considered the observed frequencies that capture the the true - albeit noised up and potentially biased - behavior of the system.
-Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $$D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $$M$ is true.
+Regrettably, the development process for traditional ocean color algorithms is founded on a statistical erroneous argument.
+The fallacy is the error of conflating empirical frequency — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials,Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena.
+These samples' probabilities are considered the observed frequencies that capture the true - albeit noised up and potentially biased - behavior of the system.
+Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $M$ is true.
 The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$.
@@
 The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
+
+Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error.
+The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities.
+In other words, it is the error of conflating the empirical frequency—derived from a limited, noisy set of observations—with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling.
+The fallout has and continues to be extremely detrimental.
+Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.
+
+In the context of statistical modelling, where $M$ is a model that represents a hypothesized relationship between independent and dependent variables, data $D$ is collected to evaluate how well this relationship holds out.
+A skilled enough model $M$ is retained and used to make predictions; it is put into production.
+However practical problems arise because under the requentist framework, model evaluation centers on the likelihood $p(D∣M)$, the probability of observing the data given the model is true.
+With enough data the frequentist assumption is that this likelihood can be directly flipped to give the probability of the model being true, $p(M∣D)$.
+The leap of faith is completed by imposing unverified assumptions to allow setting model parameters by maximizing $p(D∣M)$.
+By this logical doing so provides the best model possible.
+The historical dimensions of how this snake oil was progressively brewed are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
 
@@
 Our introduction of BART for global chlorophyll prediction opens the door for more robust monitoring of ocean biogeochemistry and improved integration of ocean color data into scientific and management applications.
+
+## Acknowledgments
+
+## Open research
+
+## References {.unnumbered}
+
+:::{#refs}
+
+:::

### 2025-04-08 manuscript/manuscript.tex
upated manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation},
  pdfauthor={Erdem M. Karaköylü; Susanne E. Craig},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Geophysical Research Letters}

\draftfalse

\begin{document}
\title{Shifting paradigms in Ocean Color: Bayesian Inference for
Uncertainty-Aware Chlorophyll Estimation}

\authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
\affiliation{1}{Consultant, }\affiliation{2}{NASA, }
\correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}

\begin{abstract}
Placeholder
\end{abstract}

\section*{Plain Language Summary}
Placeholder

\subsection{Introduction}\label{introduction}

\subsubsection{Historical Context of Chlorophyll
Algorithms}\label{historical-context-of-chlorophyll-algorithms}

Satellite ocean color observations have long been fundamental for
monitoring marine ecosystems, as they enable the global estimation of
chlorophyll‑a---a key indicator of phytoplankton biomass and ocean
productivity. Early empirical algorithms, notably developed by O'Reilly
et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al., 2000),
established the \(OCx\) family (where \(x\) denotes the number of bands
used) of polynomial regression models. These models relate blue-to-green
reflectance ratios (after log‑transformation) to in situ chlorophyll‑a,
employing either straight band ratios (BR) or maximum band ratios
(MBR)---the latter selecting the highest available blue-to-green ratio
for any given observation as input to a high‑order polynomial. These
formulations have served as the operational foundation for chlorophyll‑a
products across a broad range of satellite ocean color sensors---from
the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS,
and MERIS to more recent missions---offering a straightforward and
robust approach for Case‑1 waters. However, their performance is more
limited in optically complex Case‑2 waters and remains sensitive to
atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies.
For example, Hu et al. (Hu et al., 2012) proposed a Color Index (CI)
formulation that employs a band‑difference approach to reduce
sensitivity to residual atmospheric errors and instrument noise, with
further improvements enhancing inter‑sensor agreement (Hu et al., 2019).
The increasing availability of calibration data (e.g.,
(\textbf{Valente2015?})) and ongoing algorithmic improvements have led
to the development of additional variants of the \(OCx\)
algorithms---specifically, the OC5 and OC6 formulations. O'Reilly and
Werdell (John E. O'Reilly \& Werdell, 2019) maintain that OC5 extends
the spectral basis by incorporating the 412\,nm band, thereby exploiting
its strong signal in clear, oligotrophic waters, while OC6 replaces the
traditional denominator with the mean of the 555 and 670\,nm
reflectances, improving the dynamic range at low chlorophyll
concentrations. In total, (John E. O'Reilly \& Werdell, 2019) propose 65
versions of BR/MBR \(OCx\) type algorithms for 25 sensors---on average,
two or more variants per sensor. With this arsenal, it is hoped,
researchers are better equipped to address the wide array of bio‑optical
environments encountered in global ocean color applications.

\subsubsection{Limitations of Existing
Approaches}\label{limitations-of-existing-approaches}

Unfortunately, the development process for traditional ocean color
algorithms is founded on a statistical fallacy. Frequentism---the
classical approach---assumes that observations are random samples drawn
from the studied phenomena. These samples' probabilities are considered
the observed frequencies that capture the the true - albeit noised up
and potentially biased - behavior of the system. Under this framework,
if a phenomenon is hypothesized to be represented by a model \(M\) and
data \(D\) is collected, then the model's representativeness is
evaluated solely through the likelihood \(p(D∣M)\), which measures the
probability of observing the data given that \(M\) is true. The fallacy
arises from the mistaken belief that, with sufficiently large amounts of
data, one can simply ``flip'' the likelihood to obtain \(p(M∣D)\),
thereby yielding a probabilistic statement about \(M\). Compounding this
error is the imposition of numerous unverified assumptions---disguised
as objectivity---that lead frequentists to set the parameters of \(M\)
by maximizing \(p(D∣M)\), under the false premise that this provides the
best possible representation of the model. In essence, this fallacy
leads to a confusion between the sampling probability \(p(D∣M)\) and the
inferential probability \(p(M∣D)\); the latter when correctly obtained
quantifies the plausibility of the hypothesis represented by \(M\) -
precisely what we seek. The historical dimensions of this frequentist
illusion are fascinating in their own right, and we invite interested
readers to consult works such as McGrayne (\textbf{mcgrayne?}) and
Clayton (\textbf{clayton?}).

Bayesianism, Frequentism's archrival, provides the framoework to
establish the plausibility of \(M\) using observations through the
reinterpretation of how probability should be construed. As elaborated
extensively by Jaynes (\textbf{jaynes2003probability?}), and more
recently by practitioners like Gelman (\textbf{gelman2013bayesian?}),
and McElreath (\textbf{mcelreath2015statistical?}), probability is best
understood as a statement of our knowledge rather than as an inherent
property of the data. the Bayesian perspective treats the data, once
observed, as fixed and the model parameters as latent random variables
the uncertainty of which can---and should---be explicitly modeled via
prior distributions. Prior distributions are probabilistic constructs
that help workers encode prior knowledge. Priors can represents
knowledge elicited from subject matter experts, assumptions held by the
modeler or a complete absence thereof.

This conceptual framework dissolves the false dichotomy between
empirical and semi‑analytical approaches. Both are, in essence,
different hypotheses or approximations about how ocean color data are
generated; they merely provide alternative expressions of the likelihood
\(p(Chl | Rrs)\) to be updated in light of new data. By adopting a
Bayesian paradigm, our approach integrates prior knowledge with observed
data to yield a full posterior predictive distribution---offering not
only improved predictive performance but also principled uncertainty
quantification. In doing so, it circumvents the need for the
proliferation of narrowly tuned, sensor-specific variants (65 versions
for 25 sensors, as documented by O'Reilly and Werdell
(\textbf{OREILLY2019?})) that are symptomatic of ad hoc calibration
processes. Detailed methodological comparisons and additional model
refinements are provided in the Supplementary Material.

\subsubsection{Motivation for Bayesian Tree-Based
Models}\label{motivation-for-bayesian-tree-based-models}

To address these limitations, we explore Bayesian tree-based modeling --
specifically, Bayesian Additive Regression Trees (BART) -- as a novel
approach for ocean color chlorophyll retrieval. BART offers several
advantages for this application:

Direct Bayesian inference: As a Bayesian ensemble method, BART directly
targets the posterior distribution \(p(Chl | Rrs)\) rather than a purely
empirical fit. The model inherently provides a probabilistic mapping
from reflectance to chlorophyll. Rather than producing a single best-fit
value, BART returns a posterior mean and credible interval for each
prediction, naturally quantifying predictive uncertainty.

Uncertainty quantification: The BART framework includes uncertainty in
its predictions by design. Every estimated chlorophyll value comes with
a confidence range (e.g., 95\% credible interval) reflecting model
uncertainty and observational noise. This is a key improvement over
deterministic algorithms, enabling more reliable use of the data for
scientific and operational purposes where understanding uncertainty is
crucial e.g.~for forecasting trends.

Interpretability: Being a sum-of-trees model, BART can be interrogated
to understand the influence of each input variable. Tools such as
partial dependence plots allow us to visualize the relationship between
each spectral band and the predicted chlorophyll, marginalizing over
other bands. In contrast to high-order polynomial coefficients, these
tree-based partial dependencies are intuitively interpretable. For
instance, one can observe how changes in a particular waveband's
reflectance (holding others average) drive the chlorophyll estimate,
revealing any thresholds or non-linear responses.

BART's modeling approach is non-parametric and highly flexible, which
means it can capture complex, non-linear relationships between
multi-spectral Rrs inputs and chlorophyll without imposing a rigid
functional form. At the same time, its Bayesian regularization -- which
shrinks the contribution of each individual tree -- helps prevent
overfitting even with a rich model structure. This balance between
flexibility and regularization is well-suited to global ocean color
data, where the true Rrs--Chl relationship is known to be non-linear and
context-dependent, yet a model must generalize across diverse water
types.

The interpretability of the BART model is illustrated by partial
dependence analysis on the six input Rrs bands (centered approximately
at 412, 443, 490, 510, 555, and 670 nm). The partial dependence plots
reveal nuanced spectral--chlorophyll relationships and clear transitions
in the dominant wavelength influence across different chlorophyll
regimes. For example, the shorter blue wavelengths (412--443 nm) show a
strong inverse influence on predicted \(log(Chl_a)\) (higher blue
reflectance corresponds to lower chlorophyll) that gradually saturates
in the most oligotrophic waters, consistent with the leveling off of
blue-to-green ratios at very low Chl. The green band (\textasciitilde555
nm) exhibits a more complex, non-monotonic effect: at low reflectance
(clear water) its influence on the Chl estimate is minimal, but it
becomes increasingly important through intermediate reflectance ranges
-- reflecting the band-difference signal exploited by the CI algorithm
-- and then diminishes for very high chlorophyll where green reflectance
tends to flatten. In contrast, the red band (\textasciitilde670 nm) has
virtually no effect on the model's predictions until a threshold is
reached at elevated chlorophyll concentrations, after which its
influence steeply increases. This behavior aligns with optical
expectations, since red reflectance is negligible in low-Chl waters but
rises sharply once phytoplankton absorption in the blue saturates and
biomass is high. These partial dependence results indicate that the BART
model automatically learns the piecewise spectral logic that
oceanographers often handle via separate algorithms (blue/green ratios
for low-to-moderate Chl, red bands for high Chl). The ability to capture
such regime-dependent responses in a single unified model, and to
visualize them, underscores the interpretability and scientific insight
provided by the BART approach.

Statement of Contribution In this study, we develop and demonstrate a
new global chlorophyll retrieval model based on Bayesian Additive
Regression Trees implemented in PyMC. We train the BART model on a
large, standardized dataset of satellite remote-sensing reflectance
(Rrs) spectra matched with in situ chlorophyll measurements, using
log₁₀-transformed Chl-a as the response to stabilize variance. The
resulting model is applied globally to produce chlorophyll-a estimates
from multi-spectral satellite data, with associated uncertainty
estimates for each prediction. We show that this Bayesian tree-based
model can serve as a general-purpose ocean color algorithm that is
sensor-agnostic (provided reflectances are harmonized to common
wavebands), interpretable, and uncertainty-aware. Unlike conventional
empirical algorithms, the BART approach allows users to examine the
inferred Rrs--Chl relationships and trust the model's performance across
regimes, while also quantifying confidence in each retrieval. This work
thus contributes a novel methodological advance to satellite ocean color
science: a unified chlorophyll retrieval model that marries the
strengths of empirical algorithms (global applicability and simplicity)
with the benefits of modern Bayesian machine learning (flexibility,
interpretability, and rigorous uncertainty quantification). Our
introduction of BART for global chlorophyll prediction opens the door
for more robust monitoring of ocean biogeochemistry and improved
integration of ocean color data into scientific and management
applications.

\section{Acknowledgments}\label{acknowledgments}

\section{Open research}\label{open-research}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-hu2012novel}
Hu, C., Lee, Z., \& Franz, B. (2012). Chlorophyll aalgorithms for
oligotrophic oceans: A novel approach based on three-band reflectance
difference. \emph{Journal of Geophysical Research: Oceans},
\emph{117}(C1).
https://doi.org/\url{https://doi.org/10.1029/2011JC007395}

\bibitem[\citeproctext]{ref-hu2019improving}
Hu, C., Feng, L., Lee, Z., Franz, B. A., Bailey, S. W., Werdell, P. J.,
\& Proctor, C. W. (2019). Improving satellite global chlorophyll a data
products through algorithm refinement and data recovery. \emph{Journal
of Geophysical Research: Oceans}, \emph{124}(3), 1524--1543.
https://doi.org/\url{https://doi.org/10.1029/2019JC014941}

\bibitem[\citeproctext]{ref-oreilly2019}
O'Reilly, John E., \& Werdell, P. J. (2019). Chlorophyll algorithms for
ocean color sensors - OC4, OC5 \& OC6. \emph{Remote Sensing of
Environment}, \emph{229}, 32--47.
https://doi.org/\url{https://doi.org/10.1016/j.rse.2019.04.021}

\bibitem[\citeproctext]{ref-oreilly1998}
O'Reilly, John E., Maritorena, S., Mitchell, B. G., Siegel, D. A.,
Carder, K. L., Garver, S. A., et al. (1998). Ocean color chlorophyll
algorithms for SeaWiFS. \emph{Journal of Geophysical Research: Oceans},
\emph{103}(C11), 24937--24953.
https://doi.org/\url{https://doi.org/10.1029/98JC02160}

\bibitem[\citeproctext]{ref-oreilly2000}
O'Reilly, John E., Maritorena, S., Siegel, D. A., O'Brien, M. C., Toole,
D., Mitchell, B. G., et al. (2000). Ocean color chlorophyll a algorithms
for SeaWiFS, OC2, and OC4: Version 4. \emph{SeaWiFS Postlaunch
Calibration and Validation Analyses, Part}, \emph{3}, 9--23.

\end{CSLReferences}

\end{document}

### 2025-04-08 manuscript/manuscript.qmd
updated introduction

@@
 #### Limitations of Existing Approaches
-Regrettably, the development process for traditional ocean color algorithms is founded on a statistical erroneous argument.
-The fallacy is the error of conflating empirical frequency — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials,Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena.
-These samples' probabilities are considered the observed frequencies that capture the true - albeit noised up and potentially biased - behavior of the system.
-Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $M$ is true.
-The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$.
-Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model.
-In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek.
-The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
-
 Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error.
 The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities.
-In other words, it is the error of conflating the empirical frequency—derived from a limited, noisy set of observations—with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling.
+In other words, it is the error of conflating  empirical frequencies — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling.
 The fallout has and continues to be extremely detrimental.
@@
 
-In the context of statistical modelling, where $M$ is a model that represents a hypothesized relationship between independent and dependent variables, data $D$ is collected to evaluate how well this relationship holds out.
-A skilled enough model $M$ is retained and used to make predictions; it is put into production.
-However practical problems arise because under the requentist framework, model evaluation centers on the likelihood $p(D∣M)$, the probability of observing the data given the model is true.
-With enough data the frequentist assumption is that this likelihood can be directly flipped to give the probability of the model being true, $p(M∣D)$.
-The leap of faith is completed by imposing unverified assumptions to allow setting model parameters by maximizing $p(D∣M)$.
-By this logical doing so provides the best model possible.
-The historical dimensions of how this snake oil was progressively brewed are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
+In statistical modeling, the goal is to assess how well a hypothesized model $M$ represents the relationship between independent and dependent variables based on collected data $D$.
+Frequentist methods focus exclusively on the likelihood $p(D|M)$, estimating model parameters by maximizing this probability.
+This approach assumes that, with a sufficiently large dataset, the chosen parameters reliably reflect the true system.
+However, in practice, any finite dataset carries inherent uncertainty and noise.
+As a consequence, models developed under this framework tend to be overconfident---yielding point estimates that do not adequately capture the variability in in situ observations---and may not generalize well to new data.
+Ad hoc attempts to generate uncertainty, such as bootstrapping, further compound the problem by resampling the flawed, finite observations and thus amplifying any existing biases.
+This overconfidence, coupled with unverified assumptions presented as objectivity, can lead to misleading inferences and contribute to replication challenges.
+The historical process by which these assumptions have been enshrined is fascinating in its own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
 
-Bayesianism, Frequentism's archrival, provides the framoework to establish the plausibility of $M$ using observations through the reinterpretation of how probability should be construed.
-As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a statement of our knowledge rather than as an inherent property of the data.
-the Bayesian perspective treats the data, once observed, as fixed and the model parameters as latent random variables the uncertainty of which  can—and should—be explicitly modeled via prior distributions.
-Prior distributions are probabilistic constructs that help workers encode prior knowledge.
-Priors can represents knowledge elicited from subject matter experts, assumptions held by the modeler or a complete absence thereof.
+As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a degree of knowledge rather than as a long-run frequency.
+Accordingly in this Bayesian framework, the prior $p(M)$ encodes our current state of knowledge about the model's parameters before $D$ is observed.
+This knowledge can be elicited from experts, or if none exixts weakly informative priors can be used.
+Note that whether $M$ is based on empirically observed relationships, analytical understanding, or something in between is irrelevant to its integration into this framework.
+Once the data $D$ are collected, Bayes' theorem combines the likelihood $p(D|M)$ with the prior $p(M)$ to yield  the posterior distribution $p(M|D)$, which represents an updated level of knowledge about the model's parameters.
+Crucially this posterior can then serve as the new prior when new data become available, allowing for sequential integration of new information.
+The by-product of this approach is uncertainty estimates that more accurately reflect the complexities of the underlying system as currently understood.
+In effect, rather than relying on single point estimates based on $p(D|M)$, the posterior $p(M|D)$ can be mined to retrieve a range of plausible parameter values through credible intervals offering a more transparent and robust basis for inference.
+This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.
 
-This conceptual framework dissolves the false dichotomy between empirical and semi‑analytical approaches.
-Both are, in essence, different hypotheses or approximations about how ocean color data are generated; they merely provide alternative expressions of the likelihood
-$p(Chl | Rrs)$ to be updated in light of new data.
-By adopting a Bayesian paradigm, our approach integrates prior knowledge with observed data to yield a full posterior predictive distribution—offering not only improved predictive performance but also principled uncertainty quantification.
-In doing so, it circumvents the need for the proliferation of narrowly tuned, sensor-specific variants (65 versions for 25 sensors, as documented by O’Reilly and Werdell [@OREILLY2019]) that are symptomatic of ad hoc calibration processes.
-Detailed methodological comparisons and additional model refinements are provided in the Supplementary Material.
+In the Bayesian framework, the prior 
+p(M) encodes our current state of knowledge about the model's parameters before any data are observed.
+This knowledge can be elicited from experts or determined experimentally when expert opinion is unavailable.
+Once data
+D are collected, Bayes’ theorem combines the likelihood 
+p(D∣M) with the prior to yield the posterior distribution 
+p(M∣D), which represents our updated belief about the parameters.
+Crucially, this posterior can then serve as the new prior when new data become available, allowing for seamless sequential updating.
+In this way, whether the model is derived empirically, analytically, or somewhere in between, it is treated as a hypothesis whose parameters are updated in light of new data.
+This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.
 

### 2025-04-10 manuscript/manuscript.qmd
finished introduction draft

@@
 #### Historical Context of Chlorophyll Algorithms
-Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a—a key indicator of phytoplankton biomass and ocean productivity.
+Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a ($Chl_a$) —a key indicator of phytoplankton biomass and ocean productivity.
 Early empirical algorithms, notably developed by O’Reilly et al.
 [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models.
-These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ chlorophyll‑a, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial.
+These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial.
 These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters.
@@
 The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations.
-O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, improving the dynamic range at low chlorophyll concentrations.
+O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, with the aim of improving the dynamic range at low chlorophyll concentrations.
 In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor.
@@
 #### Limitations of Existing Approaches
-Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error.
-The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities.
-In other words, it is the error of conflating  empirical frequencies — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling.
-The fallout has and continues to be extremely detrimental.
-Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.
+Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability.
+Consider data $D$ that include predictive features (e.g.
+Remote sensing reflectance - Rrs) and prediction targets (e.g.
+$Chl_a$ or phytoplankton absorption), and a model $M$ (e.g.
+OCx) hypothesized to adequately represent the statistical association between the two.
+The sampling probability $p(D|M)$ also referred to as the likelihood, is the probability of $D$ conditioned on $M$ being "true".
+The common model fitting practice is to maximize $p(D|M)$ - likelihood maximization - achieved by tuning $M$'s parameters.
+The unspoken assumption is that making $M$ fit $D$ as well as possible yields the best version of $M$ given $D$ i.e.
+maximizing $p(M|D)$, the inferential probability.
+This sneaky swapping from $p(D|M)$ to $p(M|D)$ is a violation of the rules of conditional probability, Bayes' theorem in particular.
+Jaynes[@jaynes2003] and later Clayton through his highly readable account [@clayton] are some of the more notable whistleblowers.
+Occasionally, in simple  or well-behaved problems, with a large amount of data and no prior information available the maxima of both $p(M|D)$ and $p(D|M)$ can indeed converge on the same answer.
+Yet there is no guarantee predictability for this outcome however, and the fallout of this abuse is non-negligible, most noticeably in terms of models that don't generalize well, ad-hoc uncertainty estimation attempts, and more broadly a replication crisis.
 
-In statistical modeling, the goal is to assess how well a hypothesized model $M$ represents the relationship between independent and dependent variables based on collected data $D$.
-Frequentist methods focus exclusively on the likelihood $p(D|M)$, estimating model parameters by maximizing this probability.
-This approach assumes that, with a sufficiently large dataset, the chosen parameters reliably reflect the true system.
-However, in practice, any finite dataset carries inherent uncertainty and noise.
-As a consequence, models developed under this framework tend to be overconfident---yielding point estimates that do not adequately capture the variability in in situ observations---and may not generalize well to new data.
-Ad hoc attempts to generate uncertainty, such as bootstrapping, further compound the problem by resampling the flawed, finite observations and thus amplifying any existing biases.
-This overconfidence, coupled with unverified assumptions presented as objectivity, can lead to misleading inferences and contribute to replication challenges.
-The historical process by which these assumptions have been enshrined is fascinating in its own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].
+Some researchers have attempted to deal with these issues.
+[@seegers2018] e.g.
+have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
+Others have tried to go a step futher and switch to the Bayesian paradigm.
+E.g.
+[ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction.
+[@shi2015] proposed a probabilistic method to merge data from different sensors.
+[@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
+[@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty.
+[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model.
+[@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
+Yet most of these approaches retain variable levels of frequentism by applying only part of what is commonly referred to as the Bayesian workflow[@betancourt2019].
 
-As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a degree of knowledge rather than as a long-run frequency.
-Accordingly in this Bayesian framework, the prior $p(M)$ encodes our current state of knowledge about the model's parameters before $D$ is observed.
-This knowledge can be elicited from experts, or if none exixts weakly informative priors can be used.
-Note that whether $M$ is based on empirically observed relationships, analytical understanding, or something in between is irrelevant to its integration into this framework.
-Once the data $D$ are collected, Bayes' theorem combines the likelihood $p(D|M)$ with the prior $p(M)$ to yield  the posterior distribution $p(M|D)$, which represents an updated level of knowledge about the model's parameters.
-Crucially this posterior can then serve as the new prior when new data become available, allowing for sequential integration of new information.
-The by-product of this approach is uncertainty estimates that more accurately reflect the complexities of the underlying system as currently understood.
-In effect, rather than relying on single point estimates based on $p(D|M)$, the posterior $p(M|D)$ can be mined to retrieve a range of plausible parameter values through credible intervals offering a more transparent and robust basis for inference.
-This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.
+In this paper we propose a modified workflow to Bayesian modeling for Ocean Color emote sensing.
+We propose that this workflow include; (1) one or more model building, (2) prior formulation via expert knowledge elicitation and/or data-free simulation, (3) fitting model to data, (4) model validation, comparison and selection, (5) interpretation of results via posterior analysis, (6) prediction on new observation, (7) update the posterior when new labeled data becomes available.
+As illustration we recast some of the pre-existing $OCx$ models into their probabilistic version.
+To illustrate model comparison and selection we also build and evaluate Bayesian Additive Tree Regregssion  (BART; [@chipman2010bart]) type model.
 
-In the Bayesian framework, the prior 
-p(M) encodes our current state of knowledge about the model's parameters before any data are observed.
-This knowledge can be elicited from experts or determined experimentally when expert opinion is unavailable.
-Once data
-D are collected, Bayes’ theorem combines the likelihood 
-p(D∣M) with the prior to yield the posterior distribution 
-p(M∣D), which represents our updated belief about the parameters.
-Crucially, this posterior can then serve as the new prior when new data become available, allowing for seamless sequential updating.
-In this way, whether the model is derived empirically, analytically, or somewhere in between, it is treated as a hypothesis whose parameters are updated in light of new data.
-This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.
+### Methods
 
-#### Motivation for Bayesian Tree-Based Models
-To address these limitations, we explore Bayesian tree-based modeling – specifically, Bayesian Additive Regression Trees (BART) – as a novel approach for ocean color chlorophyll retrieval.
-BART offers several advantages for this application:
+#### Prior Elicitation and Model Formulation
 
-Direct Bayesian inference: As a Bayesian ensemble method, BART directly targets the posterior distribution $p(Chl | Rrs)$ rather than a purely empirical fit.
-The model inherently provides a probabilistic mapping from reflectance to chlorophyll.
-Rather than producing a single best-fit value, BART returns a posterior mean and credible interval for each prediction, naturally quantifying predictive uncertainty.
+#### Data Preprocessing
 
-Uncertainty quantification: The BART framework includes uncertainty in its predictions by design.
-Every estimated chlorophyll value comes with a confidence range (e.g., 95% credible interval) reflecting model uncertainty and observational noise.
-This is a key improvement over deterministic algorithms, enabling more reliable use of the data for scientific and operational purposes where understanding uncertainty is crucial e.g.
-for forecasting trends.
+####
 
-Interpretability: Being a sum-of-trees model, BART can be interrogated to understand the influence of each input variable.
-Tools such as partial dependence plots allow us to visualize the relationship between each spectral band and the predicted chlorophyll, marginalizing over other bands.
-In contrast to high-order polynomial coefficients, these tree-based partial dependencies are intuitively interpretable.
-For instance, one can observe how changes in a particular waveband’s reflectance (holding others average) drive the chlorophyll estimate, revealing any thresholds or non-linear responses.
+\section{Data Preprocessing}
 
-BART’s modeling approach is non-parametric and highly flexible, which means it can capture complex, non-linear relationships between multi-spectral Rrs inputs and chlorophyll without imposing a rigid functional form.
-At the same time, its Bayesian regularization – which shrinks the contribution of each individual tree – helps prevent overfitting even with a rich model structure.
-This balance between flexibility and regularization is well-suited to global ocean color data, where the true Rrs–Chl relationship is known to be non-linear and context-dependent, yet a model must generalize across diverse water types.
+Data for this study were acquired from multiple satellite ocean color sensors and corresponding in situ chlorophyll-\(a\) measurements obtained from sources such as the NASA Bio-Optical Marine Algorithm Data set (NOMAD) and the compilation by Valente et al.
+(2015).
+To ensure consistency across sensors, the spectral reflectance data (\(R_{rs}\)) were interpolated as needed to common wavelength centers.
 
-The interpretability of the BART model is illustrated by partial dependence analysis on the six input Rrs bands (centered approximately at 412, 443, 490, 510, 555, and 670 nm).
-The partial dependence plots reveal nuanced spectral–chlorophyll relationships and clear transitions in the dominant wavelength influence across different chlorophyll regimes.
-For example, the shorter blue wavelengths (412–443 nm) show a strong inverse influence on predicted $log(Chl_a)$ (higher blue reflectance corresponds to lower chlorophyll) that gradually saturates in the most oligotrophic waters, consistent with the leveling off of blue-to-green ratios at very low Chl.
-The green band (~555 nm) exhibits a more complex, non-monotonic effect: at low reflectance (clear water) its influence on the Chl estimate is minimal, but it becomes increasingly important through intermediate reflectance ranges – reflecting the band-difference signal exploited by the CI algorithm – and then diminishes for very high chlorophyll where green reflectance tends to flatten.
-In contrast, the red band (~670 nm) has virtually no effect on the model’s predictions until a threshold is reached at elevated chlorophyll concentrations, after which its influence steeply increases.
-This behavior aligns with optical expectations, since red reflectance is negligible in low-Chl waters but rises sharply once phytoplankton absorption in the blue saturates and biomass is high.
-These partial dependence results indicate that the BART model automatically learns the piecewise spectral logic that oceanographers often handle via separate algorithms (blue/green ratios for low-to-moderate Chl, red bands for high Chl).
-The ability to capture such regime-dependent responses in a single unified model, and to visualize them, underscores the interpretability and scientific insight provided by the BART approach.
+For the empirical \(OCx\) formulation, blue-to-green band ratios were computed for each observation.
+In particular, the maximum band ratio (MBR) was determined by taking the highest value among the available blue-band ratios (e.g., \(Rrs(443)/Rrs(555)\), \(Rrs(490)/Rrs(555)\), and \(Rrs(510)/Rrs(555)\)).
+This maximum value was then log-transformed:
+$$
+\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
+$$
 
-Statement of Contribution
+For the Color Index (CI) formulation of Hu et al. (2012), the CI was calculated as:
+$$
+\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
+$$
+and the corresponding in situ chlorophyll-\(a\) concentrations were log-transformed:
+$$
+\log \text{Chl} = \log_{10}(\text{Chl}).
+$$
+
+These transformations standardize the data to a common scale, ensuring that variability is appropriately captured for subsequent regression and uncertainty quantification.
+Detailed descriptions of the interpolation methods and quality control procedures are provided in the Supplementary Material.
+
+## Statement of Contribution
 In this study, we develop and demonstrate a new global chlorophyll retrieval model based on Bayesian Additive Regression Trees implemented in PyMC.

### 2025-04-10 manuscript/manuscript.tex
finished introduction draft

@@
 
-Unfortunately, the development process for traditional ocean color
-algorithms is founded on a statistical fallacy. Frequentism---the
-classical approach---assumes that observations are random samples drawn
-from the studied phenomena. These samples' probabilities are considered
-the observed frequencies that capture the the true - albeit noised up
-and potentially biased - behavior of the system. Under this framework,
-if a phenomenon is hypothesized to be represented by a model \(M\) and
-data \(D\) is collected, then the model's representativeness is
-evaluated solely through the likelihood \(p(D∣M)\), which measures the
-probability of observing the data given that \(M\) is true. The fallacy
-arises from the mistaken belief that, with sufficiently large amounts of
-data, one can simply ``flip'' the likelihood to obtain \(p(M∣D)\),
-thereby yielding a probabilistic statement about \(M\). Compounding this
-error is the imposition of numerous unverified assumptions---disguised
-as objectivity---that lead frequentists to set the parameters of \(M\)
-by maximizing \(p(D∣M)\), under the false premise that this provides the
-best possible representation of the model. In essence, this fallacy
-leads to a confusion between the sampling probability \(p(D∣M)\) and the
-inferential probability \(p(M∣D)\); the latter when correctly obtained
-quantifies the plausibility of the hypothesis represented by \(M\) -
-precisely what we seek. The historical dimensions of this frequentist
-illusion are fascinating in their own right, and we invite interested
-readers to consult works such as McGrayne (\textbf{mcgrayne?}) and
-Clayton (\textbf{clayton?}).
+Regrettably, the development of traditional ocean color algorithms
+relies on a fundamental statistical error. This error originates in the
+conflation of empirical frequencies --- derived from a limited, noisy
+set of observations --- with the theoretical relative frequencies that
+would be observed over an infinite number of trials, thereby ignoring
+the intrinsic uncertainty and variability inherent in finite sampling.
+The fallout has and continues to be extremely detrimental to
+observational science. Traditional, frequentist approaches, where the
+variability in sample frequencies is treated as negligible, yield
+overconfident inferences, and allegedgly significant findings that do
+not hold water, in time leading to replication crises.
 
-Bayesianism, Frequentism's archrival, provides the framoework to
-establish the plausibility of \(M\) using observations through the
-reinterpretation of how probability should be construed. As elaborated
-extensively by Jaynes (\textbf{jaynes2003probability?}), and more
-recently by practitioners like Gelman (\textbf{gelman2013bayesian?}),
-and McElreath (\textbf{mcelreath2015statistical?}), probability is best
-understood as a statement of our knowledge rather than as an inherent
-property of the data. the Bayesian perspective treats the data, once
-observed, as fixed and the model parameters as latent random variables
-the uncertainty of which can---and should---be explicitly modeled via
-prior distributions. Prior distributions are probabilistic constructs
-that help workers encode prior knowledge. Priors can represents
-knowledge elicited from subject matter experts, assumptions held by the
-modeler or a complete absence thereof.
+In statistical modeling, the goal is to assess how well a hypothesized
+model \(M\) represents the relationship between independent and
+dependent variables based on collected data \(D\). Frequentist methods
+focus exclusively on the likelihood \(p(D|M)\), estimating model
+parameters by maximizing this probability. This approach assumes that,
+with a sufficiently large dataset, the chosen parameters reliably
+reflect the true system. However, in practice, any finite dataset
+carries inherent uncertainty and noise. As a consequence, models
+developed under this framework tend to be overconfident - yielding point
+estimates that do not adequately capture the variability of
+\emph{in-situ} observations - and may not generalize well to new data.
+Ad hoc attempts to generate uncertainty, such as bootstrapping, further
+compound the problem by resampling the flawed, finite observations and
+thus amplifying any existing biases. This overconfidence, coupled with
+unverified assumptions presented as objectivity, can lead to misleading
+inferences and contribute to replication challenges. The historical
+process by which these assumptions have been enshrined is fascinating in
+its own right, and we invite interested readers to consult works such as
+McGrayne (\textbf{mcgrayne?}) and Clayton (\textbf{clayton?}).
 
-This conceptual framework dissolves the false dichotomy between
-empirical and semi‑analytical approaches. Both are, in essence,
-different hypotheses or approximations about how ocean color data are
-generated; they merely provide alternative expressions of the likelihood
-\(p(Chl | Rrs)\) to be updated in light of new data. By adopting a
-Bayesian paradigm, our approach integrates prior knowledge with observed
-data to yield a full posterior predictive distribution---offering not
-only improved predictive performance but also principled uncertainty
-quantification. In doing so, it circumvents the need for the
-proliferation of narrowly tuned, sensor-specific variants (65 versions
-for 25 sensors, as documented by O'Reilly and Werdell
-(\textbf{OREILLY2019?})) that are symptomatic of ad hoc calibration
-processes. Detailed methodological comparisons and additional model
-refinements are provided in the Supplementary Material.
+As elaborated extensively by Jaynes (\textbf{jaynes2003probability?}),
+and more recently by practitioners like Gelman
+(\textbf{gelman2013bayesian?}), and McElreath
+(\textbf{mcelreath2015statistical?}), probability is best understood as
+a degree of knowledge rather than as a long-run frequency. Accordingly
+in this Bayesian framework, the prior \(p(M)\) encodes our current state
+of knowledge about the model's parameters before \(D\) is observed. This
+knowledge can be elicited from experts, or if none exixts weakly
+informative priors can be used. Note that whether \(M\) is based on
+empirically observed relationships, analytical understanding, or
+something in between is irrelevant to its integration into this
+framework. Once the data \(D\) are collected, Bayes' theorem combines
+the likelihood \(p(D|M)\) with the prior \(p(M)\) to yield the posterior
+distribution \(p(M|D)\), which represents an updated level of knowledge
+about the model's parameters. Crucially this posterior can then serve as
+the new prior when new data become available, allowing for sequential
+integration of new information. The by-product of this approach is
+uncertainty estimates that more accurately reflect the complexities of
+the underlying system as currently understood. In effect, rather than
+relying on single point estimates based on \(p(D|M)\), the posterior
+\(p(M|D)\) can be mined to retrieve a range of plausible parameter
+values through credible intervals offering a more transparent and robust
+basis for inference. This principled integration of prior knowledge with
+observed data is central to our approach, as detailed further in the
+Supplementary Material.
+
+In this paper we recast a number of OCx-like algorithms in the Bayesian
+framework. We also explore a different kind of
 
@@
 
-Statement of Contribution In this study, we develop and demonstrate a
-new global chlorophyll retrieval model based on Bayesian Additive
-Regression Trees implemented in PyMC. We train the BART model on a
-large, standardized dataset of satellite remote-sensing reflectance
-(Rrs) spectra matched with in situ chlorophyll measurements, using
-log₁₀-transformed Chl-a as the response to stabilize variance. The
-resulting model is applied globally to produce chlorophyll-a estimates
-from multi-spectral satellite data, with associated uncertainty
-estimates for each prediction. We show that this Bayesian tree-based
-model can serve as a general-purpose ocean color algorithm that is
-sensor-agnostic (provided reflectances are harmonized to common
-wavebands), interpretable, and uncertainty-aware. Unlike conventional
-empirical algorithms, the BART approach allows users to examine the
-inferred Rrs--Chl relationships and trust the model's performance across
-regimes, while also quantifying confidence in each retrieval. This work
-thus contributes a novel methodological advance to satellite ocean color
-science: a unified chlorophyll retrieval model that marries the
-strengths of empirical algorithms (global applicability and simplicity)
-with the benefits of modern Bayesian machine learning (flexibility,
-interpretability, and rigorous uncertainty quantification). Our
-introduction of BART for global chlorophyll prediction opens the door
-for more robust monitoring of ocean biogeochemistry and improved
-integration of ocean color data into scientific and management
-applications.
+\section{Data Preprocessing}
+
+Data for this study were acquired from multiple satellite ocean color
+sensors and corresponding in situ chlorophyll-(a) measurements obtained
+from sources such as the NASA Bio-Optical Marine Algorithm Data set
+(NOMAD) and the compilation by Valente et al.~(2015). To ensure
+consistency across sensors, the spectral reflectance data ((R\_\{rs\}))
+were interpolated as needed to common wavelength centers.
+
+For the empirical (OCx) formulation, blue-to-green band ratios were
+computed for each observation. In particular, the maximum band ratio
+(MBR) was determined by taking the highest value among the available
+blue-band ratios (e.g., (Rrs(443)/Rrs(555)), (Rrs(490)/Rrs(555)), and
+(Rrs(510)/Rrs(555))). This maximum value was then log-transformed: \[
+\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
+\]
+
+For the Color Index (CI) formulation of Hu et al.~(2012), the CI was
+calculated as: \[
+\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
+\] and the corresponding in situ chlorophyll-(a) concentrations were
+log-transformed: \[
+\log \text{Chl} = \log_{10}(\text{Chl}).
+\]
+
+These transformations standardize the data to a common scale, ensuring
+that variability is appropriately captured for subsequent regression and
+uncertainty quantification. Detailed descriptions of the interpolation
+methods and quality control procedures are provided in the Supplementary
+Material.
+
+\section{Statement of Contribution}\label{statement-of-contribution}
+
+In this study, we develop and demonstrate a new global chlorophyll
+retrieval model based on Bayesian Additive Regression Trees implemented
+in PyMC. We train the BART model on a large, standardized dataset of
+satellite remote-sensing reflectance (Rrs) spectra matched with in situ
+chlorophyll measurements, using log₁₀-transformed Chl-a as the response
+to stabilize variance. The resulting model is applied globally to
+produce chlorophyll-a estimates from multi-spectral satellite data, with
+associated uncertainty estimates for each prediction. We show that this
+Bayesian tree-based model can serve as a general-purpose ocean color
+algorithm that is sensor-agnostic (provided reflectances are harmonized
+to common wavebands), interpretable, and uncertainty-aware. Unlike
+conventional empirical algorithms, the BART approach allows users to
+examine the inferred Rrs--Chl relationships and trust the model's
+performance across regimes, while also quantifying confidence in each
+retrieval. This work thus contributes a novel methodological advance to
+satellite ocean color science: a unified chlorophyll retrieval model
+that marries the strengths of empirical algorithms (global applicability
+and simplicity) with the benefits of modern Bayesian machine learning
+(flexibility, interpretability, and rigorous uncertainty
+quantification). Our introduction of BART for global chlorophyll
+prediction opens the door for more robust monitoring of ocean
+biogeochemistry and improved integration of ocean color data into
+scientific and management applications.
 

### 2025-04-11 manuscript/manuscript.qmd
updated manuscript

@@
     roles:
-      - Investigation
+      - Investigator
       - Software
       - Visualization
+      - Project administration
     affiliations:
-      - Consultant
+      - Independant Consultant
   - name: Susanne E. Craig
@@
     roles:
-      - Investigation
-      - Project administration
+      - Co-Investigator
     affiliations:
@@
 key-points:
-  - Key Points convey the main points and conclusions of the article. 
+  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
   - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
@@
 
+#### Overcoming limitations
 Some researchers have attempted to deal with these issues.
@@
 [@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
-Yet most of these approaches retain variable levels of frequentism by applying only part of what is commonly referred to as the Bayesian workflow[@betancourt2019].
+Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].
 
-In this paper we propose a modified workflow to Bayesian modeling for Ocean Color emote sensing.
-We propose that this workflow include; (1) one or more model building, (2) prior formulation via expert knowledge elicitation and/or data-free simulation, (3) fitting model to data, (4) model validation, comparison and selection, (5) interpretation of results via posterior analysis, (6) prediction on new observation, (7) update the posterior when new labeled data becomes available.
-As illustration we recast some of the pre-existing $OCx$ models into their probabilistic version.
-To illustrate model comparison and selection we also build and evaluate Bayesian Additive Tree Regregssion  (BART; [@chipman2010bart]) type model.
+#### In search of the Ocean Color Bayesian workflow
+The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g.
+[@mcelreath_stat_rethink]).
+That is not to say that all steps are applicable to all fields of research and all intents.
+The workflow can and should be tailored to the researcher's field.
+[@wolkovich2024fourstepbayesianworkflowimprovin] has proposed a four-step process for Ecological modeling.
+However there are some core steps that should be followed.
+(1) Begin with one, or better yet,  more than one conceptual models and code them up.
+Model building include prior formulation to encode existing knowledge.
+Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvemnt.
+(2) Check assumptions by a simulation process known as *Prior Predictive Checks*.
+Bayesian models are generative, meaning they can run on empty, that is produce results without data.
+This allows a sanity check of the built-in assumptions.
+Non-sensical simulation output indicate assumptions must be revisited.
+(3) Collect data and conduct exploratory data analysis.
+Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features.
+(4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information.
+Most modern packages (e.g.
+Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights.
+(5) Model compasison and selection.
+Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights.
+(6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.
+
+In this paper, we largely follow this workflow.
+We begin by leveraging some of the work eluded to above and  recast some of the pre-existing $OCx$ models into their probabilistic version.
+We also offer an alternative in the form of Bayesian Additive Regression Trees  (BART; [@chipman2010bart]).
 

### 2025-04-11 manuscript/manuscript.tex
updated manuscript

@@
 monitoring marine ecosystems, as they enable the global estimation of
-chlorophyll‑a---a key indicator of phytoplankton biomass and ocean
-productivity. Early empirical algorithms, notably developed by O'Reilly
-et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al., 2000),
-established the \(OCx\) family (where \(x\) denotes the number of bands
-used) of polynomial regression models. These models relate blue-to-green
-reflectance ratios (after log‑transformation) to in situ chlorophyll‑a,
-employing either straight band ratios (BR) or maximum band ratios
-(MBR)---the latter selecting the highest available blue-to-green ratio
-for any given observation as input to a high‑order polynomial. These
-formulations have served as the operational foundation for chlorophyll‑a
-products across a broad range of satellite ocean color sensors---from
-the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS,
-and MERIS to more recent missions---offering a straightforward and
-robust approach for Case‑1 waters. However, their performance is more
-limited in optically complex Case‑2 waters and remains sensitive to
-atmospheric correction errors.
+chlorophyll‑a (\(Chl_a\)) ---a key indicator of phytoplankton biomass
+and ocean productivity. Early empirical algorithms, notably developed by
+O'Reilly et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al.,
+2000), established the \(OCx\) family (where \(x\) denotes the number of
+bands used) of polynomial regression models. These models relate
+blue-to-green reflectance ratios (after log‑transformation) to in situ
+\(Chl_a\), employing either straight band ratios (BR) or maximum band
+ratios (MBR)---the latter selecting the highest available blue-to-green
+ratio for any given observation as input to a high‑order polynomial.
+These formulations have served as the operational foundation for
+chlorophyll‑a products across a broad range of satellite ocean color
+sensors---from the pioneering Coastal Zone Color Scanner (CZCS) through
+SeaWiFS, MODIS, and MERIS to more recent missions---offering a
+straightforward and robust approach for Case‑1 waters. However, their
+performance is more limited in optically complex Case‑2 waters and
+remains sensitive to atmospheric correction errors.
 
@@
 traditional denominator with the mean of the 555 and 670\,nm
-reflectances, improving the dynamic range at low chlorophyll
-concentrations. In total, (John E. O'Reilly \& Werdell, 2019) propose 65
-versions of BR/MBR \(OCx\) type algorithms for 25 sensors---on average,
-two or more variants per sensor. With this arsenal, it is hoped,
-researchers are better equipped to address the wide array of bio‑optical
-environments encountered in global ocean color applications.
+reflectances, with the aim of improving the dynamic range at low
+chlorophyll concentrations. In total, (John E. O'Reilly \& Werdell,
+2019) propose 65 versions of BR/MBR \(OCx\) type algorithms for 25
+sensors---on average, two or more variants per sensor. With this
+arsenal, it is hoped, researchers are better equipped to address the
+wide array of bio‑optical environments encountered in global ocean color
+applications.
 
@@
 
-Regrettably, the development of traditional ocean color algorithms
-relies on a fundamental statistical error. This error originates in the
-conflation of empirical frequencies --- derived from a limited, noisy
-set of observations --- with the theoretical relative frequencies that
-would be observed over an infinite number of trials, thereby ignoring
-the intrinsic uncertainty and variability inherent in finite sampling.
-The fallout has and continues to be extremely detrimental to
-observational science. Traditional, frequentist approaches, where the
-variability in sample frequencies is treated as negligible, yield
-overconfident inferences, and allegedgly significant findings that do
-not hold water, in time leading to replication crises.
+Regrettably, the development of traditional ocean color predictive
+models relies on a fundamental statistical error that plagues most of
+observational science today; that of conflating sampling probability,
+with inferential probability. Consider data \(D\) that include
+predictive features (e.g.~Remote sensing reflectance - Rrs) and
+prediction targets (e.g.~\(Chl_a\) or phytoplankton absorption), and a
+model \(M\) (e.g.~OCx) hypothesized to adequately represent the
+statistical association between the two. The sampling probability
+\(p(D|M)\) also referred to as the likelihood, is the probability of
+\(D\) conditioned on \(M\) being ``true''. The common model fitting
+practice is to maximize \(p(D|M)\) - likelihood maximization - achieved
+by tuning \(M\)'s parameters. The unspoken assumption is that making
+\(M\) fit \(D\) as well as possible yields the best version of \(M\)
+given \(D\) i.e.~maximizing \(p(M|D)\), the inferential probability.
+This sneaky swapping from \(p(D|M)\) to \(p(M|D)\) is a violation of the
+rules of conditional probability, Bayes' theorem in particular.
+Jaynes(\textbf{jaynes2003?}) and later Clayton through his highly
+readable account (\textbf{clayton?}) are some of the more notable
+whistleblowers. Occasionally, in simple or well-behaved problems, with a
+large amount of data and no prior information available the maxima of
+both \(p(M|D)\) and \(p(D|M)\) can indeed converge on the same answer.
+Yet there is no guarantee predictability for this outcome however, and
+the fallout of this abuse is non-negligible, most noticeably in terms of
+models that don't generalize well, ad-hoc uncertainty estimation
+attempts, and more broadly a replication crisis.
 
-In statistical modeling, the goal is to assess how well a hypothesized
-model \(M\) represents the relationship between independent and
-dependent variables based on collected data \(D\). Frequentist methods
-focus exclusively on the likelihood \(p(D|M)\), estimating model
-parameters by maximizing this probability. This approach assumes that,
-with a sufficiently large dataset, the chosen parameters reliably
-reflect the true system. However, in practice, any finite dataset
-carries inherent uncertainty and noise. As a consequence, models
-developed under this framework tend to be overconfident - yielding point
-estimates that do not adequately capture the variability of
-\emph{in-situ} observations - and may not generalize well to new data.
-Ad hoc attempts to generate uncertainty, such as bootstrapping, further
-compound the problem by resampling the flawed, finite observations and
-thus amplifying any existing biases. This overconfidence, coupled with
-unverified assumptions presented as objectivity, can lead to misleading
-inferences and contribute to replication challenges. The historical
-process by which these assumptions have been enshrined is fascinating in
-its own right, and we invite interested readers to consult works such as
-McGrayne (\textbf{mcgrayne?}) and Clayton (\textbf{clayton?}).
+Some researchers have attempted to deal with these issues. (Seegers et
+al., 2018) e.g.~have proposed alternative metrics to circumvent the
+inadequate assumptions of the frequentist approach. Others have tried to
+go a step futher and switch to the Bayesian paradigm. E.g. (Frouin \&
+Pelletier, 2013) have proposed a Bayesian inversion scheme for
+atmospheric correction. (\textbf{shi2015?}) proposed a probabilistic
+method to merge data from different sensors. (Craig \& Karaköylü, 2019)
+have proposed a Bayesian neural network approach using Hamiltonian Monte
+Carlo sampling to retrieve Inherent Optical Properties (IOP) from
+Top-of-the-atmosphere (TOA) radiance. (\textbf{werther2022?}) used
+Monte-Carlo dropout as deep neural net training method to obtain
+prediction uncertainty. (Erickson et al., 2023) have proposed using
+conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a
+forward model. (\textbf{hammout2024?}) have proposed a Bayesian Neural
+Network trained using Stochastic Variational Inference to predict
+\(Chl_a\) from ocean color observations. Yet most of these approaches
+retain variable levels of frequentism by applying only part of what is
+commonly referred to as the Bayesian workflow(\textbf{betancourt2019?}).
 
-As elaborated extensively by Jaynes (\textbf{jaynes2003probability?}),
-and more recently by practitioners like Gelman
-(\textbf{gelman2013bayesian?}), and McElreath
-(\textbf{mcelreath2015statistical?}), probability is best understood as
-a degree of knowledge rather than as a long-run frequency. Accordingly
-in this Bayesian framework, the prior \(p(M)\) encodes our current state
-of knowledge about the model's parameters before \(D\) is observed. This
-knowledge can be elicited from experts, or if none exixts weakly
-informative priors can be used. Note that whether \(M\) is based on
-empirically observed relationships, analytical understanding, or
-something in between is irrelevant to its integration into this
-framework. Once the data \(D\) are collected, Bayes' theorem combines
-the likelihood \(p(D|M)\) with the prior \(p(M)\) to yield the posterior
-distribution \(p(M|D)\), which represents an updated level of knowledge
-about the model's parameters. Crucially this posterior can then serve as
-the new prior when new data become available, allowing for sequential
-integration of new information. The by-product of this approach is
-uncertainty estimates that more accurately reflect the complexities of
-the underlying system as currently understood. In effect, rather than
-relying on single point estimates based on \(p(D|M)\), the posterior
-\(p(M|D)\) can be mined to retrieve a range of plausible parameter
-values through credible intervals offering a more transparent and robust
-basis for inference. This principled integration of prior knowledge with
-observed data is central to our approach, as detailed further in the
-Supplementary Material.
+In this paper we propose a modified workflow to Bayesian modeling for
+Ocean Color emote sensing. We propose that this workflow include; (1)
+one or more model building, (2) prior formulation via expert knowledge
+elicitation and/or data-free simulation, (3) fitting model to data, (4)
+model validation, comparison and selection, (5) interpretation of
+results via posterior analysis, (6) prediction on new observation, (7)
+update the posterior when new labeled data becomes available. As
+illustration we recast some of the pre-existing \(OCx\) models into
+their probabilistic version. To illustrate model comparison and
+selection we also build and evaluate Bayesian Additive Tree Regregssion
+(BART; (Chipman et al., 2010)) type model.
 
-In this paper we recast a number of OCx-like algorithms in the Bayesian
-framework. We also explore a different kind of
+\subsection{Methods}\label{methods}
 
-\subsubsection{Motivation for Bayesian Tree-Based
-Models}\label{motivation-for-bayesian-tree-based-models}
+\subsubsection{Prior Elicitation and Model
+Formulation}\label{prior-elicitation-and-model-formulation}
 
-To address these limitations, we explore Bayesian tree-based modeling --
-specifically, Bayesian Additive Regression Trees (BART) -- as a novel
-approach for ocean color chlorophyll retrieval. BART offers several
-advantages for this application:
+\subsubsection{Data Preprocessing}\label{data-preprocessing}
 
-Direct Bayesian inference: As a Bayesian ensemble method, BART directly
-targets the posterior distribution \(p(Chl | Rrs)\) rather than a purely
-empirical fit. The model inherently provides a probabilistic mapping
-from reflectance to chlorophyll. Rather than producing a single best-fit
-value, BART returns a posterior mean and credible interval for each
-prediction, naturally quantifying predictive uncertainty.
-
-Uncertainty quantification: The BART framework includes uncertainty in
-its predictions by design. Every estimated chlorophyll value comes with
-a confidence range (e.g., 95\% credible interval) reflecting model
-uncertainty and observational noise. This is a key improvement over
-deterministic algorithms, enabling more reliable use of the data for
-scientific and operational purposes where understanding uncertainty is
-crucial e.g.~for forecasting trends.
-
-Interpretability: Being a sum-of-trees model, BART can be interrogated
-to understand the influence of each input variable. Tools such as
-partial dependence plots allow us to visualize the relationship between
-each spectral band and the predicted chlorophyll, marginalizing over
-other bands. In contrast to high-order polynomial coefficients, these
-tree-based partial dependencies are intuitively interpretable. For
-instance, one can observe how changes in a particular waveband's
-reflectance (holding others average) drive the chlorophyll estimate,
-revealing any thresholds or non-linear responses.
-
-BART's modeling approach is non-parametric and highly flexible, which
-means it can capture complex, non-linear relationships between
-multi-spectral Rrs inputs and chlorophyll without imposing a rigid
-functional form. At the same time, its Bayesian regularization -- which
-shrinks the contribution of each individual tree -- helps prevent
-overfitting even with a rich model structure. This balance between
-flexibility and regularization is well-suited to global ocean color
-data, where the true Rrs--Chl relationship is known to be non-linear and
-context-dependent, yet a model must generalize across diverse water
-types.
-
-The interpretability of the BART model is illustrated by partial
-dependence analysis on the six input Rrs bands (centered approximately
-at 412, 443, 490, 510, 555, and 670 nm). The partial dependence plots
-reveal nuanced spectral--chlorophyll relationships and clear transitions
-in the dominant wavelength influence across different chlorophyll
-regimes. For example, the shorter blue wavelengths (412--443 nm) show a
-strong inverse influence on predicted \(log(Chl_a)\) (higher blue
-reflectance corresponds to lower chlorophyll) that gradually saturates
-in the most oligotrophic waters, consistent with the leveling off of
-blue-to-green ratios at very low Chl. The green band (\textasciitilde555
-nm) exhibits a more complex, non-monotonic effect: at low reflectance
-(clear water) its influence on the Chl estimate is minimal, but it
-becomes increasingly important through intermediate reflectance ranges
--- reflecting the band-difference signal exploited by the CI algorithm
--- and then diminishes for very high chlorophyll where green reflectance
-tends to flatten. In contrast, the red band (\textasciitilde670 nm) has
-virtually no effect on the model's predictions until a threshold is
-reached at elevated chlorophyll concentrations, after which its
-influence steeply increases. This behavior aligns with optical
-expectations, since red reflectance is negligible in low-Chl waters but
-rises sharply once phytoplankton absorption in the blue saturates and
-biomass is high. These partial dependence results indicate that the BART
-model automatically learns the piecewise spectral logic that
-oceanographers often handle via separate algorithms (blue/green ratios
-for low-to-moderate Chl, red bands for high Chl). The ability to capture
-such regime-dependent responses in a single unified model, and to
-visualize them, underscores the interpretability and scientific insight
-provided by the BART approach.
+\subsubsection{}\label{section}
 
@@
 
+\bibitem[\citeproctext]{ref-chipman2010bart}
+Chipman, H. A., George, E. I., \& McCulloch, R. E. (2010). BART:
+Bayesian additive regression trees. \emph{The Annals of Applied
+Statistics}, \emph{4}(1), 266--298.
+\url{https://doi.org/10.1214/09-AOAS285}
+
+\bibitem[\citeproctext]{ref-craig2019}
+Craig, S. E., \& Karaköylü, E. M. (2019). Bayesian models for deriving
+biogeochemical information from satellite ocean color.
+\emph{EarthArXiv}. \url{https://doi.org/10.31223/osf.io/shp6y}
+
+\bibitem[\citeproctext]{ref-erickson2023}
+Erickson, Z. K., McKinna, L. I. W., Werdell, P. J., \& Cetinić, I.
+(2023). Bayesian approach to a generalized inherent optical property
+model. \emph{Optics Express}, \emph{31}, 22790--22801.
+\url{https://doi.org/10.1364/oe.486581}
+
+\bibitem[\citeproctext]{ref-frouin2013}
+Frouin, R., \& Pelletier, B. (2013). Bayesian methodology for ocean
+color remote sensing. Retrieved from
+\url{https://hal.archives-ouvertes.fr/hal-00822032}
+
 \bibitem[\citeproctext]{ref-hu2012novel}
@@
 
+\bibitem[\citeproctext]{ref-seegers2018}
+Seegers, B. N., Stumpf, R. P., Schaeffer, B. A., Loftin, K. A., \&
+Werdell, P. J. (2018). Performance metrics for the assessment of
+satellite data products: An ocean color case study. \emph{Opt. Express},
+\emph{26}(6), 7404--7422. \url{https://doi.org/10.1364/OE.26.007404}
+
 \end{CSLReferences}

### 2025-04-18 manuscript/manuscript.qmd
Cleaned introduction

@@
 
-### Introduction
+## Introduction
 
-#### Historical Context of Chlorophyll Algorithms
-Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a ($Chl_a$) —a key indicator of phytoplankton biomass and ocean productivity.
+### Historical Context of Chlorophyll Algorithms
+Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity.
 Early empirical algorithms, notably developed by O’Reilly et al.
@@
 
-#### Limitations of Existing Approaches
-Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability.
-Consider data $D$ that include predictive features (e.g.
-Remote sensing reflectance - Rrs) and prediction targets (e.g.
-$Chl_a$ or phytoplankton absorption), and a model $M$ (e.g.
-OCx) hypothesized to adequately represent the statistical association between the two.
-The sampling probability $p(D|M)$ also referred to as the likelihood, is the probability of $D$ conditioned on $M$ being "true".
-The common model fitting practice is to maximize $p(D|M)$ - likelihood maximization - achieved by tuning $M$'s parameters.
-The unspoken assumption is that making $M$ fit $D$ as well as possible yields the best version of $M$ given $D$ i.e.
-maximizing $p(M|D)$, the inferential probability.
-This sneaky swapping from $p(D|M)$ to $p(M|D)$ is a violation of the rules of conditional probability, Bayes' theorem in particular.
-Jaynes[@jaynes2003] and later Clayton through his highly readable account [@clayton] are some of the more notable whistleblowers.
-Occasionally, in simple  or well-behaved problems, with a large amount of data and no prior information available the maxima of both $p(M|D)$ and $p(D|M)$ can indeed converge on the same answer.
-Yet there is no guarantee predictability for this outcome however, and the fallout of this abuse is non-negligible, most noticeably in terms of models that don't generalize well, ad-hoc uncertainty estimation attempts, and more broadly a replication crisis.
+### Limitations of Existing Approaches
+Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g.
+Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them.
+The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true".
+Standard model fitting maximizes this quantity (i.e.
+the likelihood) by tuning the parameters of $M$.
 
-#### Overcoming limitations
-Some researchers have attempted to deal with these issues.
-[@seegers2018] e.g.
-have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
-Others have tried to go a step futher and switch to the Bayesian paradigm.
+The unspoken assumption is that the model that best fits the data also best represents the underlying process.
+This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem.
+While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule.
+The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.
+
+In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]).
+These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well.
+In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.
+
+This epistemic shortcut has already been explicitly critiqued in the machine learning community.
+Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference.
+This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs.
+Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.
+
+### Overcoming limitations
+There have been attempts to address these issues.
+[@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
+Others have tried to go a step futher incorporate Bayesian concepts.
 E.g.
@@
 [@shi2015] proposed a probabilistic method to merge data from different sensors.
-[@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
-[@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty.
-[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model.
-[@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
+[@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
+[@werther2022] used Monte-Carlo dropout to approximate a deep BNN.
+[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty.
+[@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
 Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].
 
-#### In search of the Ocean Color Bayesian workflow
+### In search of our Bayesian workflow
 The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g.
@@
 The workflow can and should be tailored to the researcher's field.
-[@wolkovich2024fourstepbayesianworkflowimprovin] has proposed a four-step process for Ecological modeling.
+[@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for Ecological modeling.
 However there are some core steps that should be followed.
@@
 Model building include prior formulation to encode existing knowledge.
-Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvemnt.
+Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement.
 (2) Check assumptions by a simulation process known as *Prior Predictive Checks*.
@@
 This allows a sanity check of the built-in assumptions.
-Non-sensical simulation output indicate assumptions must be revisited.
+Non-sensical simulation results indicate assumptions must be revisited.
 (3) Collect data and conduct exploratory data analysis.
@@
 
-In this paper, we largely follow this workflow.
-We begin by leveraging some of the work eluded to above and  recast some of the pre-existing $OCx$ models into their probabilistic version.
-We also offer an alternative in the form of Bayesian Additive Regression Trees  (BART; [@chipman2010bart]).
+In this paper we leverage past work and recast some of the pre-existing $OCx$ models into their probabilistic version.
+We also propose alternative models, namely Bayesian Additive Regression Trees (BART, [@chipman2010bart]) as a robust flexible method that has been used for diverse application ([@linero2017].)
 
-### Methods
+## Methods
 
-#### Prior Elicitation and Model Formulation
+### Prior Elicitation and Model Formulation
 
-#### Data Preprocessing
+### Data Preprocessing
 
-####
+###
 

### 2025-04-18 manuscript/manuscript.tex
Cleaned introduction

@@
 \authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
-\affiliation{1}{Consultant, }\affiliation{2}{NASA, }
+\affiliation{1}{Independant Consultant, }\affiliation{2}{NASA, }
 \correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}
@@
 
-\subsection{Introduction}\label{introduction}
+\section{Introduction}\label{introduction}
 
-\subsubsection{Historical Context of Chlorophyll
+\subsection{Historical Context of Chlorophyll
 Algorithms}\label{historical-context-of-chlorophyll-algorithms}
@@
 
-\subsubsection{Limitations of Existing
+\subsection{Limitations of Existing
 Approaches}\label{limitations-of-existing-approaches}
@@
 observational science today; that of conflating sampling probability,
-with inferential probability. Consider data \(D\) that include
-predictive features (e.g.~Remote sensing reflectance - Rrs) and
-prediction targets (e.g.~\(Chl_a\) or phytoplankton absorption), and a
-model \(M\) (e.g.~OCx) hypothesized to adequately represent the
-statistical association between the two. The sampling probability
-\(p(D|M)\) also referred to as the likelihood, is the probability of
-\(D\) conditioned on \(M\) being ``true''. The common model fitting
-practice is to maximize \(p(D|M)\) - likelihood maximization - achieved
-by tuning \(M\)'s parameters. The unspoken assumption is that making
-\(M\) fit \(D\) as well as possible yields the best version of \(M\)
-given \(D\) i.e.~maximizing \(p(M|D)\), the inferential probability.
-This sneaky swapping from \(p(D|M)\) to \(p(M|D)\) is a violation of the
-rules of conditional probability, Bayes' theorem in particular.
-Jaynes(\textbf{jaynes2003?}) and later Clayton through his highly
-readable account (\textbf{clayton?}) are some of the more notable
-whistleblowers. Occasionally, in simple or well-behaved problems, with a
-large amount of data and no prior information available the maxima of
-both \(p(M|D)\) and \(p(D|M)\) can indeed converge on the same answer.
-Yet there is no guarantee predictability for this outcome however, and
-the fallout of this abuse is non-negligible, most noticeably in terms of
-models that don't generalize well, ad-hoc uncertainty estimation
-attempts, and more broadly a replication crisis.
+with inferential probability(Scheemaekere \& Szafarz (2011).) Consider a
+dataset \(D\) that consists of input-output pairs - e.g.~Remote sensing
+reflectance (Rrs) and chlorophyll-\emph{a} concentration \(Chl_a\) - and
+a model \(M\), such as OCx, hypothesized to describe the statistical
+association between them. The sampling probability \(p(D|M)\) represents
+the probability of observing the data \(D\) if model \(M\) were
+``true''. Standard model fitting maximizes this quantity (i.e.~the
+likelihood) by tuning the parameters of \(M\).
 
-Some researchers have attempted to deal with these issues. (Seegers et
-al., 2018) e.g.~have proposed alternative metrics to circumvent the
-inadequate assumptions of the frequentist approach. Others have tried to
-go a step futher and switch to the Bayesian paradigm. E.g. (Frouin \&
-Pelletier, 2013) have proposed a Bayesian inversion scheme for
-atmospheric correction. (\textbf{shi2015?}) proposed a probabilistic
-method to merge data from different sensors. (Craig \& Karaköylü, 2019)
-have proposed a Bayesian neural network approach using Hamiltonian Monte
-Carlo sampling to retrieve Inherent Optical Properties (IOP) from
-Top-of-the-atmosphere (TOA) radiance. (\textbf{werther2022?}) used
-Monte-Carlo dropout as deep neural net training method to obtain
-prediction uncertainty. (Erickson et al., 2023) have proposed using
-conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a
-forward model. (\textbf{hammout2024?}) have proposed a Bayesian Neural
-Network trained using Stochastic Variational Inference to predict
-\(Chl_a\) from ocean color observations. Yet most of these approaches
-retain variable levels of frequentism by applying only part of what is
-commonly referred to as the Bayesian workflow(\textbf{betancourt2019?}).
+The unspoken assumption is that the model that best fits the data also
+best represents the underlying process. This constitutes an epistemic
+fallacy; substituting \(p(D|M)\) for \(p(M|D)\), in a violation of the
+rules of conditional probability inversion as framed by Bayes' theorem.
+While in well-behaved data-rich problems, the maxima of both expressions
+may coincide, this is an exception, not the rule. The consequences of
+this mistake are far-reaching as argued by (Clayton, 2022), with
+implications extending from scientific modeling to medicine, law, and
+public policy.
 
-In this paper we propose a modified workflow to Bayesian modeling for
-Ocean Color emote sensing. We propose that this workflow include; (1)
-one or more model building, (2) prior formulation via expert knowledge
-elicitation and/or data-free simulation, (3) fitting model to data, (4)
-model validation, comparison and selection, (5) interpretation of
-results via posterior analysis, (6) prediction on new observation, (7)
-update the posterior when new labeled data becomes available. As
-illustration we recast some of the pre-existing \(OCx\) models into
-their probabilistic version. To illustrate model comparison and
-selection we also build and evaluate Bayesian Additive Tree Regregssion
-(BART; (Chipman et al., 2010)) type model.
+In science, this fallacy contributes to models fail to generalize well,
+an impetus to produce ad-hoc or post-hoc uncertainty quantification, and
+yield published results that resist replication(Cobey et al. (2024)).
+These limiations are not confined to traditional hypothesis testing;
+they are embedded in the training.; they are embedded in the training
+and deployment of modern machine learning models as well. In regression
+and classification tasks, likelihood maximization is treated as
+sufficient for inference, even though it results in a single point
+estimate and fails to account for parameter uncertainty or model
+plausibility.
 
-\subsection{Methods}\label{methods}
+This epistemic shortcut has already been explicitly critiqued in the
+machine learning community. Gal (Gal, 2016) and Ghahramani
+(\textbf{gahramani2015probabilistic?}) emphasize that most ML models
+discard uncertainty entirely and treat the output of optimization as
+inference. This leads to overconfident predctions and brittle
+generalization, echoing Clayton's concerncs. Bishop (Bishop, 2006) also
+distiguishes between the utility of models and the inferential core
+required to quantify uncertainty, underscoring that likelihood is not
+enough and that the Bernoulli fallacy pervades the very logic of applied
+machine learning.
 
-\subsubsection{Prior Elicitation and Model
+\subsection{Overcoming limitations}\label{overcoming-limitations}
+
+There have been attempts to address these issues. (Seegers et al., 2018)
+have proposed alternative metrics to circumvent the inadequate
+assumptions of the frequentist approach. Others have tried to go a step
+futher incorporate Bayesian concepts. E.g. (Frouin \& Pelletier, 2013)
+have proposed a Bayesian inversion scheme for atmospheric correction.
+(\textbf{shi2015?}) proposed a probabilistic method to merge data from
+different sensors. (Craig \& Karaköylü, 2019) have proposed a Bayesian
+neural network (BNN) approach using Hamiltonian Monte Carlo sampling to
+retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere
+(TOA) radiance. (\textbf{werther2022?}) used Monte-Carlo dropout to
+approximate a deep BNN. (Erickson et al., 2023) have proposed using
+conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian
+model to predict IOPs with uncertainty. (\textbf{hammout2024?}) have
+proposed a BNN approximation via Stochastic Variational Inference to
+predict \(Chl_a\) from ocean color observations. Yet most of these
+approaches retain variable levels of frequentism by applying only part
+of what is now commonly referred to as the Bayesian
+workflow(\textbf{bgelman2019?}).
+
+\subsection{In search of our Bayesian
+workflow}\label{in-search-of-our-bayesian-workflow}
+
+The Bayesian workflow is a recognizable and unifying way of approaching
+statistical modeling, initially championed by statisticians at Columnbia
+University (\textbf{gelman\_bda?}) subsequently quick endorsed and
+further developed by researchers dealing with scant and noisy data (e.g.
+(\textbf{mcelreath\_stat\_rethink?})). That is not to say that all steps
+are applicable to all fields of research and all intents. The workflow
+can and should be tailored to the researcher's field. (Wolkovich et al.,
+2024) has proposed a four-step process for Ecological modeling. However
+there are some core steps that should be followed. (1) Begin with one,
+or better yet, more than one conceptual models and code them up. Model
+building include prior formulation to encode existing knowledge. Priors
+are part of the model structure and make the researcher's assumptions
+transparent, providing a first avenue of critique and improvement. (2)
+Check assumptions by a simulation process known as \emph{Prior
+Predictive Checks}. Bayesian models are generative, meaning they can run
+on empty, that is produce results without data. This allows a sanity
+check of the built-in assumptions. Non-sensical simulation results
+indicate assumptions must be revisited. (3) Collect data and conduct
+exploratory data analysis. Knowing the data will be critical to
+understanding potential problems during fitting; e..g multicollinearity
+of input features. (4) Diagnostics; the fitting process and the
+resulting posterior distribution provide rich constructs that can be
+mined for a great deal of information. Most modern packages (e.g.~Stan,
+PyMC, Numpyro, etc) offer a great many tools to extract insights. (5)
+Model compasison and selection. Often a model will perform markedly
+better, which is useful in terms predictive performance, but more
+importantly in terms of generating insights. (6) Sequential update; a
+model's output, the posterior distribution becomes the new prior every
+time new data becomes available for a new sequence of model fitting.
+
+In this paper we leverage past work and recast some of the pre-existing
+\(OCx\) models into their probabilistic version. We also propose
+alternative models, namely Bayesian Additive Regression Trees (BART,
+(Chipman et al., 2010)) as a robust flexible method that has been used
+for diverse application ((Linero, 2017).)
+
+\section{Methods}\label{methods}
+
+\subsection{Prior Elicitation and Model
 Formulation}\label{prior-elicitation-and-model-formulation}
 
-\subsubsection{Data Preprocessing}\label{data-preprocessing}
+\subsection{Data Preprocessing}\label{data-preprocessing}
 
-\subsubsection{}\label{section}
+\subsection{}\label{section}
 
@@
 
+\bibitem[\citeproctext]{ref-baker2016}
+Baker, M. (2016). 1,500 scientists lift the lid on reproducibility.
+\emph{Nature}, \emph{533}(7604), 452--454.
+\url{https://doi.org/10.1038/533452a}
+
+\bibitem[\citeproctext]{ref-bishop2006pattern}
+Bishop, C. M. (2006). \emph{Pattern recognition and machine learning}.
+Springer.
+
 \bibitem[\citeproctext]{ref-chipman2010bart}
@@
 \url{https://doi.org/10.1214/09-AOAS285}
+
+\bibitem[\citeproctext]{ref-clayton2022bernoulli}
+Clayton, A. (2022). \emph{Bernoulli's fallacy: Statistical illogic and
+the crisis of modern science}. Columbia University Press. Retrieved from
+\url{https://books.google.com/books?id=BT4CzwEACAAJ}
+
+\bibitem[\citeproctext]{ref-cobey2024biomedical}
+Cobey, K. D., Ebrahimzadeh, S., Page, M. J., Thibault, R. T., Nguyen,
+P.-Y., Abu-Dalfa, F., \& Moher, D. (2024). Biomedical researchers'
+perspectives on the reproducibility of research. \emph{PLoS Biology},
+\emph{22}(11), e3002870.
 
@@
 
+\bibitem[\citeproctext]{ref-gal2016uncertainty}
+Gal, Y. (2016). \emph{Uncertainty in deep learning} (PhD thesis).
+University of Cambridge.
+
 \bibitem[\citeproctext]{ref-hu2012novel}
@@
 https://doi.org/\url{https://doi.org/10.1029/2019JC014941}
+
+\bibitem[\citeproctext]{ref-jaynes2003probability}
+Jaynes, E. T., \& Bretthorst, G. L. (2003). \emph{Probability theory:
+The logic of science}. Cambridge University Press. Retrieved from
+\url{https://books.google.com/books?id=tTN4HuUNXjgC}
+
+\bibitem[\citeproctext]{ref-linero2017}
+Linero, A. R. (2017). A review of tree-based bayesian methods.
+\emph{Communications for Statistical Applications and Methods},
+\emph{24}, 543--559. \url{https://doi.org/10.29220/CSAM.2017.24.6.543}
 
@@
 
+\bibitem[\citeproctext]{ref-DeScheemaekere2011}
+Scheemaekere, X. D., \& Szafarz, A. (2011). The inference fallacy from
+bernoulli to kolmogorov. Retrieved from
+\url{https://www.researchgate.net/publication/254450993_The_Inference_Fallacy_From_Bernoulli_to_Kolmogorov}
+
 \bibitem[\citeproctext]{ref-seegers2018}
@@
 
+\bibitem[\citeproctext]{ref-wolkovich2024fourstepbayesianworkflowimproving}
+Wolkovich, E., Davies, T. J., Pearse, W. D., \& Betancourt, M. (2024). A
+four-step bayesian workflow for improving ecological science. Retrieved
+from \url{https://arxiv.org/abs/2408.02603}
+
 \end{CSLReferences}

### 2025-04-29 manuscript/manuscript.qmd
update oc6 priors

@@
     affiliations:
-      - Independant Consultant
+      - Independant Researcher
   - name: Susanne E. Craig
@@
     affiliations:
-      - NASA
+      - NASA/UMBC
 abstract: Placeholder
 plain-language-summary: Placeholder
-keywords: []
+keywords: ['Bayesian modeling', 'Ocean Color', ]
 key-points:
   - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
-  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
-  - Key Points are not included in the word count.
+  - A Bayesian workflow is a principled guide to a fully Bayesian study
+  - 
 bibliography: references.bib
@@
 keep-tex: true
+number-sections: false
 date: last-modified
@@
 
-### Prior Elicitation and Model Formulation
+We performed the following five steps.
+(1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation and comparison.
+
+### Prior Elicitation, Model Formulation, and Prior Predictive Checks
+
+Skip to content
+You said:
+Here is my introduction to my Bayesian modeling study and the beginning of my methods section. Suggest a continuation to the prior elicitation section
+
+### Historical Context of Chlorophyll Algorithms
+Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity.
+Early empirical algorithms, notably developed by O’Reilly et al.
+[@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models.
+These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial.
+These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters.
+However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.
+
+Subsequent refinements were introduced to address these deficiencies.
+For example, Hu et al.
+[@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving].
+The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations.
+O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, with the aim of improving the dynamic range at low chlorophyll concentrations.
+In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor.
+With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.
+
+### Limitations of Existing Approaches
+Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g.
+Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them.
+The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true".
+Standard model fitting maximizes this quantity (i.e.
+the likelihood) by tuning the parameters of $M$.
+
+The unspoken assumption is that the model that best fits the data also best represents the underlying process.
+This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem.
+While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule.
+The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.
+
+In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]).
+These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well.
+In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.
+
+This epistemic shortcut has already been explicitly critiqued in the machine learning community.
+Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference.
+This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs.
+Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.
+
+### Overcoming limitations
+There have been attempts to address these issues.
+[@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
+Others have tried to go a step futher incorporate Bayesian concepts.
+E.g.
+[ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction.
+[@shi2015] proposed a probabilistic method to merge data from different sensors.
+[@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
+[@werther2022] used Monte-Carlo dropout to approximate a deep BNN.
+[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty.
+[@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
+Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].
+
+### The Bayesian workflow
+The Bayesian workflow is a recognizable and unifying approach to statistical modeling, primarily championed by statisticians at Columnbia University [@gelman_bda] subsequently endorsed and further developed by researchers dealing with scant and/or noisy data (e.g.
+[@mcelreath_stat_rethink]).
+Some researchers have modified this  researcher's field.
+[@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for ecological modeling.
+However there are some core steps that should be followed.
+(1) Begin with one, or better yet,  more than one conceptual models and code them up.
+Model building include prior formulation to encode existing knowledge.
+Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement.
+(2) Check assumptions by a simulation process known as *Prior Predictive Checks*.
+Bayesian models are generative, meaning they can run on empty, that is produce results without data.
+This allows a sanity check of the built-in assumptions.
+Non-sensical simulation results indicate assumptions must be revisited.
+(3) Collect data and conduct exploratory data analysis.
+Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features.
+(4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information.
+Most modern packages (e.g.
+Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights.
+(5) Model compasison and selection.
+Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights.
+(6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.
+
+In this paper we leverage past work and recast $OC6$ [@oreilly2019] as a Bayesian mdoel, because out of the $OCx$ family,  it integrates the most bands (i.e.
+information) into its structure.
+In addition, we also propose two more models.
+(1) Bayesian Additive Regression Trees (BART, [@chipman2010bart]); a robust flexible approach successfully used for diverse applications ([@linero2017].), and insensitive to the known multicollinearity issue between ocean color bands.
+(2) A hierarchical partial pooling linear model inspired from [@gordon1983]'s original band ratio-based linear regression, formulated to circumvent discontinuities mentioned by [@oreilly1998], while insuring maximal use of available information contained in the data set.
+
+## Methods
+
+We performed the following five steps.
+(1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation, comparison, and selection.
+
+### Prior Elicitation 
+Before fitting the Bayesian model to observational data, we performed a prior predictive evaluation based on simulated mock inputs to ensure that our model structure and prior choices were physically reasonable and aligned with established ocean color variability.
+For all 3 model types we simulated chlorophyll a with approximately log-normal distribution according to [@campbell1995].
+For $OC6$ and our hierarchical linear model we generated maximum band ratio (MBR) that roughly agree with [@oreilly2019] and references therein.
+BART prior is bases on model structured and elicitation is explained further below.
+
+#### Mock MBR and Chlorophyll Data
+
+We generated synthetic MBR inputs based on empirical knowledge of global ocean conditions.
+Specifically, the dynamic range of MBR values was assumed to span approximately 0.3 to 30, corresponding to $log_{10}(MBR)$ values between -0.5 and 1.5 [@oreilly2019].
+$log_{10}(MBR)$ values were drawn from a truncated normal distribution centered at 0.75 with a standard deviation of 0.5, ensuring denser sampling in typical oligotrophic to mesotrophic regimes while allowing coverage of extreme cases.
+The corresponding predicted $log_{10}(Chl_a)$ ⁡ values were sampled, according to the physical (detection limits) and ecological expectations of chlorophyll variability [@campbell1995] from a truncated Gaussian likelihood bounded between -3 and 3.3, corresponding to chlorophyll concentrations ranging from 0.001 to 2000 mg m⁻³.
+
+#### Model Formulation
+To implement a robust Bayesian approach, we first conducted a thorough elicitation of priors informed by domain knowledge from previously published literature, where applicable.
+The choice of distribution for the model parameters is borne out of the following assumptions.
+1. Parameter domain is $(- ∞, ∞)$ though, per [@oreilly2019], parameters are expected to be relatively close to 0
+2. Priors should be regularizing. 
+3. Parameter variance is finite.
+
+Accordingly for the $OC6$ and hierarchical models we chose weakly informative Gaussian priors.
+For the likelihood, we chose a truncated Gaussian defined over a domain The standard deviations for these priors were chosen conservatively large enough to allow for flexibility, but sufficiently tight to capture known realistic parameter ranges derived from extensive empirical validation.
+
+For the BART model, priors were selected following recommendations by [@chipman2010bart] and subsequent applied literature [@linero2017; @hill2020bayesian].
+The BART algorithm naturally encodes regularization and shrinkage through its tree structure, inherently balancing bias and variance.
+However, hyperparameters controlling tree depth and the number of trees were critical.
+We specified prior distributions for these hyperparameters based on guidelines from simulation studies and domain knowledge, starting with defaults recommended by the original authors and subsequently refining through preliminary runs and prior predictive checks.
+
+We carried out formal prior predictive checks to validate these choices.
+By generating synthetic data from the prior distributions and inspecting the resulting predictions for chlorophyll-a, we ensured that our priors neither produced unrealistic extremes nor overly constrained the model predictions.
+This procedure also allowed stakeholders with domain expertise to visually assess the suitability of our priors before fitting the model to actual data, thereby promoting transparency and interpretability.
+
+Finally, the resulting priors and model formulation and prior predictive cheks are available as a jupyter notebook in the accompanying repository for full reproducibility.
 
 ### Data Preprocessing
-
-###
 

### 2025-04-29 manuscript/manuscript.tex
update oc6 priors

@@
 \authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
-\affiliation{1}{Independant Consultant, }\affiliation{2}{NASA, }
+\affiliation{1}{Independant Researcher, }\affiliation{2}{NASA/UMBC, }
 \correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}
@@
 Satellite ocean color observations have long been fundamental for
-monitoring marine ecosystems, as they enable the global estimation of
-chlorophyll‑a (\(Chl_a\)) ---a key indicator of phytoplankton biomass
+monitoring marine ecosystems, as they enable global estimation of
+chlorophyll‑a (\(Chl_a\)) --- a key indicator of phytoplankton biomass
 and ocean productivity. Early empirical algorithms, notably developed by
@@
 \subsection{Data Preprocessing}\label{data-preprocessing}
-
-\subsection{}\label{section}
 

### 2025-05-06 manuscript/manuscript.qmd
finished data wrangling

@@
 
+Satellite ocean color remote sensing has long served as a cornerstone of marine ecosystem monitoring, offering global and synoptic coverage of surface ocean properties.
+Among these, chlorophyll-a ($Chl_a$) concentration remains a central metric, widely used as a proxy for phytoplankton biomass, primary production, and water quality.
+The retrieval of $Chl_a$ from ocean color data has evolved over decades, resulting in a diverse lineage of empirical and semi-empirical algorithms.
+The following section summarizes this historical development, which sets the stage for a critical examination of the statistical foundations underlying current approaches.
+
 ### Historical Context of Chlorophyll Algorithms
+
 Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity.
@@
 ### Limitations of Existing Approaches
-Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g.
-Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them.
-The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true".
-Standard model fitting maximizes this quantity (i.e.
-the likelihood) by tuning the parameters of $M$.
 
-The unspoken assumption is that the model that best fits the data also best represents the underlying process.
-This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem.
-While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule.
-The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.
+Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error—one that pervades much of observational science: the conflation of sampling probability with inferential probability [@jaynes2003probability; @DeScheemaekere2011].
 
-In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]).
-These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well.
-In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.
+Consider a dataset $D$ composed of input–output pairs—e.g., remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$)—and a model $M$, such as OCx, posited to represent the relationship between them.
+The sampling probability $p(D \mid M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true.
+In standard model fitting, this likelihood is maximized by adjusting the parameters of $M$ to best explain the observed data.
 
-This epistemic shortcut has already been explicitly critiqued in the machine learning community.
-Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference.
-This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs.
-Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.
+This approach tacitly assumes that the model which best fits the data also most accurately represents the underlying generative process.
+This constitutes an epistemic fallacy—treating $p(D \mid M)$ as if it were $p(M \mid D)$—a direct violation of Bayes’ theorem and the rules governing conditional probability.
+
+Although in well-behaved, data-rich cases—where the likelihood is regular, the signal strong, and the model adequately constrained—the maxima of $p(D \mid M)$ and $p(M \mid D)$ may coincide, this remains the exception—not the rule.
+
+This mistake lies at the heart of what Clayton [@clayton2022bernoulli] terms the Bernoulli Fallacy: the widespread misinterpretation of likelihood as inference, or of data-fit as belief.
+As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.
+
+In scientific modeling, this fallacy contributes to poor generalization, drives the use of ad hoc or retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate [@baker2016; @cobey2024biomedical].
+These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.
+
+In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.
+
+This epistemic shortcut has been directly critiqued in the machine learning literature.
+Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference.
+The result is overconfident predictions and brittle generalization—concerns that echo Clayton’s critique.
+
+Bishop [@bishop2006pattern] similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient—and that the Bernoulli Fallacy permeates much of applied machine learning.
 
 ### Overcoming limitations
+
 There have been attempts to address these issues.
@@
 ### Limitations of Existing Approaches
-Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g.
-Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them.
-The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true".
-Standard model fitting maximizes this quantity (i.e.
-the likelihood) by tuning the parameters of $M$.
+Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error—one that pervades much of observational science: the conflation of sampling probability with inferential probability [@jaynes2003probability; @DeScheemaekere2011].
 
-The unspoken assumption is that the model that best fits the data also best represents the underlying process.
-This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem.
-While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule.
-The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.
+Consider a dataset $D$ composed of input–output pairs—e.g., remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$), and a model $M$, such as OCx, posited to represent the relationship between them.
+The sampling probability $p(D|M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true.
+In standard model fitting, this likelihood is maximized by adjusting the parameters of $M$ to best explain the observed data.
 
-In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]).
-These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well.
-In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.
+This approach tacitly assumes that the model which best fits the data also most accurately represents the underlying generative process.
+This constitutes an epistemic fallacy—treating $p(D|M)$ as if it were $p(M | D)$; a violation of Bayes’ theorem and the rules governing conditional probability.
 
-This epistemic shortcut has already been explicitly critiqued in the machine learning community.
-Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference.
-This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs.
-Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.
+Although in well-behaved, data-rich cases—where the likelihood is regular, the signal strong, and the model adequately constrained—the maxima of $p(D|M)$ and $p(M|D)$ may coincide, this remains the exception rather than the rule.
+
+This mistake lies at the heart of what Clayton [@clayton2022bernoulli] terms the Bernoulli Fallacy: the misconception that empirical probabilities by themselves yield inferential truths.
+As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.
+
+In scientific modeling, this fallacy contributes to poor generalization, drives the use of ad hoc or retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate [@baker2016; @cobey2024biomedical].
+These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.
+
+In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.
+
+This epistemic shortcut has been directly critiqued in the machine learning literature.
+Gal [@gal2016uncertainty] and Ghahramani [@ghahramani2015probabilistic] point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference.
+The result is overconfident predictions and brittle generalization—concerns that echo Clayton’s critique.
+
+Bishop [@bishop2006pattern] similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient, and that the Bernoulli Fallacy permeates much of applied machine learning.
 

### 2025-05-16 manuscript/manuscript.qmd
updated manuscript

@@
 
-In this paper we leverage past work and recast some of the pre-existing $OCx$ models into their probabilistic version.
-We also propose alternative models, namely Bayesian Additive Regression Trees (BART, [@chipman2010bart]) as a robust flexible method that has been used for diverse application ([@linero2017].)
+In this paper we leverage past work but also propose new approaches for predicting chlorophyll from sea surface reflectance (Rrs). Overall we propose three types of models.
 
-## Methods
+## 2. Materials and Methods
 
-We performed the following five steps.
-(1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation and comparison.
+This section outlines the study region and dataset used for model development and validation, the structure and rationale for the Bayesian models, and the approach used to compare their predictive performance.
+We emphasize a subset of four representative models in the main text, while full specifications and results for all seven models are provided in the Supplementary Material.
 
-### Prior Elicitation, Model Formulation, and Prior Predictive Checks
+### 2.1 Study Area and Dataset
 
-Skip to content
-You said:
-Here is my introduction to my Bayesian modeling study and the beginning of my methods section. Suggest a continuation to the prior elicitation section
+We used satellite and in situ data from the NOMAD v2 (2008) dataset to train and validate chlorophyll-a prediction models based on remote sensing reflectance (Rrs).
+Detailed preprocessing and matching procedures were applied to ensure high-quality paired observations suitable for statistical modeling.
 
-### Historical Context of Chlorophyll Algorithms
-Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity.
-Early empirical algorithms, notably developed by O’Reilly et al.
-[@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models.
-These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial.
-These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters.
-However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.
+<!-- To expand: summarize geographic scope if applicable, temporal resolution, types of chlorophyll measurements, sensor bands used -->
 
-Subsequent refinements were introduced to address these deficiencies.
-For example, Hu et al.
-[@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving].
-The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations.
-O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, with the aim of improving the dynamic range at low chlorophyll concentrations.
-In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor.
-With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.
+- NOMAD v2 dataset: in situ measurements and coincident satellite Rrs
+- Removal of invalid or zero Rrs values
+- Computation of Max Band Ratio (MBR)
+- log-transformation of chlorophyll and MBR
+- Creation of group labels based on spectral dominance
+- Measurement method labels (fluorescence vs. HPLC)
 
-### Limitations of Existing Approaches
-Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error—one that pervades much of observational science: the conflation of sampling probability with inferential probability [@jaynes2003probability; @DeScheemaekere2011].
+### 2.2 Bayesian Modeling Framework
 
-Consider a dataset $D$ composed of input–output pairs—e.g., remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$), and a model $M$, such as OCx, posited to represent the relationship between them.
-The sampling probability $p(D|M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true.
-In standard model fitting, this likelihood is maximized by adjusting the parameters of $M$ to best explain the observed data.
+We adopt a Bayesian approach to flexibly capture the relationship between log-transformed MBR and log-transformed chlorophyll-a.
+Bayesian inference allows us to quantify uncertainty, incorporate prior knowledge, and rigorously compare alternative model structures using probabilistic criteria.
 
-This approach tacitly assumes that the model which best fits the data also most accurately represents the underlying generative process.
-This constitutes an epistemic fallacy—treating $p(D|M)$ as if it were $p(M | D)$; a violation of Bayes’ theorem and the rules governing conditional probability.
+<!-- To expand: cite PyMC, mention benefits of full posterior inference, prior predictive checking, model flexibility -->
 
-Although in well-behaved, data-rich cases—where the likelihood is regular, the signal strong, and the model adequately constrained—the maxima of $p(D|M)$ and $p(M|D)$ may coincide, this remains the exception rather than the rule.
+- Inference implemented using PyMC v5
+- Models specified using a Truncated Normal likelihood
+- Noise term (σ) modeled with increasing complexity
+- Priors: weakly informative Normal for regression terms; Gamma for σ
 
-This mistake lies at the heart of what Clayton [@clayton2022bernoulli] terms the Bernoulli Fallacy: the misconception that empirical probabilities by themselves yield inferential truths.
-As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.
+### 2.3 Overview of Candidate Models
 
-In scientific modeling, this fallacy contributes to poor generalization, drives the use of ad hoc or retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate [@baker2016; @cobey2024biomedical].
-These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.
+We developed seven Bayesian models of increasing complexity.
+These models vary in terms of functional form (polynomial vs.
+linear), structure (flat vs.
+hierarchical), error formulation (constant vs.
+heteroscedastic), and treatment of measurement-specific noise.
 
-In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.
+For brevity and clarity, we present four representative models in the main text:
+- **Model 1**: Polynomial regression (Bayesian OC6)
+- **Model 2**: Hierarchical linear regression grouped by Rrs band
+- **Model 6**: Hierarchical linear regression with group-wise heteroscedasticity
+- **Model 7**: Model 6 + additional noise term for fluorescence-based Chl
 
-This epistemic shortcut has been directly critiqued in the machine learning literature.
-Gal [@gal2016uncertainty] and Ghahramani [@ghahramani2015probabilistic] point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference.
-The result is overconfident predictions and brittle generalization—concerns that echo Clayton’s critique.
+A summary table of all models, including the three omitted from the main text, is provided below.
+Full mathematical specifications, trace diagnostics, and model performance comparisons are included in the Supplement.
 
-Bishop [@bishop2006pattern] similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient, and that the Bernoulli Fallacy permeates much of applied machine learning.
+<!-- To expand: add table summarizing models (model ID, functional form, grouping, σ structure, measurement adjustment) -->
 
-### Overcoming limitations
-There have been attempts to address these issues.
-[@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
-Others have tried to go a step futher incorporate Bayesian concepts.
-E.g.
-[ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction.
-[@shi2015] proposed a probabilistic method to merge data from different sensors.
-[@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
-[@werther2022] used Monte-Carlo dropout to approximate a deep BNN.
-[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty.
-[@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
-Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].
+### 2.4 Representative Model Structures
 
-### The Bayesian workflow
-The Bayesian workflow is a recognizable and unifying approach to statistical modeling, primarily championed by statisticians at Columnbia University [@gelman_bda] subsequently endorsed and further developed by researchers dealing with scant and/or noisy data (e.g.
-[@mcelreath_stat_rethink]).
-Some researchers have modified this  researcher's field.
-[@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for ecological modeling.
-However there are some core steps that should be followed.
-(1) Begin with one, or better yet,  more than one conceptual models and code them up.
-Model building include prior formulation to encode existing knowledge.
-Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement.
-(2) Check assumptions by a simulation process known as *Prior Predictive Checks*.
-Bayesian models are generative, meaning they can run on empty, that is produce results without data.
-This allows a sanity check of the built-in assumptions.
-Non-sensical simulation results indicate assumptions must be revisited.
-(3) Collect data and conduct exploratory data analysis.
-Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features.
-(4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information.
-Most modern packages (e.g.
-Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights.
-(5) Model compasison and selection.
-Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights.
-(6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.
+#### 2.4.1 Model 1: Bayesian Polynomial OC6 Regression
 
-In this paper we leverage past work and recast $OC6$ [@oreilly2019] as a Bayesian mdoel, because out of the $OCx$ family,  it integrates the most bands (i.e.
-information) into its structure.
-In addition, we also propose two more models.
-(1) Bayesian Additive Regression Trees (BART, [@chipman2010bart]); a robust flexible approach successfully used for diverse applications ([@linero2017].), and insensitive to the known multicollinearity issue between ocean color bands.
-(2) A hierarchical partial pooling linear model inspired from [@gordon1983]'s original band ratio-based linear regression, formulated to circumvent discontinuities mentioned by [@oreilly1998], while insuring maximal use of available information contained in the data set.
+This model mirrors the classical OC6 algorithm with a polynomial regression on log(MBR). It provides a simple, non-hierarchical baseline for model comparisons.
 
-## Methods
+- Polynomial structure similar to O'Reilly et al. (2019)
+- Constant σ
+- No group structure or measurement adjustment
 
-We performed the following five steps.
-(1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation, comparison, and selection.
+#### 2.4.2 Model 2: Hierarchical Linear Regression (4-band Grouping)
 
-### Prior Elicitation 
-Before fitting the Bayesian model to observational data, we performed a prior predictive evaluation based on simulated mock inputs to ensure that our model structure and prior choices were physically reasonable and aligned with established ocean color variability.
-For all 3 model types we simulated chlorophyll a with approximately log-normal distribution according to [@campbell1995].
-For $OC6$ and our hierarchical linear model we generated maximum band ratio (MBR) that roughly agree with [@oreilly2019] and references therein.
-BART prior is bases on model structured and elicitation is explained further below.
+This model introduces partial pooling by grouping observations according to the dominant Rrs band used in the MBR numerator.
+It aims to address known discontinuities in empirical band-ratio algorithms.
 
-#### Mock MBR and Chlorophyll Data
+- Partial pooling on slope and intercept terms
+- Shared hyperpriors across groups
+- Constant σ
 
-We generated synthetic MBR inputs based on empirical knowledge of global ocean conditions.
-Specifically, the dynamic range of MBR values was assumed to span approximately 0.3 to 30, corresponding to $log_{10}(MBR)$ values between -0.5 and 1.5 [@oreilly2019].
-$log_{10}(MBR)$ values were drawn from a truncated normal distribution centered at 0.75 with a standard deviation of 0.5, ensuring denser sampling in typical oligotrophic to mesotrophic regimes while allowing coverage of extreme cases.
-The corresponding predicted $log_{10}(Chl_a)$ ⁡ values were sampled, according to the physical (detection limits) and ecological expectations of chlorophyll variability [@campbell1995] from a truncated Gaussian likelihood bounded between -3 and 3.3, corresponding to chlorophyll concentrations ranging from 0.001 to 2000 mg m⁻³.
+#### 2.4.3 Model 6: Hierarchical Linear Regression with Group-wise Linear σ
 
-#### Model Formulation
-To implement a robust Bayesian approach, we first conducted a thorough elicitation of priors informed by domain knowledge from previously published literature, where applicable.
-The choice of distribution for the model parameters is borne out of the following assumptions.
-1. Parameter domain is $(- ∞, ∞)$ though, per [@oreilly2019], parameters are expected to be relatively close to 0
-2. Priors should be regularizing. 
-3. Parameter variance is finite.
+This model extends Model 2 by allowing the noise term σ to vary by group as a log-linear function of MBR, capturing heteroscedasticity commonly observed in chlorophyll datasets.
 
-Accordingly for the $OC6$ and hierarchical models we chose weakly informative Gaussian priors.
-For the likelihood, we chose a truncated Gaussian defined over a domain The standard deviations for these priors were chosen conservatively large enough to allow for flexibility, but sufficiently tight to capture known realistic parameter ranges derived from extensive empirical validation.
+- Group-specific linear model for log(σ)
+- Partial pooling of σ slope/intercept parameters
 
-For the BART model, priors were selected following recommendations by [@chipman2010bart] and subsequent applied literature [@linero2017; @hill2020bayesian].
-The BART algorithm naturally encodes regularization and shrinkage through its tree structure, inherently balancing bias and variance.
-However, hyperparameters controlling tree depth and the number of trees were critical.
-We specified prior distributions for these hyperparameters based on guidelines from simulation studies and domain knowledge, starting with defaults recommended by the original authors and subsequently refining through preliminary runs and prior predictive checks.
+#### 2.4.4 Model 7: Hierarchical Linear σ + Measurement-Method-Specific Noise
 
-We carried out formal prior predictive checks to validate these choices.
-By generating synthetic data from the prior distributions and inspecting the resulting predictions for chlorophyll-a, we ensured that our priors neither produced unrealistic extremes nor overly constrained the model predictions.
-This procedure also allowed stakeholders with domain expertise to visually assess the suitability of our priors before fitting the model to actual data, thereby promoting transparency and interpretability.
+Building on Model 6, this model introduces an additional noise term for chlorophyll values derived from fluorescence, accounting for the increased uncertainty relative to HPLC-based measurements.
 
-Finally, the resulting priors and model formulation and prior predictive cheks are available as a jupyter notebook in the accompanying repository for full reproducibility.
+- Same structure as Model 6
+- Additional variance term for fluorescence-flagged observations
 
-### Data Preprocessing
+### 2.5 Prior Specification and Inference Settings
 
-\section{Data Preprocessing}
+All models use weakly informative priors and were fit using Hamiltonian Monte Carlo (NUTS) in PyMC. Convergence and effective sampling were confirmed via standard diagnostics.
 
-Data for this study were acquired from multiple satellite ocean color sensors and corresponding in situ chlorophyll-\(a\) measurements obtained from sources such as the NASA Bio-Optical Marine Algorithm Data set (NOMAD) and the compilation by Valente et al.
-(2015).
-To ensure consistency across sensors, the spectral reflectance data (\(R_{rs}\)) were interpolated as needed to common wavelength centers.
+<!-- To expand: mention specific prior distributions used, tuning parameters, number of chains/draws -->
 
-For the empirical \(OCx\) formulation, blue-to-green band ratios were computed for each observation.
-In particular, the maximum band ratio (MBR) was determined by taking the highest value among the available blue-band ratios (e.g., \(Rrs(443)/Rrs(555)\), \(Rrs(490)/Rrs(555)\), and \(Rrs(510)/Rrs(555)\)).
-This maximum value was then log-transformed:
-$$
-\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
-$$
+- Priors: Normal for regression terms, Gamma for σ
+- Chains, tuning steps, and draws per model
+- Convergence assessed using R-hat and ESS
+- Trace plots and diagnostics in Supplementary Material
 
-For the Color Index (CI) formulation of Hu et al. (2012), the CI was calculated as:
-$$
-\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
-$$
-and the corresponding in situ chlorophyll-\(a\) concentrations were log-transformed:
-$$
-\log \text{Chl} = \log_{10}(\text{Chl}).
-$$
+### 2.6 Model Comparison and Predictive Evaluation
 
-These transformations standardize the data to a common scale, ensuring that variability is appropriately captured for subsequent regression and uncertainty quantification.
-Detailed descriptions of the interpolation methods and quality control procedures are provided in the Supplementary Material.
+Models were compared using Leave-One-Out Cross Validation (LOO-CV) based on Pareto Smoothed Importance Sampling (PSIS), a robust criterion for evaluating out-of-sample predictive performance.
 
-## Statement of Contribution
-In this study, we develop and demonstrate a new global chlorophyll retrieval model based on Bayesian Additive Regression Trees implemented in PyMC.
-We train the BART model on a large, standardized dataset of satellite remote-sensing reflectance (Rrs) spectra matched with in situ chlorophyll measurements, using log₁₀-transformed Chl-a as the response to stabilize variance.
-The resulting model is applied globally to produce chlorophyll-a estimates from multi-spectral satellite data, with associated uncertainty estimates for each prediction.
-We show that this Bayesian tree-based model can serve as a general-purpose ocean color algorithm that is sensor-agnostic (provided reflectances are harmonized to common wavebands), interpretable, and uncertainty-aware.
-Unlike conventional empirical algorithms, the BART approach allows users to examine the inferred Rrs–Chl relationships and trust the model’s performance across regimes, while also quantifying confidence in each retrieval.
-This work thus contributes a novel methodological advance to satellite ocean color science: a unified chlorophyll retrieval model that marries the strengths of empirical algorithms (global applicability and simplicity) with the benefits of modern Bayesian machine learning (flexibility, interpretability, and rigorous uncertainty quantification).
-Our introduction of BART for global chlorophyll prediction opens the door for more robust monitoring of ocean biogeochemistry and improved integration of ocean color data into scientific and management applications.
+<!-- To expand: cite ArviZ, explain ELPD, SE(ΔELPD), ranking strategy -->
+
+- Model comparison was performed using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)**, implemented via the `arviz.loo()` function.
+- Models were ranked based on their **Expected Log Predictive Density (ELPD)**, a Bayesian measure of out-of-sample predictive accuracy.
+- Differences in ELPD between models were evaluated alongside their **standard errors (SE)** to assess whether improvements were credibly distinguishable.
+- **Predictive accuracy** was assessed using:
+  - **Posterior Predictive Checks (PPCs)**: to visualize the degree to which simulated data resembled observed outcomes.
+  - **Highest Density Interval (HDI) coverage**: to evaluate the calibration of predictive intervals relative to held-out data.
+- No frequentist criteria (e.g., RMSE, R²) were used at any stage of model evaluation or selection.
+
+## 3. Results
+
+This section presents convergence diagnostics, posterior inference, and model comparison using fully Bayesian metrics.
+No frequentist evaluation criteria were used.
+Instead, we rely on information-theoretic model comparison (LOO-CV), posterior predictive checks, and visualization of uncertainty via highest density intervals (HDIs).
+
+### 3.1 MCMC Convergence
+
+All four models converged successfully.
+The Gelman-Rubin statistic (R-hat ≈ 1.0) and large effective sample sizes (ESS > 1000) across all parameters indicate robust posterior estimation.
+Diagnostic plots and sampling traces are provided in the Supplementary Material.
+
+- All R-hat < 1.01
+- No divergences
+- Sufficient tail and bulk ESS
+- Full diagnostics in Supplement
+
+### 3.2 Predictive Performance via Leave-One-Out Cross Validation
+
+We compared models using Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV), a fully Bayesian approach to evaluating out-of-sample predictive accuracy.
+
+**Table 1.** Leave-One-Out Comparison (ELPD ± SE):
+
+<!-- Insert Quarto table of ELPD, ΔELPD, SE, p_loo -->
+
+- Model 7 ranked highest in expected log predictive density (ELPD)
+- Model 6 performed comparably, with ΔELPD within 1 SE
+- Model 1 (OC6 polynomial) had the weakest predictive performance
+- All models had acceptable Pareto-k diagnostics (see Supplement)
+
+### 3.3 Posterior Predictive Checks (PPC)
+
+We used posterior predictive checks (PPCs) to evaluate the ability of each model to reproduce key features of the observed log-chlorophyll distribution. These include:
+
+- Empirical CDF comparisons
+- HDI envelopes overlaid on the data
+- Predictive median vs. observed values
+
+**Figure 1.** Posterior predictive envelopes vs. observed log(Chl)  
+**Figure 2.** ECDF overlays with posterior predictive means and HDIs  
+**Figure 3.** Rug plot of observations with overlaid posterior predictive draws
+
+Findings:
+- Model 7 captures full data distribution, especially for fluorescence-labeled values
+- Model 1 underestimates tails (under-dispersed)
+- Model 6 accurately reflects group-level variance
+
+### 3.4 Posterior Distributions and Interpretability
+
+Posterior distributions for key parameters illustrate how model structure affects inference. We show both:
+- Group-level slopes and intercepts (hierarchical structure)
+- Group-wise σ (heteroscedasticity)
+- Fluorescence-specific variance term (Model 7)
+
+**Figure 4.** Forest plot of slope and intercept posteriors  
+**Figure 5.** Posterior of σ across MBR groups  
+**Figure 6.** Posterior of added noise term for fluorescence
+
+Key insights:
+- Rrs510-dominant group had the steepest slope (Model 2 & 6)
+- Fluorescence noise term is credibly nonzero (Model 7)
+- Posterior variance structure supports heteroscedasticity as a key feature
+
+### 3.5 Summary of Bayesian Comparison
+
+- **Model 7**: Best predictive performance (highest ELPD), best uncertainty calibration
+- **Model 6**: Nearly equivalent performance; simpler if fluorescence uncertainty not needed
+- **Model 2**: Gains from partial pooling; underperforms without variance modeling
+- **Model 1**: Legacy structure underfits tails, overconfident predictions
+
+All results support the Bayesian hierarchical modeling framework with heteroscedasticity and measurement-aware noise as optimal for chlorophyll-a estimation.
+
+## 4. Discussion
+
+This section interprets the results in the context of previous chlorophyll-a modeling work, with particular emphasis on the added value of Bayesian modeling.
+We highlight the importance of accounting for heteroscedasticity and measurement-type uncertainty, and discuss how this contributes to improved predictive reliability and scientific interpretability.
+
+### 4.1 Value of Bayesian Framework over Classical Approaches
+
+Our results demonstrate the clear advantages of fully Bayesian modeling for chlorophyll-a prediction from satellite Rrs. Unlike deterministic or frequentist models:
+
+- Bayesian methods explicitly model parameter uncertainty and provide HDIs instead of misleading confidence intervals
+- Posterior predictive distributions enable direct evaluation of model calibration
+- Bayesian model comparison via LOO-CV avoids overfitting and guards against inappropriate complexity
+
+<!-- Expand: reinforce critique of RMSE-based selection; cite key Bayesian modeling references -->
+
+### 4.2 Importance of Hierarchical Structuring
+
+Models that leveraged group-wise structure (based on dominant Rrs bands) consistently outperformed simpler counterparts.
+
+- Hierarchical partial pooling improved both prediction and parameter identifiability
+- Group-level trends reflected known biogeophysical distinctions (e.g., clear vs. turbid water types)
+- Intercepts and slopes varied systematically across spectral regimes
+
+<!-- Expand: relate group-level results to known Case 1 vs. Case 2 water differences -->
+
+### 4.3 Role of Heteroscedasticity in Chlorophyll Retrieval
+
+Accounting for non-constant noise was essential for accurate uncertainty quantification.
+
+- Models with linear or hierarchical σ terms captured increased variance at higher MBR or Chl
+- Constant-σ models (e.g., OC6) underrepresented predictive uncertainty, especially in extremes
+- Explicit modeling of σ structure produced calibrated HDIs and superior posterior predictive checks
+
+<!-- Expand: link back to observed data behavior and prior studies showing variance scaling with concentration -->
+
+### 4.4 Measurement-Specific Error Modeling
+
+Model 7, which introduced a separate noise term for fluorescence-derived Chl, outperformed all others.
+
+- Fluorescence-based measurements have higher and more variable error
+- This heterogeneity cannot be ignored in merged datasets
+- Bayesian modeling allows this variance to be encoded explicitly without discarding data
+
+<!-- Expand: cite studies on fluorescence bias, show that Bayesian methods allow uncertainty to “absorb” method bias -->
+
+### 4.5 Implications for Satellite Algorithm Development
+
+The findings support incorporating Bayesian elements into future operational chlorophyll algorithms.
+
+- Hierarchical modeling offers a path to generalize across bioregions and sensors
+- Posterior uncertainty estimates can support downstream applications (e.g., forecasting, ecological thresholds)
+- Bayesian frameworks are sensor-agnostic: Rrs inputs can vary as long as priors are adapted
+
+<!-- Optional: comment on feasibility of deploying these methods, or emulating them with approximate Bayesian techniques -->
+
+### 4.6 Limitations and Future Work
+
+As with any modeling effort, our study has limitations:
+
+- All results are conditional on the data from NOMAD; further validation on independent datasets is needed
+- More sophisticated noise structures (e.g., nonparametric σ) could be explored
+- Spatial or spatio-temporal extensions were not considered, but would be feasible in PyMC
+
+<!-- Expand: suggest extensions (Gaussian Processes, spatial priors, stacking), and invite reproducibility -->
 

### 2025-05-17 manuscript/manuscript.qmd
updated introduction

@@
 
-There have been attempts to address these issues.
-[@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach.
-Others have tried to go a step futher incorporate Bayesian concepts.
-E.g.
-[ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction.
-[@shi2015] proposed a probabilistic method to merge data from different sensors.
-[@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance.
-[@werther2022] used Monte-Carlo dropout to approximate a deep BNN.
-[@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty.
-[@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations.
-Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].
+Several recent efforts have attempted to address the limitations of classical retrieval models.
+For instance, [@seegers2018] proposed alternative evaluation metrics to move beyond restrictive frequentist assumptions.
+Others have incorporated Bayesian elements into the modeling pipeline: [@frouin2013] applied Bayesian inversion for atmospheric correction, [@shi2015] used probabilistic fusion for multi-sensor data, and [@craig2019] employed Hamiltonian Monte Carlo to train Bayesian neural networks (BNNs) for retrieving inherent optical properties (IOPs) from top-of-atmosphere radiance.
+Similarly, [@werther2022] introduced Monte Carlo dropout as an approximate BNN strategy, while [@erickson2023] recast the Generalized Inherent Optical Property (GIOP) framework using conjugate Bayesian linear models.
+Most recently, [@hammout2024] developed a BNN surrogate using stochastic variational inference for chlorophyll-a prediction from Rrs.
 
-### In search of our Bayesian workflow
-The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g.
-[@mcelreath_stat_rethink]).
-That is not to say that all steps are applicable to all fields of research and all intents.
-The workflow can and should be tailored to the researcher's field.
-[@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for Ecological modeling.
-However there are some core steps that should be followed.
-(1) Begin with one, or better yet,  more than one conceptual models and code them up.
-Model building include prior formulation to encode existing knowledge.
-Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement.
-(2) Check assumptions by a simulation process known as *Prior Predictive Checks*.
-Bayesian models are generative, meaning they can run on empty, that is produce results without data.
-This allows a sanity check of the built-in assumptions.
-Non-sensical simulation results indicate assumptions must be revisited.
-(3) Collect data and conduct exploratory data analysis.
-Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features.
-(4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information.
-Most modern packages (e.g.
-Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights.
-(5) Model compasison and selection.
-Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights.
-(6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.
+While these studies mark important progress, they often adopt only isolated components of what has come to be known as the Bayesian workflow [@bgelman2019].
+Key aspects—such as principled prior specification, posterior predictive checking, and formal model comparison—remain largely absent.
+As a result, uncertainty is frequently approximated rather than inferred, and model structure is rarely treated as a variable to be interrogated.
 
-In this paper we leverage past work but also propose new approaches for predicting chlorophyll from sea surface reflectance (Rrs). Overall we propose three types of models.
+This paper builds on prior work and extends it by applying a fully Bayesian modeling framework for chlorophyll-a prediction from sea surface reflectance (Rrs), with explicit treatment of uncertainty, model complexity, and measurement error.
 
@@
 
-This section outlines the study region and dataset used for model development and validation, the structure and rationale for the Bayesian models, and the approach used to compare their predictive performance.
-We emphasize a subset of four representative models in the main text, while full specifications and results for all seven models are provided in the Supplementary Material.
+This section outlines the dataset, model structures, inference methods, and model evaluation criteria used in this study.
+I focus on a fully Bayesian approach to modeling chlorophyll-a from satellite remote sensing reflectance (Rrs), emphasizing transparency, uncertainty quantification, and principled model comparison.
+Four representative models are presented in the main text, while all seven model formulations and results are included in the Supplement.
 
@@
 
-We used satellite and in situ data from the NOMAD v2 (2008) dataset to train and validate chlorophyll-a prediction models based on remote sensing reflectance (Rrs).
-Detailed preprocessing and matching procedures were applied to ensure high-quality paired observations suitable for statistical modeling.
+I used the NOMAD v2 (2008) dataset, which includes quality-controlled in situ measurements of chlorophyll-a matched with coincident satellite Rrs data.
+These measurements span a range of oceanographic conditions, enabling the development of models that generalize across water types.
 
-<!-- To expand: summarize geographic scope if applicable, temporal resolution, types of chlorophyll measurements, sensor bands used -->
+To prepare the data:
 
-- NOMAD v2 dataset: in situ measurements and coincident satellite Rrs
-- Removal of invalid or zero Rrs values
-- Computation of Max Band Ratio (MBR)
-- log-transformation of chlorophyll and MBR
-- Creation of group labels based on spectral dominance
-- Measurement method labels (fluorescence vs. HPLC)
+- **Chlorophyll-a**:  
+  - Measurements obtained using both HPLC and fluorescence methods  
+  - Combined into a single log-transformed target variable (log₁₀ Chl)  
+  - A binary flag identifies the measurement method for each sample
+
+- **Remote sensing reflectance (Rrs)**:  
+  - Rrs values equal to zero were excluded  
+  - A derived feature, the Maximum Band Ratio (MBR), was computed as:  
+    - Numerator: max(Rrs411, Rrs443, Rrs489, Rrs510)  
+    - Denominator: Rrs555 + Rrs670  
+  - MBR was log-transformed to create the primary input variable
+
+- **Grouping and flags**:  
+  - A categorical variable (`MBR_flag`) was used to indicate the dominant Rrs band used in the MBR numerator  
+  - This grouping informed partial pooling structures in hierarchical models
 
@@
 
-We adopt a Bayesian approach to flexibly capture the relationship between log-transformed MBR and log-transformed chlorophyll-a.
-Bayesian inference allows us to quantify uncertainty, incorporate prior knowledge, and rigorously compare alternative model structures using probabilistic criteria.
+All models were implemented in PyMC v5 and estimated using Hamiltonian Monte Carlo with the No-U-Turn Sampler (NUTS).
+I modeled log-transformed chlorophyll-a as a function of log-transformed MBR, using a truncated normal likelihood to ensure predictions remained within physically plausible bounds.
 
-<!-- To expand: cite PyMC, mention benefits of full posterior inference, prior predictive checking, model flexibility -->
+Key components of the modeling framework include:
 
-- Inference implemented using PyMC v5
-- Models specified using a Truncated Normal likelihood
-- Noise term (σ) modeled with increasing complexity
-- Priors: weakly informative Normal for regression terms; Gamma for σ
+- **Likelihood**:  
+  - Truncated normal with support between –3.0 and 3.2 (log₁₀ Chl)
+
+- **Noise structure** (`σ`):  
+  - Varied across models, with increasing structural complexity  
+  - Some models use constant σ; others include heteroscedastic or hierarchical formulations
+
+- **Priors**:  
+  - Weakly informative priors were used throughout  
+  - Regression coefficients received Normal priors  
+  - All σ terms were modeled with Gamma distributions, allowing flexible but strictly positive support
+
+- **Software**:  
+  - PyMC for model specification and inference  
+  - ArviZ for posterior diagnostics, posterior predictive checks, and model comparison
 
@@
 
-We developed seven Bayesian models of increasing complexity.
-These models vary in terms of functional form (polynomial vs.
-linear), structure (flat vs.
-hierarchical), error formulation (constant vs.
-heteroscedastic), and treatment of measurement-specific noise.
+I developed seven Bayesian models representing a progression of modeling complexity and assumptions:
 
-For brevity and clarity, we present four representative models in the main text:
-- **Model 1**: Polynomial regression (Bayesian OC6)
-- **Model 2**: Hierarchical linear regression grouped by Rrs band
-- **Model 6**: Hierarchical linear regression with group-wise heteroscedasticity
-- **Model 7**: Model 6 + additional noise term for fluorescence-based Chl
+- **Model 1**: Polynomial regression (Bayesian OC6-style baseline)  
+- **Model 2**: Hierarchical linear regression, grouped by dominant MBR numerator band  
+- **Model 3**: Hierarchical linear regression extended to include all Rrs spectral dominance groups  
+- **Model 4**: Model 2 with group-specific constant σ  
+- **Model 5**: Global linear heteroscedastic model  
+- **Model 6**: Model 2 with group-wise linear heteroscedasticity  
+- **Model 7**: Model 6 plus an added dispersion term for fluorescence-based measurements
 
-A summary table of all models, including the three omitted from the main text, is provided below.
-Full mathematical specifications, trace diagnostics, and model performance comparisons are included in the Supplement.
-
-<!-- To expand: add table summarizing models (model ID, functional form, grouping, σ structure, measurement adjustment) -->
+The main text focuses on Models 1, 2, 6, and 7, which span the range of key structural variations. Full equations, priors, and diagnostic results for all models are provided in the Supplement.
 
@@
 
-This model mirrors the classical OC6 algorithm with a polynomial regression on log(MBR). It provides a simple, non-hierarchical baseline for model comparisons.
+This model follows the OC6 formulation proposed by O’Reilly et al. (2019), adapted to a Bayesian framework. It uses a fourth-order polynomial in log(MBR) to estimate log(Chl).
 
-- Polynomial structure similar to O'Reilly et al. (2019)
-- Constant σ
-- No group structure or measurement adjustment
+- No grouping or partial pooling  
+- Constant noise term σ shared across all observations  
+- Gamma prior placed on σ
 
@@
 
-This model introduces partial pooling by grouping observations according to the dominant Rrs band used in the MBR numerator.
-It aims to address known discontinuities in empirical band-ratio algorithms.
+This model introduces partial pooling across four groups defined by the dominant MBR numerator band. Each group receives its own slope and intercept, with shared hyperpriors.
 
-- Partial pooling on slope and intercept terms
-- Shared hyperpriors across groups
-- Constant σ
+- Grouping: Rrs411, Rrs443, Rrs489, Rrs510  
+- Partial pooling via shared hyperpriors  
+- Constant σ across all groups
 
@@
 
-This model extends Model 2 by allowing the noise term σ to vary by group as a log-linear function of MBR, capturing heteroscedasticity commonly observed in chlorophyll datasets.
+To accommodate heteroscedasticity, this model allows σ to vary with log(MBR) within each group, modeled as a log-linear relationship. Parameters are estimated hierarchically across groups.
 
-- Group-specific linear model for log(σ)
-- Partial pooling of σ slope/intercept parameters
+- log(σ) = α + β × log(MBR), with group-specific α and β  
+- Transformed via exp(log(σ))  
+- Partial pooling of σ slope/intercept terms
 
-#### 2.4.4 Model 7: Hierarchical Linear σ + Measurement-Method-Specific Noise
+#### 2.4.4 Model 7: Hierarchical Linear σ + Measurement-Specific Dispersion
 
-Building on Model 6, this model introduces an additional noise term for chlorophyll values derived from fluorescence, accounting for the increased uncertainty relative to HPLC-based measurements.
+This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates.
 
-- Same structure as Model 6
-- Additional variance term for fluorescence-flagged observations
+- Additional noise component applied when chl_flag = fluorescence  
+- All other structure inherited from Model 6
 
@@
 
-All models use weakly informative priors and were fit using Hamiltonian Monte Carlo (NUTS) in PyMC. Convergence and effective sampling were confirmed via standard diagnostics.
+I used weakly informative priors across all models to balance flexibility and regularization:
 
-<!-- To expand: mention specific prior distributions used, tuning parameters, number of chains/draws -->
+- **Regression terms**: Normal(0, 1) unless otherwise stated  
+- **Hyperpriors**: Normal and HalfNormal for hierarchical parameters  
+- **Noise terms** (`σ`): Gamma(γ, φ), chosen to reflect plausible dispersion values
 
-- Priors: Normal for regression terms, Gamma for σ
-- Chains, tuning steps, and draws per model
-- Convergence assessed using R-hat and ESS
-- Trace plots and diagnostics in Supplementary Material
+Inference was conducted using:
+
+- 4 chains per model  
+- 1000–1500 tuning steps  
+- ≥2000 posterior samples per chain
+
+I assessed convergence using R-hat (target < 1.01) and effective sample size (ESS > 1000). Full trace plots and diagnostics are included in the Supplement.
 
@@
 
-Models were compared using Leave-One-Out Cross Validation (LOO-CV) based on Pareto Smoothed Importance Sampling (PSIS), a robust criterion for evaluating out-of-sample predictive performance.
+I compared model performance using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)**, implemented via the `arviz.loo()` function.
+This Bayesian metric estimates the expected log predictive density (ELPD), a principled measure of out-of-sample predictive accuracy.
 
-<!-- To expand: cite ArviZ, explain ELPD, SE(ΔELPD), ranking strategy -->
+- Models were ranked by ELPD  
+- ΔELPD values were interpreted relative to their standard errors  
+- Pareto-k diagnostics confirmed LOO reliability (full table in Supplement)
 
-- Model comparison was performed using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)**, implemented via the `arviz.loo()` function.
-- Models were ranked based on their **Expected Log Predictive Density (ELPD)**, a Bayesian measure of out-of-sample predictive accuracy.
-- Differences in ELPD between models were evaluated alongside their **standard errors (SE)** to assess whether improvements were credibly distinguishable.
-- **Predictive accuracy** was assessed using:
-  - **Posterior Predictive Checks (PPCs)**: to visualize the degree to which simulated data resembled observed outcomes.
-  - **Highest Density Interval (HDI) coverage**: to evaluate the calibration of predictive intervals relative to held-out data.
-- No frequentist criteria (e.g., RMSE, R²) were used at any stage of model evaluation or selection.
+I evaluated predictive accuracy and calibration using:
+
+- **Posterior Predictive Checks (PPCs)**:  
+  - Simulated draws overlaid on observed distributions  
+  - Empirical CDF comparisons and interval coverage
+
+- **HDI coverage**:  
+  - Fraction of observed values falling within 95% posterior predictive intervals
+
+No frequentist evaluation criteria such as RMSE, MAE, or R² were used at any stage as these do not make sense in this context where for each observations, a probability distribution of the prediction is produced.
 

### 2025-05-18 manuscript/manuscript.qmd
updated methods section  intro

@@
 keep-tex: true
-number-sections: false
+number-sections: true
 date: last-modified
@@
 
-I used the NOMAD v2 (2008) dataset, which includes quality-controlled in situ measurements of chlorophyll-a matched with coincident satellite Rrs data.
+For pedagogical reasons, I used the well known NOMAD [@werdell2005improved] dataset, which includes quality-controlled in situ measurements of chlorophyll-a matched with coincident satellite Rrs data.
 These measurements span a range of oceanographic conditions, enabling the development of models that generalize across water types.
+Sea-surface reflectances (Rrs) in the 6 visible SeaWiFS bands (411, 443, 489, 510, 550, 670nm) constitue the bulk of the data.
+As prediction targets corresponding chlorophyll concentrations using both fluorescence and HPLC were retained.
 
@@
 
-This section presents convergence diagnostics, posterior inference, and model comparison using fully Bayesian metrics.
-No frequentist evaluation criteria were used.
-Instead, we rely on information-theoretic model comparison (LOO-CV), posterior predictive checks, and visualization of uncertainty via highest density intervals (HDIs).
+This section presents the outputs of posterior inference, model comparison using Bayesian criteria, and posterior predictive diagnostics for the four primary models.
+Results from the three supplementary models are provided separately in the Supplement.
 

### 2025-05-19 manuscript/manuscript.qmd
finished methods/data preprocessing section

@@
 
-### 2.1 Study Area and Dataset
+### 2.1 Data Description and Preparation Overview
 
 For pedagogical reasons, I used the well known NOMAD [@werdell2005improved] dataset, which includes quality-controlled in situ measurements of chlorophyll-a matched with coincident satellite Rrs data.
-These measurements span a range of oceanographic conditions, enabling the development of models that generalize across water types.
-Sea-surface reflectances (Rrs) in the 6 visible SeaWiFS bands (411, 443, 489, 510, 550, 670nm) constitue the bulk of the data.
-As prediction targets corresponding chlorophyll concentrations using both fluorescence and HPLC were retained.
+These measurements span a range of oceanographic conditions, enabling the development of models that are expected to generalize across water types.
+The data were loaded using the Python Pandas library [@reback2020pandas], which was used for all subsequent cleaning, preprocessing, and transformation operations.
 
-To prepare the data:
+*Raw Data* - Sea-surface reflectances (Rrs) in the 6 visible SeaWiFS bands (411, 443, 489, 510, 550, 670nm) constitue the bulk of the data.
+Chlorophyll concentrations, measured via fluorescence or HPLC or both, consituted the target variables.
+Any row containing an invalid data point (null, zero, or negative) was removed, as proper missing/invalid data imputation is a subject onto itself and beyond the scope of this paper.
 
-- **Chlorophyll-a**:  
-  - Measurements obtained using both HPLC and fluorescence methods  
-  - Combined into a single log-transformed target variable (log₁₀ Chl)  
-  - A binary flag identifies the measurement method for each sample
+*Data Preprocessing and Transformation* - Chlorophyll measurements from the two different methods were combined into a single data.
+An auxiliary column was used to flag the measurement method as 'fluo' of 'hplc'.
+The chlorophyll was then log-transformed to stabilize its variance.
+The max band ratio (MBR) was computed as $\frac {\max(Rrs411, Rrs443, Rrs489, Rrs)}{Rrs555 + Rrs670}$, similar to the OC6 model proposed by [@oreilly2019].
+Another ancillary column was created to captured the actual Rrs band used for each set observations.
+This numerator type as well as the chlorophyll measurement type are useful to build more hierarchical partial pooling models that can take advantage of such data groups.
+A sample from the fully processed data used in the subsequent models can be seen in [@data-table].
 
-- **Remote sensing reflectance (Rrs)**:  
-  - Rrs values equal to zero were excluded  
-  - A derived feature, the Maximum Band Ratio (MBR), was computed as:  
-    - Numerator: max(Rrs411, Rrs443, Rrs489, Rrs510)  
-    - Denominator: Rrs555 + Rrs670  
-  - MBR was log-transformed to create the primary input variable
+:::
 
-- **Grouping and flags**:  
-  - A categorical variable (`MBR_flag`) was used to indicate the dominant Rrs band used in the MBR numerator  
-  - This grouping informed partial pooling structures in hierarchical models
+|   | log_MBR              | MBR_flag   | log_chl              | hplc_flag   |
+|----:|:-----------------------|:-----------|:-----------------------|:------------|
+| 905 | -0.2545615131431814  | Rrs510     |  0.5837653682849998  | hplc        |
+|  91 | -0.3641327253450254  | Rrs510     |  1.0511525224473812  | fluo        |
+| 528 | -0.0737170739918758  | Rrs510     |  0.14363923527454328 | fluo        |
+| 969 |  0.5813600497877703  | Rrs411     | -0.8272803386505035  | fluo        |
+| 552 |  0.7134341855521412  | Rrs443     | -0.9318141382538383  | hplc        |
+::: A random 5-row sample from the Pandas dataframe holding the data used in subsequent models.
+*log_MBR* is the log-transformed Max.
+Band Ratio (MBR) of sea surface reflectances, where the numerator is Max(Rrs411, Rrs443, Rrs489, Rrs510) and the denominator is computed from the sum of Rrs555 and Rrs670.
+*MBR_flag* indicates which of the 4 possible Rrs bands is used in the numerator for that particular row.
+*log_chl* is log transformed in-situ measured chlorophyll *a* concentration, originally in $mg\m ^{-3}$.
+*hplc_flag* is a binary variable indicating whether the measurement was obtained via HPLC or fluorometry.
+"-flag" are used during modeling for grouping data for hierarchical partial pooling models (see below.) {#data-table}
+:::
 

### 2025-05-19 manuscript/manuscript.qmd
updated model overview

@@
 I focus on a fully Bayesian approach to modeling chlorophyll-a from satellite remote sensing reflectance (Rrs), emphasizing transparency, uncertainty quantification, and principled model comparison.
-Four representative models are presented in the main text, while all seven model formulations and results are included in the Supplement.
+Four  models are presented in the main text, while all six model formulations and results are included in the Supplement.
 
@@
 
-All models were implemented in PyMC v5 and estimated using Hamiltonian Monte Carlo with the No-U-Turn Sampler (NUTS).
-I modeled log-transformed chlorophyll-a as a function of log-transformed MBR, using a truncated normal likelihood to ensure predictions remained within physically plausible bounds.
+A sound Bayesian modeling principle is to develop several models with increasing levels of complexity as insight is progressively built.
+I largely follow this approach with the exception that the first model is built from a known formulation, a 4th order polynomial.
+All models use $log(MBR)$ as main input and $log(chl)$ as target, some use $MBR\_flag$ as additional input, the last model also uses $hlpc_flag$.
+The following are key components shared across models.
 
-Key components of the modeling framework include:
+*Likelihood* - A truncated gaussian distribution with support constrained to -3.0 to 3.2 on the $log_{10}$ scale,corresponding to 0.001 - 1600 $mg\ m^{-3}$ to reflect both detectable realistic levels of marine surface chlorophyll.
 
-- **Likelihood**:  
-  - Truncated normal with support between –3.0 and 3.2 (log₁₀ Chl)
+*Noise Structure* - Noise was incorporated as likelihood dispersion parameter $σ$; some of the models use a constant $σ$, others take heteroscedasticity into account.
 
-- **Noise structure** (`σ`):  
-  - Varied across models, with increasing structural complexity  
-  - Some models use constant σ; others include heteroscedastic or hierarchical formulations
+*Priors* - To estimate model posterior distribution, and produce predictions with uncertainty following data fitting, all model parameters receive a prior distribution, making explicit model assumptions.
+Often there will be enough data to overrule the priors, which are here mostly a conduit to the Posterior.
+An important caveat is that care should be taken not to assign a 0-probability through a poorly specified prior where support might actually be needed.
+Less importantly, more efficient use of the search space can be done through priors that assign 0-probability where support is clearly unreasonable.
+Thus, all regression parameters receive Normal (Gaussian) 0-centered weakly informative priors.
+And all $σ$ terms received Gamma priors that provide flexible but strictly positive support.
 
-- **Priors**:  
-  - Weakly informative priors were used throughout  
-  - Regression coefficients received Normal priors  
-  - All σ terms were modeled with Gamma distributions, allowing flexible but strictly positive support
-
-- **Software**:  
-  - PyMC for model specification and inference  
-  - ArviZ for posterior diagnostics, posterior predictive checks, and model comparison
+*Software* -  All models were implemented with the Python probabilistic programming library PyMC V.5 [@abril2023pymc] and estimated using a variant of Hamiltonian Monte Carlo known as the No-U-Turn Sampler (NUTS).
+Model diagnostics inference evaluation and model comparisons were done through the Arviz package.
 
@@
 
-I developed seven Bayesian models representing a progression of modeling complexity and assumptions:
+I developed six models representing a progression of modeling complexity and assumptions:
 
@@
 - **Model 2**: Hierarchical linear regression, grouped by dominant MBR numerator band  
-- **Model 3**: Hierarchical linear regression extended to include all Rrs spectral dominance groups  
-- **Model 4**: Model 2 with group-specific constant σ  
+- **Model 4**: Model 2 with group-specific constant likelihood variance  
 - **Model 5**: Global linear heteroscedastic model  

### 2025-05-21 manuscript/manuscript.qmd
added supplementary materials and moved eqns there

@@
 
-:::
+:::{#data-table}
 
@@
 | 552 |  0.7134341855521412  | Rrs443     | -0.9318141382538383  | hplc        |
-::: A random 5-row sample from the Pandas dataframe holding the data used in subsequent models.
+
+A random 5-row sample from the Pandas dataframe holding the data used in subsequent models.
 *log_MBR* is the log-transformed Max.
@@
 *MBR_flag* indicates which of the 4 possible Rrs bands is used in the numerator for that particular row.
-*log_chl* is log transformed in-situ measured chlorophyll *a* concentration, originally in $mg\m ^{-3}$.
+*log_chl* is log transformed in-situ measured chlorophyll *a* concentration, originally in $mg\ m^{-3}$.
 *hplc_flag* is a binary variable indicating whether the measurement was obtained via HPLC or fluorometry.
-"-flag" are used during modeling for grouping data for hierarchical partial pooling models (see below.) {#data-table}
+"-flag" are used during modeling for grouping data for hierarchical partial pooling models, explained further below.
+These groupings can be seen in [@fig-eda]
+
+:::
+
+::: {#fig-eda layout-ncol=2}
+
+![](images/eda_numerator.png){#fig-eda-mbr}
+
+![](images/eda_hplc.png){#fig-eda-hplc}
+
+Two ways of grouping the data. [@fig-eda-mbr], max band ratio numerator; [@fig-eda-hplc], chlorophyll measurement method.
 :::
@@
 
-A sound Bayesian modeling principle is to develop several models with increasing levels of complexity as insight is progressively built.
-I largely follow this approach with the exception that the first model is built from a known formulation, a 4th order polynomial.
-All models use $log(MBR)$ as main input and $log(chl)$ as target, some use $MBR\_flag$ as additional input, the last model also uses $hlpc_flag$.
+Several models of increasing levels were built.
+All models use $log(MBR)$ as primary input and $log(chl)$ as target, some use $MBR\_flag$ as additional input, the last model also uses $hlpc_flag$.
 The following are key components shared across models.
 
-*Likelihood* - A truncated gaussian distribution with support constrained to -3.0 to 3.2 on the $log_{10}$ scale,corresponding to 0.001 - 1600 $mg\ m^{-3}$ to reflect both detectable realistic levels of marine surface chlorophyll.
+*Likelihood* - A truncated gaussian distribution with support constrained to -3.0 to 3.2 on the $log_{10}$ scale,corresponding to $0.001 - 1600 mg\ m^{-3}$ to reflect both detectable realistic levels of marine surface chlorophyll.
 
@@
 
-- No grouping or partial pooling  
-- Constant noise term σ shared across all observations  
-- Gamma prior placed on σ
+This model can be represented by the following set of probabilistic equations:
+
+\begin{align*}
+    \alpha &\sim \text{Normal}(0, 1) \\
+    \beta_1 &\sim \text{Normal}(0, 1) \\
+    \beta_2 &\sim \text{Normal}(0, 1) \\
+    \beta_3 &\sim \text{Normal}(0, 1) \\
+    \beta_4 &\sim \text{Normal}(0, 1) \\
+    \mu &= \alpha + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 \\
+    \sigma &\sim \text{Gamma}(2, 2) \\
+    \text{likelihood} &\sim \text{TruncatedNormal}(\mu, \sigma, -3, 3.3)
+\end{align*}
+
+::: {#fig-oc6-struct}
+
+![](images/oc6_structure.svg)
+
+:::
 

### 2025-05-21 manuscript/supplement.qmd
added supplementary materials and moved eqns there

---
title: "Supplementary Material"
subtitle: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation"
format:
  pdf:
    toc: true
    number-sections: true
    keep-tex: true
bibliography: references.bib
keep-tex: true
---

This document provides supplementary material supporting the manuscript titled _"Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation"_, submitted to *Remote Sensing of Environment*.

# Supplementary Methods

- Full Bayesian model specifications
- Prior predictive checks
- Diagnostic metrics for convergence (e.g., $\hat{R}$, ESS)
- Model comparison procedures (WAIC, LOO)

## Model 1 - Fourth Order Polynomial Regression
- model introduction (relation to Oreilly and Werdell 2019)

### Model Description and Priors
- brief word description

### Mathematical Description
\begin{align*}
    \alpha &\sim \text{Normal}(0, 1) \\
    \beta_1 &\sim \text{Normal}(0, 1) \\
    \beta_2 &\sim \text{Normal}(0, 1) \\
    \beta_3 &\sim \text{Normal}(0, 1) \\
    \beta_4 &\sim \text{Normal}(0, 1) \\
    \mu_i &= \alpha + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 \\
    \sigma &\sim \text{Gamma}(2, 2) \\
    y_i &\sim \text{TruncatedNormal}(\mu, \sigma, -3, 3.2)
\end{align*}

### Model Limitations
- Discussion of limitations (e.g., fixed structure, sensitivity to polynomial degree, difficulty of interpretation)

## Model 2: Hierarchical linear regression, grouped by dominant MBR numerator band

- model introduction (relation to Oreilly and Werdell 2019)

### Model Description and Priors
- brief word description

### Mathematical Description
- latex equations
### Model Limitations
- Discussion of limitations

## Model 3: Similar to Model 2, with group-specific constant likelihood variance  
- model introduction (relation to Oreilly and Werdell 2019)

### Model Description and Priors
- brief word description

### Mathematical Description
- latex equations
### Model Limitations
- Discussion of limitations

## Model 4: Global linear heteroscedastic model 
- model introduction (relation to Oreilly and Werdell 2019)

### Model Description and Priors
- brief word description

### Mathematical Description
- latex equations
### Model Limitations
- Discussion of limitations

## Model 5: Hierarchical Linear Regression with Linear Group-Wise Heteroscedasticity 
- model introduction

### Model Description and Priors
- brief word description of the idea behind the model and the priors used.

\begin{align*}
    \alpha_\mu &\sim \text{Normal}(0, 1) \\
    \alpha_\sigma &\sim \text{Exponential}(1) \\
    \beta_\mu &\sim \text{Normal}(0, 1) \\
    \beta_\sigma &\sim \text{Exponential}(1) \\
    \sigma_{\alpha_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\alpha_\sigma} &\sim \text{Exponential}(1) \\
    \sigma_{\beta_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\beta_\sigma} &\sim \text{Exponential}(1) \\
    \alpha_j &\sim \text{Normal}(\alpha_\mu, \alpha_\sigma) \quad \text{for } j = 1, \dots, J \\
    \beta_j &\sim \text{Normal}(\beta_\mu, \beta_\sigma) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\alpha_j} &\sim \text{Normal}(\sigma_{\alpha_\mu}, \sigma_{\alpha_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\beta_j} &\sim \text{Normal}(\sigma_{\beta_\mu}, \sigma_{\beta_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \mu_i &= \alpha_{g_i} + \beta_{g_i} \cdot \log\text{MBR}_i \quad \text{for } i = 1, \dots, N \\
    \log \sigma_i &= \sigma_{\alpha_{g_i}} + \sigma_{\beta_{g_i}} \cdot \log\text{MBR}_i \\
    \sigma_i &= \exp(\log \sigma_i) \\
    \log\text{Chl}_i &\sim \text{TruncatedNormal}(\mu_i, \sigma_i, -3, 3.2)
\end{align*}

# Supplementary Figures

```{figure} figs/ppc_example.png
---
width: 85%
fig-cap: "Figure S1: Posterior predictive check comparing observed and simulated chlorophyll-a distributions across groups."
---

### 2025-05-21 manuscript/manuscript.qmd
moved eqns to Suppl. Updated model 5 in methods

@@
   - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
-  - A Bayesian workflow is a principled guide to a fully Bayesian study
+  - These models are a principled way to take uncertainty into account
   - 
@@
 
-### 2.3 Overview of Candidate Models
+###  Overview of Candidate Models
 
@@
 - **Model 2**: Hierarchical linear regression, grouped by dominant MBR numerator band  
-- **Model 4**: Model 2 with group-specific constant likelihood variance  
-- **Model 5**: Global linear heteroscedastic model  
-- **Model 6**: Model 2 with group-wise linear heteroscedasticity  
-- **Model 7**: Model 6 plus an added dispersion term for fluorescence-based measurements
+- **Model 3**: Similart to Model 2, with group-specific constant likelihood variance  
+- **Model 4**: Global linear heteroscedastic model  
+- **Model 5**: Similar to Model 2, with group-wise linear heteroscedasticity  
+- **Model 6**: Model 6 plus an added dispersion term for fluorescence-based measurements
 
-The main text focuses on Models 1, 2, 6, and 7, which span the range of key structural variations. Full equations, priors, and diagnostic results for all models are provided in the Supplement.
+The main text focuses on Models 1, 2, 5, and 6, which span the range of key structural variations. Full equations, priors, and diagnostic results for all models are provided in the Supplement.
 
-### 2.4 Representative Model Structures
+### Representative Model Structures
+Models are described here as sets of equations and Directed Acyclic Graphs (DAG), with the former providing numerical details while the latters helps visualize model structure and flow.
 
-#### 2.4.1 Model 1: Bayesian Polynomial OC6 Regression
+#### Model 1: Bayesian 4th Order Polynomia Regression
 
-This model follows the OC6 formulation proposed by O’Reilly et al. (2019), adapted to a Bayesian framework. It uses a fourth-order polynomial in log(MBR) to estimate log(Chl).
-
-This model can be represented by the following set of probabilistic equations:
-
-\begin{align*}
-    \alpha &\sim \text{Normal}(0, 1) \\
-    \beta_1 &\sim \text{Normal}(0, 1) \\
-    \beta_2 &\sim \text{Normal}(0, 1) \\
-    \beta_3 &\sim \text{Normal}(0, 1) \\
-    \beta_4 &\sim \text{Normal}(0, 1) \\
-    \mu &= \alpha + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 \\
-    \sigma &\sim \text{Gamma}(2, 2) \\
-    \text{likelihood} &\sim \text{TruncatedNormal}(\mu, \sigma, -3, 3.3)
-\end{align*}
+This model follows the OC6 formulation proposed by O’Reilly et al.
+(2019), with the difference that the denominator is the sum of $Rrs555$ and $Rrs670$ rather than the mean.
+The model mean is a fourth-order polynomial of $log(MBR)$ to estimate $log(Chl)$.
+The set of proabilistic equations that define this model can be found in the Supplemental Material Section.
+This model can be represented by the following set of probabilistic equations where $X$ represents $log(MBR)$, and $y_i$ correspondonds to the likelihood of the $i^{th}$ $log(Chl)$ observation
 
@@
 
-![](images/oc6_structure.svg)
+![](images/oc6_structure.png)
+
+Model 1 DAG showing model parameters and distributions used as priors.
+The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more).
+The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations contained in the dataframe and show the number of data points available for fitting.
 
@@
 
-#### 2.4.2 Model 2: Hierarchical Linear Regression (4-band Grouping)
+#### Model 2: Hierarchical Linear Regression (4-band Grouping)
 
-This model introduces partial pooling across four groups defined by the dominant MBR numerator band. Each group receives its own slope and intercept, with shared hyperpriors.
+*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510.
+Each group receives its own slope and intercept with shared hyperpriors.
+On the one hand each group has its own submodel that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.
+$X_i$ is the same as for Model 1, i.e.
+$log(MBR)$, assigned a given observation to group $i$ (@fig-eda-mbr).
+This model retains a constant dispersion parameter $σ$, similar to *Model 1*.
+Below is the system of equations corresponding to this model, where $j$ is the group index and $i$ is the $i^{th}$ obasrvation.
+Mathematical description of the model can be found in the Supplementary Material.
+@fig-model2-struct illustrates the model structure.
 
-- Grouping: Rrs411, Rrs443, Rrs489, Rrs510  
-- Partial pooling via shared hyperpriors  
-- Constant σ across all groups
+::: {#fig-model2-struct}
 
-#### 2.4.3 Model 6: Hierarchical Linear Regression with Group-wise Linear σ
+![](images/model2_structure.png)
 
-To accommodate heteroscedasticity, this model allows σ to vary with log(MBR) within each group, modeled as a log-linear relationship. Parameters are estimated hierarchically across groups.
+Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510).
+Model parameter prior mean and variance have Gaussian and Exponential hyperpriors.
+This structure allows the sharing of information across groups at model fitting time.
+:::
 
-- log(σ) = α + β × log(MBR), with group-specific α and β  
-- Transformed via exp(log(σ))  
-- Partial pooling of σ slope/intercept terms
+#### Model 5: Hierarchical Linear Regression with Linear Heteroscedasticity
 
-#### 2.4.4 Model 7: Hierarchical Linear σ + Measurement-Specific Dispersion
+The data plotted in @fig-eda-mbr suggests heteroscedasticity - variance that is not constant.
+Morevoer variance seems to change with the data.
+Models 3 and 4 (see Supplemental Material section) investigated the role of groupwise constant heteroscedasticity and global linear heteroscedasticity, respectively.
+To investigate this further, Model 5 include groupwise linear heteroscedasticity.
+Thus,  for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed to the likelihood.
+Similarly to the model mean $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors.
+Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.
+illustrates the model structure.
+
+::: {#fig-model5-struct}
+
+![](images/model5_structure.png)
+
+Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.
+Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2s.
+:::
+
+####  Model 6:  Measurement-Specific Dispersion
 
@@
 
-### 2.5 Prior Specification and Inference Settings
+### Inference Settings
 
@@
 
-### 2.6 Model Comparison and Predictive Evaluation
+### Model Comparison and Predictive Evaluation
 
@@
 
-## 3. Results
+## Results
 
@@
 
-### 3.1 MCMC Convergence
+### MCMC Convergence
 
@@
 
-### 3.2 Predictive Performance via Leave-One-Out Cross Validation
+### Predictive Performance via Leave-One-Out Cross Validation
 

### 2025-05-21 manuscript/manuscript.qmd
updated methods section for model 6

@@
 
-Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.
+Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.
 Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2s.
@@
 
+::: {#fig-model6-struct}
+
+![](images/model6_structure.png)
+
+Model 6 - Similar to Model 5; multi-level hierarchical linear model partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and variance.
+However the linear variance model gets an additional parameter $γ$ that allows for the model dispersion to depend on the chlorophyll measurement method (fluorescence or HPLC).
+While this doesn't affect future data prediction, for which ground truth mearuement are optional, it does allow the development of insight on the effect of chloroophyll measurement method on model ucertainty.
+:::
+
 ### Inference Settings
@@
 - **Regression terms**: Normal(0, 1) unless otherwise stated  
-- **Hyperpriors**: Normal and HalfNormal for hierarchical parameters  
+- **Hyperpriors**: Normal and Exponential for hierarchical parameters  
 - **Noise terms** (`σ`): Gamma(γ, φ), chosen to reflect plausible dispersion values

### 2025-05-22 manuscript/manuscript.qmd
updated methods section

@@
 
-## 2. Materials and Methods
+## Materials and Methods
 
@@
 
-### 2.1 Data Description and Preparation Overview
+### Data Description and Preparation Overview
 
@@
 
-### 2.2 Bayesian Modeling Framework
+### Bayesian Modeling Framework
 
@@
 
-*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510.
+*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510 (*cf*.
+@fig-eda-mbr).
 Each group receives its own slope and intercept with shared hyperpriors.
-On the one hand each group has its own submodel that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.
-$X_i$ is the same as for Model 1, i.e.
-$log(MBR)$, assigned a given observation to group $i$ (@fig-eda-mbr).
-This model retains a constant dispersion parameter $σ$, similar to *Model 1*.
-Below is the system of equations corresponding to this model, where $j$ is the group index and $i$ is the $i^{th}$ obasrvation.
+On the one hand each group has its own sub-model that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.
 Mathematical description of the model can be found in the Supplementary Material.
@@
 This structure allows the sharing of information across groups at model fitting time.
+
 :::
@@
 
-The data plotted in @fig-eda-mbr suggests heteroscedasticity - variance that is not constant.
-Morevoer variance seems to change with the data.
-Models 3 and 4 (see Supplemental Material section) investigated the role of groupwise constant heteroscedasticity and global linear heteroscedasticity, respectively.
-To investigate this further, Model 5 include groupwise linear heteroscedasticity.
-Thus,  for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed to the likelihood.
-Similarly to the model mean $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors.
+The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data.
+Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively.
+Details of these models can be found in the Supplementary Material section.
+To investigate this further, Model 5 looks at group-wise linear heteroscedasticity.
+Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood.
+Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors.
 Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.
@@
 
-::: {#fig-model6-struct}
+::: {#fig-model6-struct fig-pos="H"}
 
@@
 
-### Inference Settings
+### Prior Specifications and Model Fitting
 
-I used weakly informative priors across all models to balance flexibility and regularization:
+*Priors* - I used weakly informative priors across all models to balance flexibility and regularization.
+For regression parameters  I used Gaussian priors, which in the case of single-level models like Model 1, were given a $0$ mean and variance $1$, to encourage parameter sparsity.
+Hyperpriors of prior means were set to Gaussans of mean 0 and variance 1 to encourage sparsity.
+Hyperpriors of prior variances set to  Exponential distribution with rate set to $$1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise.
+This hyperprior also encourages uncertainty shrinkage across groups if possible.
+The likelihood dispersion $σ$, when constant was set, to a Gamma distribution of shape and rate both set to $2$.
+Prior soundness was assessed using *Prior Predictive Checks*.
+This procedure takes advantage of the fact taht Bayesian models are generative to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting.
+This was plotted as Cumulative Distribution Function (CDF) of $log(Chl)$.
+See *Results* section for a Prior Predictive Checks plot example; plots for all models can be found in the Supplementary Material section.
 
-- **Regression terms**: Normal(0, 1) unless otherwise stated  
-- **Hyperpriors**: Normal and Exponential for hierarchical parameters  
-- **Noise terms** (`σ`): Gamma(γ, φ), chosen to reflect plausible dispersion values
-
-Inference was conducted using:
-
-- 4 chains per model  
-- 1000–1500 tuning steps  
-- ≥2000 posterior samples per chain
-
-I assessed convergence using R-hat (target < 1.01) and effective sample size (ESS > 1000). Full trace plots and diagnostics are included in the Supplement.
+*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior.
+Sampler settings were as follows.
+Four independent sampling runs referred to as *chains* were used to assess sampler convergence.
+Prior to sampling, 1000 tuning steps were used for each chain to maximize sampler efficiency.
+The 2000 samples were recorded for each chain totalling 8000 samples.
+I assessed sampler convergence using the Gelman-Rubin statistic, $\hat{R}$ and effective sample size($ESS$).
+Convergence is acceptable for $\hat{R}$ < 1.01 and $ESS > 1000$.
+Full trace plots and diagnostics are included in the Supplementary Material section.
+Goodness of fit was assessed using *Posterior plots*, which show  *Posterior Predictive Checks, which show the expected
 
@@
 
-I compared model performance using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)**, implemented via the `arviz.loo()` function.
-This Bayesian metric estimates the expected log predictive density (ELPD), a principled measure of out-of-sample predictive accuracy.
+I compared model performance using **Pareto Smoothed Importance Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)** [@vehtari2017].
+This Bayesian metric estimates the expected log predictive density (ELPD), a principled measure of out-of-sample predictive accuracy, that uses the entire posterior predictive distribution
 

### 2025-05-23 manuscript/manuscript.qmd
manuscript update

@@
 citation:
-  container-title: Geophysical Research Letters
+  container-title: Remote Sensing of Environment
 keep-tex: true
@@
 This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates.
-
-- Additional noise component applied when chl_flag = fluorescence  
-- All other structure inherited from Model 6
+A reasonable assumption is that HPLC measurements are less noisy than fluorescence.
+Thus $σ$ is given the expression:
 
@@
 However the linear variance model gets an additional parameter $γ$ that allows for the model dispersion to depend on the chlorophyll measurement method (fluorescence or HPLC).
-While this doesn't affect future data prediction, for which ground truth mearuement are optional, it does allow the development of insight on the effect of chloroophyll measurement method on model ucertainty.
+While this does not necessarily affect future , for which ground truth mearuement are optional, it does allow the development of insight on the effect of chloroophyll measurement method on model ucertainty.
 :::
@@
 
-*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior.
+*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior; see the Supplementary Materials section for more details.
 Sampler settings were as follows.

### 2025-05-23 manuscript/supplement.qmd
manuscript update

@@
 - brief word description of the idea behind the model and the priors used.
-
+### Mathematical Description
 \begin{align*}
@@
 
+## Model 6:
+
+### Mathematical Description
+
+\begin{align*}
+    \alpha_\mu &\sim \text{Normal}(0, 1) \\
+    \alpha_\sigma &\sim \text{Exponential}(1) \\
+    \beta_\mu &\sim \text{Normal}(0, 1) \\
+    \beta_\sigma &\sim \text{Exponential}(1) \\
+    \sigma_{\alpha_\mu} &\sim \text{Normal}(0, 1) \\
+    \sigma_{\alpha_\sigma} &\sim \text{Exponential}(1) \\
+    \sigma_{\beta_\mu} &\sim \text{Normal}(0, 1) \\
+    \sigma_{\beta_\sigma} &\sim \text{Exponential}(1) \\
+    \alpha_j &\sim \text{Normal}(\alpha_\mu, \alpha_\sigma) \quad \text{for } j = 1, \dots, J \\
+    \beta_j &\sim \text{Normal}(\beta_\mu, \beta_\sigma) \quad \text{for } j = 1, \dots, J \\
+    \sigma_{\alpha_j} &\sim \text{Normal}(\sigma_{\alpha_\mu}, \sigma_{\alpha_\sigma}) \quad \text{for } j = 1, \dots, J \\
+    \sigma_{\beta_j} &\sim \text{Normal}(\sigma_{\beta_\mu}, \sigma_{\beta_\sigma}) \quad \text{for } j = 1, \dots, J \\
+    \gamma &\sim \text{Normal}(0, 1) \quad \text{(fluorescence-specific noise effect)} \\
+    \mu_i &= \alpha_{g_i} + \beta_{g_i} \cdot \log\text{MBR}_i \quad \text{for } i = 1, \dots, N \\
+    \log \sigma_i &= \sigma_{\alpha_{g_i}} + \sigma_{\beta_{g_i}} \cdot \log\text{MBR}_i + \gamma \cdot (1 - \text{ChlType}_i) \\
+    \sigma_i &= \exp(\log \sigma_i) \\
+    \log\text{Chl}_i &\sim \text{TruncatedNormal}(\mu_i, \sigma_i, -3, 3.2)
+\end{align*}
+
+## Model Fitting
+
+### The Posterior Distribution
+
+Mathematically the **posterior distribution** of the model parameters, given the observed data, is defined by Bayes' Theorem. For the simpler *Model 1* this looks like:
+
+$$p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma | X) = \frac{p(X | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)}{p(X)}$$
+
+Where:
+* $p(\text{data} | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)$ is the **likelihood function**, which is the product of the individual likelihoods for each observed data point $y_i$:
+    $$
+    p(\text{data} | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) = \prod_{i=1}^{N} \text{TruncatedNormal}(y_i | \mu_i, \sigma, -3, 3.3)
+    $$
+    with
+    $$
+    \mu_i = \alpha + \beta_1 X_i + \beta_2 X_i^2 + \beta_3 X_i^3 + \beta_4 X_i^4
+    $$
+* $p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)$ is the **joint prior distribution** of all parameters, which with the underlying assumption that they are conditionally independent, is the product of their individual prior distributions:
+    $$
+    p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) = p(\alpha) \, p(\beta_1) \, p(\beta_2) \, p(\beta_3) \, p(\beta_4) \, p(\sigma)
+    $$
+    From your model:
+    * $p(\alpha) = \text{Normal}(0, 1)$
+    * $p(\beta_j) = \text{Normal}(0, 1)$ for $j=1,2,3,4$
+    * $p(\sigma) = \text{Gamma}(2, 2)$
+* $p(X)$ is the **marginal likelihood** (also known as the evidence), which is a normalizing constant:
+    $$
+    p(X) = \int p(X | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, d\alpha \, d\beta_1 \, d\beta_2 \, d\beta_3 \, d\beta_4 \, d\sigma
+    $$
+
+With many parameters, these integrals become quick intractable.
+Thus for often then none, the full posterior distribution will have to be approximated.
+The most accurate approximation is Markov Chain Monte Carlo (MCMC) sampling  based on accepted/rejected probabilistic jumps in the posterior space.
+MCMC algorithms have become quite since the early Metropolis-type algorithms.
+In this study I used the NUTS sampler, discussed next.
+
+### Approximating the Posterior with HMC and the NUTS Sampler
+Hamiltonian Monte Carlo (HMC) is an MCMC method that conceptualizes parameters as particles moving on a potential energy surface derived from the negative log-posterior.
+It introduces momentum variables and simulates Hamiltonian dynamics through "leapfrog steps" to propose new parameter values, allowing for more efficient exploration of complex parameter spaces compared to simpler MCMC techniques.
+The No-U-Turn Sampler (NUTS) enhances HMC by adaptively determining the optimal number of leapfrog steps in each iteration, preventing inefficient "U-turns" in the sampling trajectory.
+This makes NUTS a highly efficient and robust algorithm for approximating posterior distributions.
+The pre-print article by @betancourt2018
+
+### Posterior Predictive
+
 # Supplementary Figures
 
-```{figure} figs/ppc_example.png
+```{figure} images/ppc_example.png
 ---

### 2025-05-28 manuscript/manuscript.qmd
updated model 6 explanation

@@
     affiliations:
-      - Independant Researcher
+      - Independent Researcher
   - name: Susanne E. Craig
@@
 The set of proabilistic equations that define this model can be found in the Supplemental Material Section.
-This model can be represented by the following set of probabilistic equations where $X$ represents $log(MBR)$, and $y_i$ correspondonds to the likelihood of the $i^{th}$ $log(Chl)$ observation
+The model structure and inference flow is represented in @fig-model1_struct as a Directed Acyclic Graph (DAG)
 
-::: {#fig-oc6-struct}
+::: {#fig-oc6-struct pos=H}
 
-![](images/oc6_structure.png)
+![](images/model1_structure.png)
 
-Model 1 DAG showing model parameters and distributions used as priors.
+Model 1 Directed Acyclic Graph showing model parameters and their prior distributions and how they relate .
 The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more).
@@
 
-#### Model 2: Hierarchical Linear Regression (4-band Grouping)
+#### Model 2: Hierarchical Linear Regression (HLR)
 
@@
 @fig-eda-mbr).
-Each group receives its own slope and intercept with shared hyperpriors.
-On the one hand each group has its own sub-model that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.
+Each group receives its own slope and intercept, and these parameters have shared hyperpriors.
+Consequently, each group has its own sub-model that can capture details specific to it and the shared hyperpriors allow for the sharing of information between groups when fitting the model.
+This approach maximizes the use of information contained in the available data.
+Hierarchical models are usually result in uncertainty shrinkage.
 Mathematical description of the model can be found in the Supplementary Material.
@@
 
-#### Model 5: Hierarchical Linear Regression with Linear Heteroscedasticity
+#### Model 5: HLR with Added Linear Heteroscedastic Likelihood Dispersion
 
@@
 To investigate this further, Model 5 looks at group-wise linear heteroscedasticity.
-Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood.
+Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood.
 Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors.
@@
 Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.
-Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2s.
+Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2.
 :::
 
-####  Model 6:  Measurement-Specific Dispersion
+####  Model 6:  Augmenting Model 5 with Measurement-Specific Dispersion
 
-This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates.
 A reasonable assumption is that HPLC measurements are less noisy than fluorescence.
-Thus $σ$ is given the expression:
+This model adds a measurement-method-specific dispersion term to the structure of Model 6, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates.
+A reasonable assumption is that HPLC measurements are less noisy than fluorescence.
+The likelihood dispersion, $σ$, is given the expression $log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})$, where $chl_{type\_idx} = 1$ if *in-situ* chlorophyll was measured with HPLC, $0$ otherwise.
+As a result, $σ$ is the same as in Model 5 if $chl_{type\_idx} = 1$, since the extra term is nullified.
+In the case of fluorescence measurement, on the other hand, $log(σ)$ receives an extra noise term, providing an effective measurement of noise difference between the two methods.
+
+@fig-model6-struct shows the inference DAG of Model 6.
 
@@
 
-*Priors* - I used weakly informative priors across all models to balance flexibility and regularization.
-For regression parameters  I used Gaussian priors, which in the case of single-level models like Model 1, were given a $0$ mean and variance $1$, to encourage parameter sparsity.
-Hyperpriors of prior means were set to Gaussans of mean 0 and variance 1 to encourage sparsity.
-Hyperpriors of prior variances set to  Exponential distribution with rate set to $$1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise.
+*Priors* - I used weakly informative priors across all models to balance flexibility with regularization.
+For regression parameters  I used Gaussian priors, In the case of the  single-level  Model 1, this meant priors with $0$ mean and variance of $1$ to encourage parameter sparsity.
+In the case of Models 2-6, which are multi-level models, hyperpriors of prior means were set to Gaussans of mean 0 and variance 1, again to encourage sparsity.
+Hyperpriors of prior variances were  Exponential distributions with rate set to $1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise.
 This hyperprior also encourages uncertainty shrinkage across groups if possible.
-The likelihood dispersion $σ$, when constant was set, to a Gamma distribution of shape and rate both set to $2$.
+In the case of homoscedasticity, or group-wise heteroscedasticity, I used a Gamma prior of shape and scale of $2$ for the likelihood dispersion $σ$, which provides strictly postivie support.
 Prior soundness was assessed using *Prior Predictive Checks*.
-This procedure takes advantage of the fact taht Bayesian models are generative to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting.
-This was plotted as Cumulative Distribution Function (CDF) of $log(Chl)$.
+This procedure takes advantage of the fact taht Bayesian models are generative, meaning the model can be sampled without data, to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting.
+Prior predictive $log(Chl)$ values were plotted as Cumulative Distribution Function (CDF) of $log(Chl)$.
+@fig-model1-priorpc provides
+
+::: {#fig-model1-priorpc fig-pos="H"}
+
+![](images/model1_prior.png)
+
+Prior Predictive Checks, here shown for Model 1.
+This step allows the practitioner to verify the soundness of model assumptions - choice of priors, the model formulation, etc.
+This can be done even before data is collected as the generative nature of Bayesian models means output can be produce based solely on the priors and the model's mathematical expression alone.
+Results are presented as Cumulative Distribution Function.
+In blue, 500 draws were performed each containing a prescribed number of observations.
+The dashed line represents the CDF of the simulation mean, In black, the CDF of the data is shown for comparison.
+
+:::
+
 See *Results* section for a Prior Predictive Checks plot example; plots for all models can be found in the Supplementary Material section.

### 2025-05-28 manuscript/manuscript.qmd
model 6 corrections

@@
 The set of proabilistic equations that define this model can be found in the Supplemental Material Section.
-The model structure and inference flow is represented in @fig-model1_struct as a Directed Acyclic Graph (DAG)
+The model structure and inference flow is represented in @fig-model1-struct as a Directed Acyclic Graph (DAG).
 
-::: {#fig-oc6-struct pos=H}
-
+::: {#fig-model1-struct pos=H}
 ![](images/model1_structure.png)
@@
 The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more).
-The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations contained in the dataframe and show the number of data points available for fitting.
+The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations available.
 
@@
 ::: {#fig-model2-struct}
-
 ![](images/model2_structure.png)
-
 Model 2 - Multi-level hierarchical level with partial pooling across $MBR$ numerator groups (Rrs411, Rrs443, Rrs489, Rrs510).
@@
 This structure allows the sharing of information across groups at model fitting time.
-
 :::
@@
 #### Model 5: HLR with Added Linear Heteroscedastic Likelihood Dispersion
-
 The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data.
@@
 ::: {#fig-model5-struct}
-
 ![](images/model5_structure.png)
-
 Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.
@@
 
-####  Model 6:  Augmenting Model 5 with Measurement-Specific Dispersion
-
+####  Model 6: Impact of Chl Measurement Method on Noise
 A reasonable assumption is that HPLC measurements are less noisy than fluorescence.
-This model adds a measurement-method-specific dispersion term to the structure of Model 6, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates.
-A reasonable assumption is that HPLC measurements are less noisy than fluorescence.
+This model adds a measurement-method-specific dispersion term to the structure of Model 5, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates.
 The likelihood dispersion, $σ$, is given the expression $log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})$, where $chl_{type\_idx} = 1$ if *in-situ* chlorophyll was measured with HPLC, $0$ otherwise.
@@
 In the case of fluorescence measurement, on the other hand, $log(σ)$ receives an extra noise term, providing an effective measurement of noise difference between the two methods.
-
 @fig-model6-struct shows the inference DAG of Model 6.
 
-::: {#fig-model6-struct fig-pos="H"}
+::: {#fig-model6-struct}
 
@@
 In the case of Models 2-6, which are multi-level models, hyperpriors of prior means were set to Gaussans of mean 0 and variance 1, again to encourage sparsity.
-Hyperpriors of prior variances were  Exponential distributions with rate set to $1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise.
+Variance hyperpriors were given Exponential distributions with rate set to $1$ with significant density near 0, whichß encourages similarity across groups unless the data mandates otherwise.
 This hyperprior also encourages uncertainty shrinkage across groups if possible.
@@
 Prior predictive $log(Chl)$ values were plotted as Cumulative Distribution Function (CDF) of $log(Chl)$.
-@fig-model1-priorpc provides
+@fig-model1-priorpc shows such a plot for Model 1.
 
-::: {#fig-model1-priorpc fig-pos="H"}
+::: {#fig-model1-priorpc}
 

### 2025-06-05 manuscript/manuscript.tex
folder update

--- 2025-04-11 manuscript/manuscript.tex
@@
   pdfauthor={Erdem M. Karaköylü; Susanne E. Craig},
+  pdfkeywords={Bayesian modeling, Ocean Color},
   colorlinks=true,
@@
 
-\journalname{Geophysical Research Letters}
+\journalname{Remote Sensing of Environment}
 
@@
 \authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
-\affiliation{1}{Consultant, }\affiliation{2}{NASA, }
+\affiliation{1}{Independent Researcher, }\affiliation{2}{NASA/UMBC, }
 \correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}
@@
 
-\subsection{Introduction}\label{introduction}
+\section{Introduction}\label{introduction}
 
-\subsubsection{Historical Context of Chlorophyll
+Satellite ocean color remote sensing has long served as a cornerstone of
+marine ecosystem monitoring, offering global and synoptic coverage of
+surface ocean properties. Among these, chlorophyll-a (\(Chl_a\))
+concentration remains a central metric, widely used as a proxy for
+phytoplankton biomass, primary production, and water quality. The
+retrieval of \(Chl_a\) from ocean color data has evolved over decades,
+resulting in a diverse lineage of empirical and semi-empirical
+algorithms. The following section summarizes this historical
+development, which sets the stage for a critical examination of the
+statistical foundations underlying current approaches.
+
+\subsection{Historical Context of Chlorophyll
 Algorithms}\label{historical-context-of-chlorophyll-algorithms}
@@
 
-\subsubsection{Limitations of Existing
+\subsection{Limitations of Existing
 Approaches}\label{limitations-of-existing-approaches}
 
-Regrettably, the development of traditional ocean color predictive
-models relies on a fundamental statistical error that plagues most of
-observational science today; that of conflating sampling probability,
-with inferential probability. Consider data \(D\) that include
-predictive features (e.g.~Remote sensing reflectance - Rrs) and
-prediction targets (e.g.~\(Chl_a\) or phytoplankton absorption), and a
-model \(M\) (e.g.~OCx) hypothesized to adequately represent the
-statistical association between the two. The sampling probability
-\(p(D|M)\) also referred to as the likelihood, is the probability of
-\(D\) conditioned on \(M\) being ``true''. The common model fitting
-practice is to maximize \(p(D|M)\) - likelihood maximization - achieved
-by tuning \(M\)'s parameters. The unspoken assumption is that making
-\(M\) fit \(D\) as well as possible yields the best version of \(M\)
-given \(D\) i.e.~maximizing \(p(M|D)\), the inferential probability.
-This sneaky swapping from \(p(D|M)\) to \(p(M|D)\) is a violation of the
-rules of conditional probability, Bayes' theorem in particular.
-Jaynes(\textbf{jaynes2003?}) and later Clayton through his highly
-readable account (\textbf{clayton?}) are some of the more notable
-whistleblowers. Occasionally, in simple or well-behaved problems, with a
-large amount of data and no prior information available the maxima of
-both \(p(M|D)\) and \(p(D|M)\) can indeed converge on the same answer.
-Yet there is no guarantee predictability for this outcome however, and
-the fallout of this abuse is non-negligible, most noticeably in terms of
-models that don't generalize well, ad-hoc uncertainty estimation
-attempts, and more broadly a replication crisis.
+Regrettably, the development of traditional ocean color algorithms is
+grounded in a fundamental statistical error---one that pervades much of
+observational science: the conflation of sampling probability with
+inferential probability (Jaynes \& Bretthorst, 2003;
+\textbf{DeScheemaekere2011?}).
 
-Some researchers have attempted to deal with these issues. (Seegers et
-al., 2018) e.g.~have proposed alternative metrics to circumvent the
-inadequate assumptions of the frequentist approach. Others have tried to
-go a step futher and switch to the Bayesian paradigm. E.g. (Frouin \&
-Pelletier, 2013) have proposed a Bayesian inversion scheme for
-atmospheric correction. (\textbf{shi2015?}) proposed a probabilistic
-method to merge data from different sensors. (Craig \& Karaköylü, 2019)
-have proposed a Bayesian neural network approach using Hamiltonian Monte
-Carlo sampling to retrieve Inherent Optical Properties (IOP) from
-Top-of-the-atmosphere (TOA) radiance. (\textbf{werther2022?}) used
-Monte-Carlo dropout as deep neural net training method to obtain
-prediction uncertainty. (Erickson et al., 2023) have proposed using
-conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a
-forward model. (\textbf{hammout2024?}) have proposed a Bayesian Neural
-Network trained using Stochastic Variational Inference to predict
-\(Chl_a\) from ocean color observations. Yet most of these approaches
-retain variable levels of frequentism by applying only part of what is
-commonly referred to as the Bayesian workflow(\textbf{betancourt2019?}).
+Consider a dataset \(D\) composed of input--output pairs---e.g., remote
+sensing reflectance (Rrs) and chlorophyll-a concentration
+(\(Chl_a\))---and a model \(M\), such as OCx, posited to represent the
+relationship between them. The sampling probability \(p(D \mid M)\)
+denotes the probability of observing data \(D\) under the assumption
+that model \(M\) is true. In standard model fitting, this likelihood is
+maximized by adjusting the parameters of \(M\) to best explain the
+observed data.
 
-In this paper we propose a modified workflow to Bayesian modeling for
-Ocean Color emote sensing. We propose that this workflow include; (1)
-one or more model building, (2) prior formulation via expert knowledge
-elicitation and/or data-free simulation, (3) fitting model to data, (4)
-model validation, comparison and selection, (5) interpretation of
-results via posterior analysis, (6) prediction on new observation, (7)
-update the posterior when new labeled data becomes available. As
-illustration we recast some of the pre-existing \(OCx\) models into
-their probabilistic version. To illustrate model comparison and
-selection we also build and evaluate Bayesian Additive Tree Regregssion
-(BART; (Chipman et al., 2010)) type model.
+This approach tacitly assumes that the model which best fits the data
+also most accurately represents the underlying generative process. This
+constitutes an epistemic fallacy---treating \(p(D \mid M)\) as if it
+were \(p(M \mid D)\)---a direct violation of Bayes' theorem and the
+rules governing conditional probability.
 
-\subsection{Methods}\label{methods}
+Although in well-behaved, data-rich cases---where the likelihood is
+regular, the signal strong, and the model adequately constrained---the
+maxima of \(p(D \mid M)\) and \(p(M \mid D)\) may coincide, this remains
+the exception---not the rule.
 
-\subsubsection{Prior Elicitation and Model
-Formulation}\label{prior-elicitation-and-model-formulation}
+This mistake lies at the heart of what Clayton (Clayton, 2022) terms the
+Bernoulli Fallacy: the widespread misinterpretation of likelihood as
+inference, or of data-fit as belief. As Clayton argues, this logical
+misstep has far-reaching consequences, with implications that extend
+beyond science to domains such as medicine, law, and public policy.
 
-\subsubsection{Data Preprocessing}\label{data-preprocessing}
+In scientific modeling, this fallacy contributes to poor generalization,
+drives the use of ad hoc or retrospective uncertainty quantification,
+and underlies many published results that later prove difficult to
+replicate (Baker, 2016; Cobey et al., 2024). These limitations are not
+restricted to classical hypothesis testing; they persist in the training
+and deployment of modern machine learning models as well.
 
-\subsubsection{}\label{section}
+In regression and classification, maximizing likelihood is often treated
+as sufficient for inference---despite yielding only a single point
+estimate and ignoring both parameter uncertainty and the plausibility of
+alternative models.
 
-\section{Data Preprocessing}
+This epistemic shortcut has been directly critiqued in the machine
+learning literature. Gal (Gal, 2016) and Ghahramani
+(\textbf{gahramani2015probabilistic?}) point out that most ML models
+discard uncertainty altogether, treating the outcome of an optimization
+as if it were an inference. The result is overconfident predictions and
+brittle generalization---concerns that echo Clayton's critique.
 
-Data for this study were acquired from multiple satellite ocean color
-sensors and corresponding in situ chlorophyll-(a) measurements obtained
-from sources such as the NASA Bio-Optical Marine Algorithm Data set
-(NOMAD) and the compilation by Valente et al.~(2015). To ensure
-consistency across sensors, the spectral reflectance data ((R\_\{rs\}))
-were interpolated as needed to common wavelength centers.
+Bishop (Bishop, 2006) similarly distinguishes between the utility of
+predictive models and the inferential scaffolding required to quantify
+uncertainty, reinforcing the notion that likelihood alone is
+insufficient---and that the Bernoulli Fallacy permeates much of applied
+machine learning.
 
-For the empirical (OCx) formulation, blue-to-green band ratios were
-computed for each observation. In particular, the maximum band ratio
-(MBR) was determined by taking the highest value among the available
-blue-band ratios (e.g., (Rrs(443)/Rrs(555)), (Rrs(490)/Rrs(555)), and
-(Rrs(510)/Rrs(555))). This maximum value was then log-transformed: \[
-\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
-\]
+\subsection{Overcoming limitations}\label{overcoming-limitations}
 
-For the Color Index (CI) formulation of Hu et al.~(2012), the CI was
-calculated as: \[
-\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
-\] and the corresponding in situ chlorophyll-(a) concentrations were
-log-transformed: \[
-\log \text{Chl} = \log_{10}(\text{Chl}).
-\]
+Several recent efforts have attempted to address the limitations of
+classical retrieval models. For instance, (Seegers et al., 2018)
+proposed alternative evaluation metrics to move beyond restrictive
+frequentist assumptions. Others have incorporated Bayesian elements into
+the modeling pipeline: (Frouin \& Pelletier, 2013) applied Bayesian
+inversion for atmospheric correction, (\textbf{shi2015?}) used
+probabilistic fusion for multi-sensor data, and (Craig \& Karaköylü,
+2019) employed Hamiltonian Monte Carlo to train Bayesian neural networks
+(BNNs) for retrieving inherent optical properties (IOPs) from
+top-of-atmosphere radiance. Similarly, (\textbf{werther2022?})
+introduced Monte Carlo dropout as an approximate BNN strategy, while
+(Erickson et al., 2023) recast the Generalized Inherent Optical Property
+(GIOP) framework using conjugate Bayesian linear models. Most recently,
+(\textbf{hammout2024?}) developed a BNN surrogate using stochastic
+variational inference for chlorophyll-a prediction from Rrs.
 
-These transformations standardize the data to a common scale, ensuring
-that variability is appropriately captured for subsequent regression and
-uncertainty quantification. Detailed descriptions of the interpolation
-methods and quality control procedures are provided in the Supplementary
-Material.
+While these studies mark important progress, they often adopt only
+isolated components of what has come to be known as the Bayesian
+workflow (\textbf{bgelman2019?}). Key aspects---such as principled prior
+specification, posterior predictive checking, and formal model
+comparison---remain largely absent. As a result, uncertainty is
+frequently approximated rather than inferred, and model structure is
+rarely treated as a variable to be interrogated.
 
-\section{Statement of Contribution}\label{statement-of-contribution}
+This paper builds on prior work and extends it by applying a fully
+Bayesian modeling framework for chlorophyll-a prediction from sea
+surface reflectance (Rrs), with explicit treatment of uncertainty, model
+complexity, and measurement error.
 
-In this study, we develop and demonstrate a new global chlorophyll
-retrieval model based on Bayesian Additive Regression Trees implemented
-in PyMC. We train the BART model on a large, standardized dataset of
-satellite remote-sensing reflectance (Rrs) spectra matched with in situ
-chlorophyll measurements, using log₁₀-transformed Chl-a as the response
-to stabilize variance. The resulting model is applied globally to
-produce chlorophyll-a estimates from multi-spectral satellite data, with
-associated uncertainty estimates for each prediction. We show that this
-Bayesian tree-based model can serve as a general-purpose ocean color
-algorithm that is sensor-agnostic (provided reflectances are harmonized
-to common wavebands), interpretable, and uncertainty-aware. Unlike
-conventional empirical algorithms, the BART approach allows users to
-examine the inferred Rrs--Chl relationships and trust the model's
-performance across regimes, while also quantifying confidence in each
-retrieval. This work thus contributes a novel methodological advance to
-satellite ocean color science: a unified chlorophyll retrieval model
-that marries the strengths of empirical algorithms (global applicability
-and simplicity) with the benefits of modern Bayesian machine learning
-(flexibility, interpretability, and rigorous uncertainty
-quantification). Our introduction of BART for global chlorophyll
-prediction opens the door for more robust monitoring of ocean
-biogeochemistry and improved integration of ocean color data into
-scientific and management applications.
+\section{Materials and Methods}\label{materials-and-methods}
+
+This section outlines the dataset, model structures, inference methods,
+and model evaluation criteria used in this study. I focus on a fully
+Bayesian approach to modeling chlorophyll-a from satellite remote
+sensing reflectance (Rrs), emphasizing transparency, uncertainty
+quantification, and principled model comparison. Four models are
+presented in the main text, while all six model formulations and results
+are included in the Supplement.
+
+\subsection{Data Description and Preparation
+Overview}\label{data-description-and-preparation-overview}
+
+For pedagogical reasons, I used the well known NOMAD (Werdell \& Bailey,
+2005) dataset, which includes quality-controlled in situ measurements of
+chlorophyll-a matched with coincident satellite Rrs data. These
+measurements span a range of oceanographic conditions, enabling the
+development of models that are expected to generalize across water
+types. The data were loaded using the Python Pandas library (team,
+2020), which was used for all subsequent cleaning, preprocessing, and
+transformation operations.
+
+\emph{Raw Data} - Sea-surface reflectances (Rrs) in the 6 visible
+SeaWiFS bands (411, 443, 489, 510, 550, 670nm) constitue the bulk of the
+data. Chlorophyll concentrations, measured via fluorescence or HPLC or
+both, consituted the target variables. Any row containing an invalid
+data point (null, zero, or negative) was removed, as proper
+missing/invalid data imputation is a subject onto itself and beyond the
+scope of this paper.
+
+\emph{Data Preprocessing and Transformation} - Chlorophyll measurements
+from the two different methods were combined into a single data. An
+auxiliary column was used to flag the measurement method as `fluo' of
+`hplc'. The chlorophyll was then log-transformed to stabilize its
+variance. The max band ratio (MBR) was computed as
+\(\frac {\max(Rrs411, Rrs443, Rrs489, Rrs)}{Rrs555 + Rrs670}\), similar
+to the OC6 model proposed by (John E. O'Reilly \& Werdell, 2019).
+Another ancillary column was created to captured the actual Rrs band
+used for each set observations. This numerator type as well as the
+chlorophyll measurement type are useful to build more hierarchical
+partial pooling models that can take advantage of such data groups. A
+sample from the fully processed data used in the subsequent models can
+be seen in (\textbf{data-table?}).
+
+\phantomsection\label{data-table}
+\begin{longtable}[]{@{}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.0641}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.3077}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.1538}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.3077}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.1667}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+log\_MBR
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+MBR\_flag
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+log\_chl
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+hplc\_flag
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+905 & -0.2545615131431814 & Rrs510 & 0.5837653682849998 & hplc \\
+91 & -0.3641327253450254 & Rrs510 & 1.0511525224473812 & fluo \\
+528 & -0.0737170739918758 & Rrs510 & 0.14363923527454328 & fluo \\
+969 & 0.5813600497877703 & Rrs411 & -0.8272803386505035 & fluo \\
+552 & 0.7134341855521412 & Rrs443 & -0.9318141382538383 & hplc \\
+\end{longtable}
+
+A random 5-row sample from the Pandas dataframe holding the data used in
+subsequent models. \emph{log\_MBR} is the log-transformed Max. Band
+Ratio (MBR) of sea surface reflectances, where the numerator is
+Max(Rrs411, Rrs443, Rrs489, Rrs510) and the denominator is computed from
+the sum of Rrs555 and Rrs670. \emph{MBR\_flag} indicates which of the 4
+possible Rrs bands is used in the numerator for that particular row.
+\emph{log\_chl} is log transformed in-situ measured chlorophyll \emph{a}
+concentration, originally in \(mg\ m^{-3}\). \emph{hplc\_flag} is a
+binary variable indicating whether the measurement was obtained via HPLC
+or fluorometry. ``-flag'' are used during modeling for grouping data for
+hierarchical partial pooling models, explained further below. These
+groupings can be seen in Figure~\ref{fig-eda}
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/eda_numerator.png}}
+
+}
+
+\subcaption{\label{fig-eda-mbr}}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/eda_hplc.png}}
+
+}
+
+\subcaption{\label{fig-eda-hplc}}
+
+\end{minipage}%
+
+\caption{\label{fig-eda}Two ways of grouping the data.
+Figure~\ref{fig-eda-mbr}, max band ratio numerator;
+Figure~\ref{fig-eda-hplc}, chlorophyll measurement method.}
+
+\end{figure}%
+
+\subsection{Bayesian Modeling
+Framework}\label{bayesian-modeling-framework}
+
+Several models of increasing levels were built. All models use
+\(log(MBR)\) as primary input and \(log(chl)\) as target, some use
+\(MBR\_flag\) as additional input, the last model also uses
+\(hlpc_flag\). The following are key components shared across models.
+
+\emph{Likelihood} - A truncated gaussian distribution with support
+constrained to -3.0 to 3.2 on the \(log_{10}\) scale,corresponding to
+\(0.001 - 1600 mg\ m^{-3}\) to reflect both detectable realistic levels
+of marine surface chlorophyll.
+
+\emph{Noise Structure} - Noise was incorporated as likelihood dispersion
+parameter \(σ\); some of the models use a constant \(σ\), others take
+heteroscedasticity into account.
+
+\emph{Priors} - To estimate model posterior distribution, and produce
+predictions with uncertainty following data fitting, all model
+parameters receive a prior distribution, making explicit model
+assumptions. Often there will be enough data to overrule the priors,
+which are here mostly a conduit to the Posterior. An important caveat is
+that care should be taken not to assign a 0-probability through a poorly
+specified prior where support might actually be needed. Less
+importantly, more efficient use of the search space can be done through
+priors that assign 0-probability where support is clearly unreasonable.
+Thus, all regression parameters receive Normal (Gaussian) 0-centered
+weakly informative priors. And all \(σ\) terms received Gamma priors
+that provide flexible but strictly positive support.
+
+\emph{Software} - All models were implemented with the Python
+probabilistic programming library PyMC V.5 (Abril-Pla et al., 2023) and
+estimated using a variant of Hamiltonian Monte Carlo known as the
+No-U-Turn Sampler (NUTS). Model diagnostics inference evaluation and
+model comparisons were done through the Arviz package.
+
+\subsection{Overview of Candidate
+Models}\label{overview-of-candidate-models}
+
+I developed six models representing a progression of modeling complexity
+and assumptions:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Model 1}: Polynomial regression (Bayesian OC6-style
+  baseline)\\
+\item
+  \textbf{Model 2}: Hierarchical linear regression, grouped by dominant
+  MBR numerator band\\
+\item
+  \textbf{Model 3}: Similart to Model 2, with group-specific constant
+  likelihood variance\\
+\item
+  \textbf{Model 4}: Global linear heteroscedastic model\\
+\item
+  \textbf{Model 5}: Similar to Model 2, with group-wise linear
+  heteroscedasticity\\
+\item
+  \textbf{Model 6}: Model 6 plus an added dispersion term for
+  fluorescence-based measurements
+\end{itemize}
+
+The main text focuses on Models 1, 2, 5, and 6, which span the range of
+key structural variations. Full equations, priors, and diagnostic
+results for all models are provided in the Supplement.
+
+\subsection{Representative Model
+Structures}\label{representative-model-structures}
+
+Models are described here as sets of equations and Directed Acyclic
+Graphs (DAG), with the former providing numerical details while the
+latters helps visualize model structure and flow.
+
+\subsubsection{Model 1: Bayesian 4th Order Polynomia
+Regression}\label{model-1-bayesian-4th-order-polynomia-regression}
+
+This model follows the OC6 formulation proposed by O'Reilly et
+al.~(2019), with the difference that the denominator is the sum of
+\(Rrs555\) and \(Rrs670\) rather than the mean. The model mean is a
+fourth-order polynomial of \(log(MBR)\) to estimate \(log(Chl)\). The
+set of proabilistic equations that define this model can be found in the
+Supplemental Material Section. The model structure and inference flow is
+represented in Figure~\ref{fig-model1-struct} as a Directed Acyclic
+Graph (DAG).
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/model1_structure.png}}
+
+}
+
+\caption{\label{fig-model1-struct}Model 1 Directed Acyclic Graph showing
+model parameters and their prior distributions and how they relate . The
+intercept \(α\) and the polynomial slopes, \(β1\)-\(β4\) all have
+Gaussian priors; \(σ\), the likelihood dispersion parameter, receives a
+Gamma prior; the likelihood itself is constrained by a Truncated Normal
+distribution (see text for more). The plate containing the likelihood is
+indexed (\(obs\_idx\)) indicating the likelihood will be computed for
+each of the 1082 observations available.}
+
+\end{figure}%
+
+\subsubsection{Model 2: Hierarchical Linear Regression
+(HLR)}\label{model-2-hierarchical-linear-regression-hlr}
+
+\emph{Model 2} introduces partial pooling across four groups defined by
+the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510
+(\emph{cf}. Figure~\ref{fig-eda-mbr}). Each group receives its own slope
+and intercept, and these parameters have shared hyperpriors.
+Consequently, each group has its own sub-model that can capture details
+specific to it and the shared hyperpriors allow for the sharing of
+information between groups when fitting the model. This approach
+maximizes the use of information contained in the available data.
+Hierarchical models are usually result in uncertainty shrinkage.
+Mathematical description of the model can be found in the Supplementary
+Material. Figure~\ref{fig-model2-struct} illustrates the model
+structure.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/model2_structure.png}}
+
+}
+
+\caption{\label{fig-model2-struct}}
+
+\end{figure}%
+
+\subsubsection{Model 5: HLR with Added Linear Heteroscedastic Likelihood
+Dispersion}\label{model-5-hlr-with-added-linear-heteroscedastic-likelihood-dispersion}
+
+The data plotted in Figure~\ref{fig-eda} suggests heteroscedasticity -
+variance that is not constant, but rather changes depending on the data.
+Models 3 and 4 investigate the role of group-wise constant
+heteroscedasticity and global linear heteroscedasticity, respectively.
+Details of these models can be found in the Supplementary Material
+section. To investigate this further, Model 5 looks at group-wise linear
+heteroscedasticity. Thus, for a given MBR group \(j\) \(σ_j\) is modeled
+as a log-linear relationship such that \(log(σ) = α_j + β_j × log(MBR)\)
+and then transformed via \(exp(log(σ_j))\) before being passed on to the
+likelihood. Similar to the model mean, \(μ_j\), \(σ_j\)'s slope and
+intercept terms are partially pooled and so have common hyperpriors.
+Mathematical descriiption of the model can be found in the Supplementary
+Material section; Figure~\ref{fig-model5-struct}. illustrates the model
+structure.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/model5_structure.png}}
+
+}
+
+\caption{\label{fig-model5-struct}}
+
+\end{figure}%
+
+\subsubsection{Model 6: Augmenting Model 5 with Measurement-Specific
+Dispersion}\label{model-6-augmenting-model-5-with-measurement-specific-dispersion}
+
+A reasonable assumption is that HPLC measurements are less noisy than
+fluorescence. This model adds a measurement-method-specific dispersion
+term to the structure of Model 6, to capture the suspected increase in
+variability associated with fluorescence-based chlorophyll estimates.
+The likelihood dispersion, \(σ\), is given the expression
+\(log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})\),
+where \(chl_{type\_idx} = 1\) if \emph{in-situ} chlorophyll was measured
+with HPLC, \(0\) otherwise. As a result, \(σ\) is the same as in Model 5
+if \(chl_{type\_idx} = 1\), since the extra term is nullified. In the
+case of fluorescence measurement, on the other hand, \(log(σ)\) receives
+an extra noise term, providing an effective measurement of noise
+difference between the two methods. Figure~\ref{fig-model6-struct} shows
+the inference DAG of Model 6.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/model6_structure.png}}
+
+}
+
+\caption{\label{fig-model6-struct}Model 6 - Similar to Model 5;
+multi-level hierarchical linear model partial pooling across \(MBR\)
+groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and
+variance. However the linear variance model gets an additional parameter
+\(γ\) that allows for the model dispersion to depend on the chlorophyll
+measurement method (fluorescence or HPLC). While this does not
+necessarily affect future , for which ground truth mearuement are
+optional, it does allow the development of insight on the effect of
+chloroophyll measurement method on model ucertainty.}
+
+\end{figure}%
+
+\subsection{Prior Specifications and Model
+Fitting}\label{prior-specifications-and-model-fitting}
+
+\emph{Priors} - I used weakly informative priors across all models to
+balance flexibility with regularization. For regression parameters I
+used Gaussian priors, In the case of the single-level Model 1, this
+meant priors with \(0\) mean and variance of \(1\) to encourage
+parameter sparsity. In the case of Models 2-6, which are multi-level
+models, hyperpriors of prior means were set to Gaussans of mean 0 and
+variance 1, again to encourage sparsity. Variance hyperpriors were given
+Exponential distributions with rate set to \(1\) with significant
+density near 0, whichß encourages similarity across groups unless the
+data mandates otherwise. This hyperprior also encourages uncertainty
+shrinkage across groups if possible. In the case of homoscedasticity, or
+group-wise heteroscedasticity, I used a Gamma prior of shape and scale
+of \(2\) for the likelihood dispersion \(σ\), which provides strictly
+postivie support. Prior soundness was assessed using \emph{Prior
+Predictive Checks}. This procedure takes advantage of the fact taht
+Bayesian models are generative, meaning the model can be sampled without
+data, to insure that values of (\(log(Chl)\)) predicted by the model are
+reasonable even before model fitting. Prior predictive \(log(Chl)\)
+values were plotted as Cumulative Distribution Function (CDF) of
+\(log(Chl)\). Figure~\ref{fig-model1-priorpc} shows such a plot for
+Model 1.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/model1_prior.png}}
+
+}
+
+\caption{\label{fig-model1-priorpc}Prior Predictive Checks, here shown
+for Model 1. This step allows the practitioner to verify the soundness
+of model assumptions - choice of priors, the model formulation, etc.
+This can be done even before data is collected as the generative nature
+of Bayesian models means output can be produce based solely on the
+priors and the model's mathematical expression alone. Results are
+presented as Cumulative Distribution Function. In blue, 500 draws were
+performed each containing a prescribed number of observations. The
+dashed line represents the CDF of the simulation mean, In black, the CDF
+of the data is shown for comparison.}
+
+\end{figure}%
+
+See \emph{Results} section for a Prior Predictive Checks plot example;
+plots for all models can be found in the Supplementary Material section.
+
+\emph{Model Fitting} - The model posterior was approximated using the No
+U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian
+Monte Carlo (HMC), which uses particle physics to sample the posterior;
+see the Supplementary Materials section for more details. Sampler
+settings were as follows. Four independent sampling runs referred to as
+\emph{chains} were used to assess sampler convergence. Prior to
+sampling, 1000 tuning steps were used for each chain to maximize sampler
+efficiency. The 2000 samples were recorded for each chain totalling 8000
+samples. I assessed sampler convergence using the Gelman-Rubin
+statistic, \(\hat{R}\) and effective sample size(\(ESS\)). Convergence
+is acceptable for \(\hat{R}\) \textless{} 1.01 and \(ESS > 1000\). Full
+trace plots and diagnostics are included in the Supplementary Material
+section. Goodness of fit was assessed using \emph{Posterior plots},
+which show *Posterior Predictive Checks, which show the expected
+
+\subsection{Model Comparison and Predictive
+Evaluation}\label{model-comparison-and-predictive-evaluation}
+
+I compared model performance using \textbf{Pareto Smoothed Importance
+Sampling Leave-One-Out Cross-Validation (PSIS-LOO-CV)} (Vehtari et al.,
+2017). This Bayesian metric estimates the expected log predictive
+density (ELPD), a principled measure of out-of-sample predictive
+accuracy, that uses the entire posterior predictive distribution
+
+\begin{itemize}
+\tightlist
+\item
+  Models were ranked by ELPD\\
+\item
+  ΔELPD values were interpreted relative to their standard errors\\
+\item
+  Pareto-k diagnostics confirmed LOO reliability (full table in
+  Supplement)
+\end{itemize}
+
+I evaluated predictive accuracy and calibration using:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Posterior Predictive Checks (PPCs)}:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Simulated draws overlaid on observed distributions\\
+  \item
+    Empirical CDF comparisons and interval coverage
+  \end{itemize}
+\item
+  \textbf{HDI coverage}:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Fraction of observed values falling within 95\% posterior predictive
+    intervals
+  \end{itemize}
+\end{itemize}
+
+No frequentist evaluation criteria such as RMSE, MAE, or R² were used at
+any stage as these do not make sense in this context where for each
+observations, a probability distribution of the prediction is produced.
+
+\section{Results}\label{results}
+
+This section presents the outputs of posterior inference, model
+comparison using Bayesian criteria, and posterior predictive diagnostics
+for the four primary models. Results from the three supplementary models
+are provided separately in the Supplement.
+
+\subsection{MCMC Convergence}\label{mcmc-convergence}
+
+All four models converged successfully. The Gelman-Rubin statistic
+(R-hat ≈ 1.0) and large effective sample sizes (ESS \textgreater{} 1000)
+across all parameters indicate robust posterior estimation. Diagnostic
+plots and sampling traces are provided in the Supplementary Material.
+
+\begin{itemize}
+\tightlist
+\item
+  All R-hat \textless{} 1.01
+\item
+  No divergences
+\item
+  Sufficient tail and bulk ESS
+\item
+  Full diagnostics in Supplement
+\end{itemize}
+
+\subsection{Predictive Performance via Leave-One-Out Cross
+Validation}\label{predictive-performance-via-leave-one-out-cross-validation}
+
+We compared models using Pareto Smoothed Importance Sampling
+Leave-One-Out Cross-Validation (PSIS-LOO-CV), a fully Bayesian approach
+to evaluating out-of-sample predictive accuracy.
+
+\textbf{Table 1.} Leave-One-Out Comparison (ELPD ± SE):
+
+\begin{itemize}
+\tightlist
+\item
+  Model 7 ranked highest in expected log predictive density (ELPD)
+\item
+  Model 6 performed comparably, with ΔELPD within 1 SE
+\item
+  Model 1 (OC6 polynomial) had the weakest predictive performance
+\item
+  All models had acceptable Pareto-k diagnostics (see Supplement)
+\end{itemize}
+
+\subsection{3.3 Posterior Predictive Checks
+(PPC)}\label{posterior-predictive-checks-ppc}
+
+We used posterior predictive checks (PPCs) to evaluate the ability of
+each model to reproduce key features of the observed log-chlorophyll
+distribution. These include:
+
+\begin{itemize}
+\tightlist
+\item
+  Empirical CDF comparisons
+\item
+  HDI envelopes overlaid on the data
+\item
+  Predictive median vs.~observed values
+\end{itemize}
+
+\textbf{Figure 1.} Posterior predictive envelopes vs.~observed
+log(Chl)\\
+\textbf{Figure 2.} ECDF overlays with posterior predictive means and
+HDIs\\
+\textbf{Figure 3.} Rug plot of observations with overlaid posterior
+predictive draws
+
+Findings: - Model 7 captures full data distribution, especially for
+fluorescence-labeled values - Model 1 underestimates tails
+(under-dispersed) - Model 6 accurately reflects group-level variance
+
+\subsection{3.4 Posterior Distributions and
+Interpretability}\label{posterior-distributions-and-interpretability}
+
+Posterior distributions for key parameters illustrate how model
+structure affects inference. We show both: - Group-level slopes and
+intercepts (hierarchical structure) - Group-wise σ (heteroscedasticity)
+- Fluorescence-specific variance term (Model 7)
+
+\textbf{Figure 4.} Forest plot of slope and intercept posteriors\\
+\textbf{Figure 5.} Posterior of σ across MBR groups\\
+\textbf{Figure 6.} Posterior of added noise term for fluorescence
+
+Key insights: - Rrs510-dominant group had the steepest slope (Model 2 \&
+6) - Fluorescence noise term is credibly nonzero (Model 7) - Posterior
+variance structure supports heteroscedasticity as a key feature
+
+\subsection{3.5 Summary of Bayesian
+Comparison}\label{summary-of-bayesian-comparison}
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Model 7}: Best predictive performance (highest ELPD), best
+  uncertainty calibration
+\item
+  \textbf{Model 6}: Nearly equivalent performance; simpler if
+  fluorescence uncertainty not needed
+\item
+  \textbf{Model 2}: Gains from partial pooling; underperforms without
+  variance modeling
+\item
+  \textbf{Model 1}: Legacy structure underfits tails, overconfident
+  predictions
+\end{itemize}
+
+All results support the Bayesian hierarchical modeling framework with
+heteroscedasticity and measurement-aware noise as optimal for
+chlorophyll-a estimation.
+
+\section{4. Discussion}\label{discussion}
+
+This section interprets the results in the context of previous
+chlorophyll-a modeling work, with particular emphasis on the added value
+of Bayesian modeling. We highlight the importance of accounting for
+heteroscedasticity and measurement-type uncertainty, and discuss how
+this contributes to improved predictive reliability and scientific
+interpretability.
+
+\subsection{4.1 Value of Bayesian Framework over Classical
+Approaches}\label{value-of-bayesian-framework-over-classical-approaches}
+
+Our results demonstrate the clear advantages of fully Bayesian modeling
+for chlorophyll-a prediction from satellite Rrs. Unlike deterministic or
+frequentist models:
+
+\begin{itemize}
+\tightlist
+\item
+  Bayesian methods explicitly model parameter uncertainty and provide
+  HDIs instead of misleading confidence intervals
+\item
+  Posterior predictive distributions enable direct evaluation of model
+  calibration
+\item
+  Bayesian model comparison via LOO-CV avoids overfitting and guards
+  against inappropriate complexity
+\end{itemize}
+
+\subsection{4.2 Importance of Hierarchical
+Structuring}\label{importance-of-hierarchical-structuring}
+
+Models that leveraged group-wise structure (based on dominant Rrs bands)
+consistently outperformed simpler counterparts.
+
+\begin{itemize}
+\tightlist
+\item
+  Hierarchical partial pooling improved both prediction and parameter
+  identifiability
+\item
+  Group-level trends reflected known biogeophysical distinctions (e.g.,
+  clear vs.~turbid water types)
+\item
+  Intercepts and slopes varied systematically across spectral regimes
+\end{itemize}
+
+\subsection{4.3 Role of Heteroscedasticity in Chlorophyll
+Retrieval}\label{role-of-heteroscedasticity-in-chlorophyll-retrieval}
+
+Accounting for non-constant noise was essential for accurate uncertainty
+quantification.
+
+\begin{itemize}
+\tightlist
+\item
+  Models with linear or hierarchical σ terms captured increased variance
+  at higher MBR or Chl
+\item
+  Constant-σ models (e.g., OC6) underrepresented predictive uncertainty,
+  especially in extremes
+\item
+  Explicit modeling of σ structure produced calibrated HDIs and superior
+  posterior predictive checks
+\end{itemize}
+
+\subsection{4.4 Measurement-Specific Error
+Modeling}\label{measurement-specific-error-modeling}
+
+Model 7, which introduced a separate noise term for fluorescence-derived
+Chl, outperformed all others.
+
+\begin{itemize}
+\tightlist
+\item
+  Fluorescence-based measurements have higher and more variable error
+\item
+  This heterogeneity cannot be ignored in merged datasets
+\item
+  Bayesian modeling allows this variance to be encoded explicitly
+  without discarding data
+\end{itemize}
+
+\subsection{4.5 Implications for Satellite Algorithm
+Development}\label{implications-for-satellite-algorithm-development}
+
+The findings support incorporating Bayesian elements into future
+operational chlorophyll algorithms.
+
+\begin{itemize}
+\tightlist
+\item
+  Hierarchical modeling offers a path to generalize across bioregions
+  and sensors
+\item
+  Posterior uncertainty estimates can support downstream applications
+  (e.g., forecasting, ecological thresholds)
+\item
+  Bayesian frameworks are sensor-agnostic: Rrs inputs can vary as long
+  as priors are adapted
+\end{itemize}
+
+\subsection{4.6 Limitations and Future
+Work}\label{limitations-and-future-work}
+
+As with any modeling effort, our study has limitations:
+
+\begin{itemize}
+\tightlist
+\item
+  All results are conditional on the data from NOMAD; further validation
+  on independent datasets is needed
+\item
+  More sophisticated noise structures (e.g., nonparametric σ) could be
+  explored
+\item
+  Spatial or spatio-temporal extensions were not considered, but would
+  be feasible in PyMC
+\end{itemize}
 
@@
 
-\bibitem[\citeproctext]{ref-chipman2010bart}
-Chipman, H. A., George, E. I., \& McCulloch, R. E. (2010). BART:
-Bayesian additive regression trees. \emph{The Annals of Applied
-Statistics}, \emph{4}(1), 266--298.
-\url{https://doi.org/10.1214/09-AOAS285}
+\bibitem[\citeproctext]{ref-abril2023pymc}
+Abril-Pla, O., Andreani, V., Carroll, C., Dong, L., Fonnesbeck, C. J.,
+Kochurov, M., et al. (2023). PyMC: A modern, and comprehensive
+probabilistic programming framework in python. \emph{PeerJ Computer
+Science}, \emph{9}, e1516. \url{https://doi.org/10.7717/peerj-cs.1516}
+
+\bibitem[\citeproctext]{ref-baker2016}
+Baker, M. (2016). 1,500 scientists lift the lid on reproducibility.
+\emph{Nature}, \emph{533}(7604), 452--454.
+\url{https://doi.org/10.1038/533452a}
+
+\bibitem[\citeproctext]{ref-bishop2006pattern}
+Bishop, C. M. (2006). \emph{Pattern recognition and machine learning}.
+Springer.
+
+\bibitem[\citeproctext]{ref-clayton2022bernoulli}
+Clayton, A. (2022). \emph{Bernoulli's fallacy: Statistical illogic and
+the crisis of modern science}. Columbia University Press. Retrieved from
+\url{https://books.google.com/books?id=BT4CzwEACAAJ}
+
+\bibitem[\citeproctext]{ref-cobey2024biomedical}
+Cobey, K. D., Ebrahimzadeh, S., Page, M. J., Thibault, R. T., Nguyen,
+P.-Y., Abu-Dalfa, F., \& Moher, D. (2024). Biomedical researchers'
+perspectives on the reproducibility of research. \emph{PLoS Biology},
+\emph{22}(11), e3002870.
 
@@
 https://doi.org/\url{https://doi.org/10.1029/2019JC014941}
+
+\bibitem[\citeproctext]{ref-jaynes2003probability}
+Jaynes, E. T., \& Bretthorst, G. L. (2003). \emph{Probability theory:
+The logic of science}. Cambridge University Press. Retrieved from
+\url{https://books.google.com/books?id=tTN4HuUNXjgC}
 
@@
 
+\bibitem[\citeproctext]{ref-reback2020pandas}
+team, T. pandas development. (2020, February). Pandas-dev/pandas: pandas
+(Version latest). Zenodo. \url{https://doi.org/10.5281/zenodo.3509134}
+
+\bibitem[\citeproctext]{ref-vehtari2017}
+Vehtari, A., Gelman, A., \& Gabry, J. (2017). Practical bayesian model
+evaluation using leave-one-out cross-validation and WAIC.
+\emph{Statistics and Computing}, \emph{27}(5), 1413--1432.
+\url{https://doi.org/10.1007/s11222-016-9696-4}
+
+\bibitem[\citeproctext]{ref-werdell2005improved}
+Werdell, P. J., \& Bailey, S. W. (2005). An improved bio-optical data
+set for ocean color algorithm development and satellite data product
+validation. \emph{Remote Sensing of Environment}, \emph{98}(1),
+122--140.
+
 \end{CSLReferences}

### 2025-08-02 manuscript/main.tex
updated overleaf ms

\documentclass[preprint,authoryear]{elsarticle}

% --------------------------------------------------
% Packages
% --------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{lineno} % for line numbers (optional)
\usepackage{color}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{siunitx} % for SI units
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{natbib}

% --------------------------------------------------
% Line numbers and spacing (remove before final version if needed)
% --------------------------------------------------
% \linenumbers
\doublespacing

% --------------------------------------------------
% Journal info
% --------------------------------------------------
\journal{Remote Sensing of Environment}

% --------------------------------------------------
% Document Start
% --------------------------------------------------
\begin{document}

\begin{frontmatter}

% \title{Shifting Paradigms: A Complete Bayesian Workflow for Ocean Color Remote Sensing}
\title{Redefining Uncertainty: A Complete Bayesian Workflow for Ocean Color Remote Sensing}

\author[inst1]{Erdem M. Karaköylü}
\ead{erdemk@protonmail.com}

\affiliation[inst1]{organization={Independent Consultant}, 
            city={University Park},
            postcode={20782}, 
            country={USA}}

% --------------------------------------------------
% Abstract
% --------------------------------------------------
\begin{abstract}
[Your abstract goes here.]
\end{abstract}

\begin{keyword}
Ocean Color \sep Bayesian Workflow \sep Chlorophyll Model \sep Bayesian modeling 
\end{keyword}

\end{frontmatter}

% --------------------------------------------------
% Main Body
% --------------------------------------------------

\section{Introduction}

Satellite ocean color remote sensing has long served as a cornerstone of marine ecosystem monitoring, offering global and synoptic coverage of surface ocean properties. Among these, chlorophyll-a ($Chl_a$) concentration remains a central quantity, widely used as a proxy for phytoplankton biomass, primary production and water quality. The retrieval of $Chl_a$ from ocean color data has evolved over decades, resulting in a diverse lineage of empirical and semi-empirical algorithms. The following section summarizes this historical development, which sets the stage for a critical examination of the statistical foundations underlying current approaches.

\subsection{Background}

Early empirical algorithms, notably the $OCx$ family developed by O’Reilly et al. \citep{oreilly1998, oreilly2000}, established a statistical template for retrieving chlorophyll-a ($Chl_a$) concentration from ocean color data. These models relate log-transformed blue-to-green reflectance ratios to in situ $Chl_a$, utilizing either direct band ratios (BR) or maximum band ratios (MBR)—the latter selecting the largest blue-to-green ratio for a given observation and applying a high-order polynomial fit. Their empirical simplicity and practical robustness made these polynomial regressions the operational foundation for chlorophyll products across successive satellite sensors (e.g., CZCS, SeaWiFS, MODIS, MERIS). They proved particularly effective in Case-1 waters, where phytoplankton dominate the optical signal.

However, performance degrades in optically complex Case-2 waters, where non-phytoplankton components (e.g., suspended sediments, colored dissolved organic matter) disrupt the assumed reflectance-$Chl_a$ relationship. These models are also sensitive to atmospheric correction errors, especially in the blue spectral region. Yet, the OCx family represents a deterministic and frequentist modeling tradition, yielding single-point predictions with fixed coefficients and offering no formal quantification of parameter or predictive uncertainty.

Subsequent refinements have addressed these limitations. For example, the Color Index (CI) method \citep{hu2012} employs a band-difference approach to reduce sensitivity to sensor noise and atmospheric residuals. Ongoing efforts have led to newer algorithm variants such as OC5 and OC6 \citep{oreilly2019}, which incorporate additional bands or modified ratio formulations to better capture variability across chlorophyll regimes.

\subsection{Limitations of Existing Approaches}

Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error - one that pervades much of observational science: the conflation of sampling probability with inferential probability \citep{jaynes2003probability,descheemaekere2011}.

Consider a data set $D$ composed of input-output pairs, for example remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$) - and a model $M$, representing the relationship between them. The sampling probability $p(D|M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true. Standard model fitting whatever the specific form, is akin to maximizing this likelihood by adjusting parameters of $M$ to best explain the observed data.

This approach tacitly assumes that the model that best fits the training data also most accurately represents the underlying generative process. This constitutes an epistemic fallacy; treating the sampling probability $p(D|M)$ as though its conditionality was reversed; i.e., $p(M|D)$. Although in well-behaved data-rich cases, the maxima of $p(D|M)$ and $p(M|D)$ may coincide, this remains the exception—not the rule.

This mistake lies at the heart of what \cite{clayton2022bernoulli} terms Bernoulli's Fallacy: the widespread misinterpretation of sampling probability as inference probability. As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.

This fallacy contributes to poor model generalization, drives the use of ad hoc, retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate \cite{baker2016, cobey2024biomedical}. These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.

In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.

This epistemic shortcut has been directly critiqued in the machine learning literature. \cite{gal2016uncertainty} and \cite{ghahramani2015probabilistic} point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference. The result is overconfident predictions and brittle generalization, in line with Clayton’s critique. \cite{bishop2006pattern} similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient.

\subsection{Overcoming limitations}

In oceanographic remote sensing, several recent efforts have attempted to address the limitations of classical models. For instance, \cite{seegers2018} proposed alternative evaluation metrics to move beyond restrictive frequentist assumptions. Others have introduced Bayesian elements into the modeling pipeline: \cite{frouin2013} applied Bayesian inversion for atmospheric correction, \cite{shi2015} used probabilistic fusion for multi-sensor data, and \cite{craig2019} employed Hamiltonian Monte Carlo to train Bayesian neural networks (BNNs) for retrieving inherent optical properties (IOPs) from top-of-atmosphere radiance. Similarly, \cite{werther2022} used Monte Carlo dropout to approximate Bayesian inference in IOP retrieval, while \cite{werther2025} benchmarked multiple probabilistic neural network architectures for quantifying aleatoric and epistemic uncertainty. \cite{erickson2023} recast the Generalized Inherent Optical Property (GIOP) framework using conjugate Bayesian linear models.

These and other studies mark important progress. However, as \cite{werther2023} correctly argues, embracing uncertainty requires more than scatter plots with error bars. Despite growing interest in Bayesian tools for ocean color modeling, the full Bayesian workflow \citep{gelman2020bayesianworkflow, wolkovich2024fourstepbayesianworkflowimproving} remains underutilized. One exception is Craig and Karaköylü \citep{craig2019}, who explicitly define uncertainty in Bayesian terms—as distributions over unknowns—and construct models that natively express posterior uncertainty. Even there, however, evaluation still relies on frequentist metrics such as \(R^2\) and $MAE$. More commonly, probabilistic components are bolted onto otherwise frequentist pipelines, leaving the foundational definition of uncertainty untouched.

In the ocean color community, uncertainty is most commonly defined following the GUM convention \citep{gum2008} as a parameter characterizing the dispersion of values that could reasonably be attributed to the measurand \citep{ioccg2020uncertainties}. This frequentist view emphasizes measurement variability, typically expressed through standard deviations or confidence intervals, based on assumptions about repeated sampling and error propagation.

By contrast, in Bayesian inference all unknowns - such as model parameters - are assigned probability distributions that are updated through observed data. This view naturally encompasses both aleatoric and epistemic uncertainty, and it does not treat interval estimates as fixed or convention-bound. A Bayesian credible interval represents a direct probabilistic statement — e.g., “there is an 87\% probability that the true value lies in this interval, given the data and model” — and its width can be tailored to suit scientific or decision-making relevance.

Bridging this gap requires more than Bayesian components; it requires Bayesian thinking. In what follows, I recast a foundational chlorophyll retrieval model as a fully Bayesian workflow, where uncertainty is not an afterthought but a central object of inference; an object that quantifies what the practitioner does not know about the underlying data-generating process.

\section{Materials and Methods}

\subsection{Dataset and Preprocessing}

I used the familiar NASA Ocean Biology Processing Group's NOMAD dataset \citep{werdell2005improved}, which contains quality-controlled in situ chlorophyll-a measurements matched with satellite-derived remote sensing reflectance (Rrs) observations. The dataset spans a wide range of oceanographic conditions, enabling model development with potential for broad generalization.

I retained visible Rrs bands from SeaWiFS (411, 443, 489, 510, 555, and 670 nm) and removed observations with invalid (null, zero, or negative) values. Chlorophyll-a concentrations measured using HPLC or fluorescence were $log_{10}$-transformed to stabilize variance.

To construct the primary predictor, I computed the Maximum Band Ratio (MBR), similar to OC6 proposed by \cite{oreilly2019}, which uses all 6 available bands:

$$
MBR = \frac{\max(Rrs_{411}, Rrs_{443}, Rrs_{489}, Rrs_{510})}{Rrs_{555} + Rrs_{670}}
$$

I then log-transformed MBR and created a categorical grouping variable based on which numerator band was dominant for each observation. An additional binary flag distinguished chlorophyll measurement method (HPLC or fluorescence).

\subsection{Modeling Approach}

I developed a sequence of Bayesian regression models to estimate log-transformed chlorophyll-a from log-transformed MBR. Models were implemented using PyMC v5 \citep{abril2023pymc} and fitted via the No-U-Turn Sampler - NUTS, \citep{homan2014} - an adaptive variant of Hamiltonian Monte Carlo. I assessed convergence using Gelman-Rubin's $\hat{R}$ diagnostic, effective sample size (ESS), and visual trace plot inspection \cite{mcelreath2020}.
$\hat{R}$ measures how well independent MCMC chains converge; a value of 1.00-1.01 is indicative of good convergence. ESS estimates how much independent information is available for each model parameter in potentially autocorrelated MCMC chains; ESS greater than 1000 suggests stable estimates. Both metrics are suggestive, and unmet criteria are not necessarily deal breakers but call for heightened vigilance.

Once trained each model's predictive skill was evaluated using several approaches. The first uses Posterior Predictive Checks (PPC). Similar to prior predictive checks, this compares training data to trained model simulated output.

Models were then evaluated for generalization (how well they can predict on out-of-sample data) and uncertainty calibration (how realistic prediction uncertainties are). For this I used the Leave-One-Out cross validation (LOO) Probability Integral Transform (PIT); \textit{cf. e.g.} \cite{nguyen2025}. The Probability Integral Transform states that if a random variable $X$ has a continuous distribution with cumulative distribution function (CDF) $F_X$, then the random variable $Y = F_X(X)$ is uniformly distributed. This property is used in LOO-PIT to check if the model's predictions are well-calibrated. In LOO-PIT, the model is fit on all data except one observation, and the CDF of the posterior predictive distribution is used to compute the PIT for each observation. If the model is well-calibrated, the LOO-PIT values should be uniformly distributed. To reduce computational complexity, LOO does not require refitting the model for each data point. Instead, it uses a Pareto Smoothed Importance Sampling (PSIS) approximation, which re-weights draws from the full posterior to approximate leave-one-out posteriors in a stable and efficient manner \citep{vehtari2017}.
The Leave-One-Out (LOO) Probability Integral Transform (PIT) is a method used in Bayesian statistics for model checking. It combines the concepts of Leave-One-Out Cross-Validation and Probability Integral Transform to assess the calibration of a model's predictions (see \citealt{karakoylublog} for a brief tutorial). The LOO-PIT method helps identify model misspecifications by comparing the distribution of LOO-PIT values to a uniform distribution. If the model is not well-calibrated, the LOO-PIT values will deviate from uniformity.

PSIS-LOO also yields a scalar metric for model comparison: the expected log predictive density (ELPD, \citealt{vehtari2017}). This quantity summarizes how well a model is expected to predict new data. Models with higher ELPD values are preferred. In practice, I compared models by their ELPD and corresponding standard errors, and considered differences significant when the ELPD gap exceeded its uncertainty. In combination with the previously described analysis steps, this provides a robust, fully Bayesian approach to selecting among competing models.

Finally, I computed predictive interval coverage: the proportion of observed values falling within the 94\% highest density intervals (HDI - the narrowest possible interval)  of the posterior predictive distribution. This was done for both in-sample  and out-of-sample datasets to assess generalization. A SeaBASS matchup dataset, containing 53 clean and complete observations after preprocessing, was used for out-of-sample-validation.

\subsubsection{Model Progression}

Modeling proceeded iteratively. Initial model performance and diagnostics motivated refinement of structure and assumptions to direct subsequent model development. Table ~\ref{tab:model-overview} summarizes the three main models discussed in the main text.

\begin{table}[H]
\centering
\caption{Summary of model progression. Each model incrementally addresses limitations in structure or variance.}
\label{tab:model-overview}
\begin{tabular}{llp{7.5cm}}
\toprule
\textbf{Model} & \textbf{Type} & \textbf{Key Characteristics} \\
\midrule
Model 1 & Polynomial Regression & Bayesian re-framing of OCx (OC6-style), with a 4th-order polynomial on log(MBR) predicting log(Chl) \\
Model 2 & Hierarchical Linear Regression & Partial pooling across MBR numerator groups, each with its own intercept and slope \\
Model 5 & Heteroskedastic HLR & Extends Model 2 by modeling log($\sigma$) as a linear function of log(MBR), with group-specific parameters \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Model Structures and Priors}

Model 1 uses a single-level polynomial structure with Gaussian priors on all regression coefficients and a Gamma prior on the shared dispersion parameter $\sigma$. Model 2 introduces a hierarchical structure with group-specific slopes and intercepts based on the dominant MBR numerator band. Unlike the 4th-order polynomial used in Model 1, this model uses a simple linear form and shares information across groups through partial pooling. This added structure is expected to capture spectral variability more effectively and yield improved calibration. Model 5 adds heteroskedasticity by modeling $log(\sigma_i)$ as a linear function of log(MBR), with group-specific slope and intercepts.

All group-level parameters were given Normal priors centered at zero, and hyper-prior standard deviations followed Exponential(1) distributions to allow regularization via partial pooling. Prior predictive checks were performed for all models to ensure reasonable behavior before fitting.

Figure ~\ref{fig:model1-struct} shows the structure of Model 1 as a directed acyclic graph (DAG). DAGs for Models 2 and 5 are provided in the Supplementary Material.

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{figures/model1_structure.pdf}
\caption{Model 1 Directed Acyclic Graph (DAG). Gaussian priors are used for the intercept $\alpha$ and polynomial coefficients $\beta_1$--$\beta_4$. The dispersion parameter $\sigma$ has a Gamma prior. Observations are modeled via a truncated normal distribution.}
\label{fig:model1-struct}
\end{figure}

\section{Results}

\subsection{Model Performance Overview}

Three models were fit to the data, each extending the previous in structural complexity. Model 1, a Bayesian analogue of OCx-style polynomial regression, served as a baseline. Model 2 introduced group-specific structure via hierarchical partial pooling by dominant MBR numerator band. Model 5 extended Model 2 by introducing heteroskedasticity, allowing the dispersion parameter to vary with log(MBR) across groups.

\subsection{Model Evaluation: Prior Predictive, Posterior Predictive, and LOO-PIT}

Standard evaluation metrics such as $R^2$ or RMSE assess how closely point predictions match observed data. However, these are not appropriate for Bayesian models, which yield full posterior distributions rather than single-value estimates. In this context, predictive accuracy must be judged not only by central tendency but also by how well the model represents uncertainty.

To that end, I evaluate each model using prior and posterior predictive checks and LOO-PIT diagnostics, which respectively assess in-sample fit and out-of-sample calibration. These diagnostics are presented below for Models 1, 2, and 5, beginning with Model 1.

\subsubsection{Model 1: Polynomial Regression}

The top two panels in Figure~\ref{fig:model1-eval} display kernel density estimates (KDEs) comparing simulated and observed distributions of log-transformed chlorophyll-a. KDEs provide smooth, non-parametric estimates of probability density and are used here to visually assess how well the model captures the shape of the data-generating process.

The top-left panel shows the prior predictive distribution. Simulated draws from the model under the prior (gray lines) span a wide range of values, as expected given the weakly informative priors. The mean of these simulations (orange dashed line) is diffuse and relatively flat, indicating that the prior allows the model to generate a wide range of plausible outcomes. The observed data KDE (black line) does not influence the prior predictive draws, as the model has not yet been trained on the data; it is included here solely to facilitate comparison with the posterior predictive check in the top-right panel.

The top-right panel compares the posterior predictive distribution to the same observed KDE. The black line represents the empirical distribution of the data, estimated from the NOMAD dataset. It is unimodal, with a dominant peak near log(Chl) = 0 (i.e., 1 mg~m$^{-3}$), typical of mesotrophic or moderately productive coastal waters. A secondary shoulder appears near log(Chl) $\approx$ -0.5, corresponding to lower chlorophyll concentrations (0.3 mg~m$^{-3}$ and below) associated with oligotrophic oceanic regions. This structure indicates that the dataset includes a range of ecological regimes, with a skew toward oceanic conditions.

Posterior predictive draws (gray lines) cluster tightly around the observed KDE and differ markedly from the diffuse prior predictive draws, suggesting that the model has learned meaningfully from the data. The posterior predictive mean (orange dashed line) successfully captures the overall support and central tendency of the observed distribution. However, it underestimates the height of the primary mode, producing a smoother, more flattened density. This suggests that while the model has fit the broad structure of the data, it fails to fully represent the concentrated central mass, indicating a tendency toward overdispersed predictions.

The bottom two panels assess model calibration using leave-one-out probability integral transform (LOO-PIT) diagnostics. Two complementary views are shown: the left panel presents the density of LOO-PIT values via a kernel density estimate (KDE), while the right panel shows the cumulative deviation of those values from the ideal uniform distribution. The KDE is sensitive to local concentration of probability mass—such as over- or underdispersion—while the ECDF-minus-uniform plot emphasizes systematic calibration drift across the distribution. Together, they provide a more complete picture of model uncertainty than either view alone.

In a well-calibrated model, PIT values—computed by evaluating the posterior predictive cumulative distribution function at each held-out observation—should follow a uniform distribution. The KDE panel (bottom-left) shows the LOO-PIT density (black line) compared against 500 KDEs of values drawn from a uniform distribution (thin blue lines). A horizontal dashed line at $y = 1$ denotes the target density for a calibrated model. The observed KDE exhibits a distinct U-shape, with density elevated in the tails and suppressed in the center, indicating that the model is \textbf{underconfident}, or equivalently, \textbf{overdispersed}—its predictive intervals are too wide and assign insufficient mass near the typical observations.

The bottom-right panel shows the difference between the empirical cumulative distribution function (ECDF) of the LOO-PIT values and the ideal uniform ECDF. A horizontal dashed line at $y = 0$ represents the calibration target. The shaded band shows a 94\% credible interval for this difference under ideal conditions, based on repeated uniform draws. The curve for Model 1 deviates below the reference line in the central quantiles and rises above it in the tails, confirming the pattern seen in the KDE panel. Together, these diagnostics show that while Model 1 captures the central trend of the data, its posterior predictive distributions are too diffuse.

\begin{figure}[H]
\centering
\includegraphics[width=0.99\textwidth]{figures/model1_evaluation.pdf}
\caption{Model 1: Top row shows prior and posterior predictive distributions. Bottom row shows model calibration assessed using LOO-PIT plots. Left panel shows the kernel density estimate (KDE) of the LOO-PIT values. For a well-calibrated model, the distribution should be approximately flat, indicating that observed outcomes are consistent with the posterior predictive distributions. Deviations from flatness, such as peaks or dips, reflect under- or overconfidence in different parts of the predictive distribution. The right panel shows the difference between the empirical cumulative distribution function (ECDF) of the LOO-PIT values and the ideal uniform CDF }
\label{fig:model1-eval}
\end{figure}

\subsubsection{Model 2: Hierarchical Linear Regression}

Figure~\ref{fig:model2-eval} shows the model evaluation panels for Model 2. The prior predictive simulations (top-left) are broader than in Model 1, reflecting the additional flexibility introduced by the hierarchical structure. The prior predictive mean (orange dashed) is flat and diffuse, suggesting weakly informative priors that allow a wide range of generative behaviors.

The posterior predictive check (top-right) shows a notable improvement over Model 1. Posterior draws cluster tightly around the observed KDE, especially near the primary mode. This reflects the effect of partial pooling across MBR groups, which concentrates predictions around dominant patterns in the data. However, the model fails to fully capture the secondary shoulder near $\text{log(Chl)} \approx -0.5$, suggesting that some variation remains unaccounted for — possibly due to shrinkage of uncertainty toward the group mean.

The LOO-PIT panels are interpreted using the same diagnostic framework described for Model 1. The KDE reveals local deviations from calibration, while the ECDF-minus-uniform plot captures cumulative miscalibration trends. Here, the LOO-PIT diagnostics (bottom row) further support the improved performance. The KDE panel (bottom-left) shows smaller deviations from the uniform reference (dashed line at $y = 1$) compared to Model 1. The y-axis range here is narrower—approximately 0.5 to 1.3—indicating that the magnitude of miscalibration is significantly reduced. Similarly, the ECDF-minus-uniform plot (bottom-right) shows deviations that remain well within the 94\% credible interval, with a y-axis span of just ±0.025. These compressed ranges reflect a more calibrated model. Mild signs of overdispersion persist—visible as a small negative dip in the central quantiles—but overall, Model 2 represents a substantial improvement in both fit and uncertainty calibration relative to the baseline.

\begin{figure}[H]
\centering
\includegraphics[width=0.99\textwidth]{figures/model2_evaluation.pdf}
\caption{Model 2: Prior/posterior predictive checks (top), LOO-PIT diagnostics (bottom).}
\label{fig:model2-eval}
\end{figure}

\subsubsection{Model 5: Heteroskedastic Hierarchical Linear Regression}

Model 5 extends the hierarchical structure of Model 2 by allowing the likelihood variance $\sigma$ to vary linearly with log(MBR), using group-specific intercepts and slopes. This accommodates heteroskedasticity across spectral groups and better reflects the variance structure of the data. By explicitly modeling the dispersion, the model is expected to improve both fit and calibration, particularly in the tails of the predictive distribution.

Figure~\ref{fig:model5-eval} presents the four-panel evaluation for Model 5. The prior predictive simulations (top-left) are the most diffuse of any model—expected, given the added flexibility in both the mean and variance components. The prior predictive mean (orange dashed line) is nearly flat. As before, the observed KDE (black line) is included only for visual comparison. The posterior predictive check (top-right) shows excellent agreement between the posterior draws and the observed KDE. The model captures both the main mode near $\text{log(Chl)} = 0$ and the secondary shoulder near -0.5—something neither Model 1 nor 2 accomplished. The predictive mean aligns closely with the observed distribution, and the vertical spread among draws reflects uncertainty in both the mean and variance terms.

The LOO-PIT panels are interpreted using the same diagnostic framework described for Model 1. The KDE panel (bottom-left) shows a nearly flat density centered around the ideal reference line at $ y = 1$ , with minor undulations well within the light blue band representing draws from the uniform distribution. The vertical range is now the tightest of all three models—approximately 0.8 to 1.2—indicating minimal deviation from ideal calibration. The ECDF-minus-uniform plot (bottom-right) shows similarly tight behavior. The LOO-PIT curve remains entirely within the 94\% credible interval and deviates by no more than ±0.02 from the reference line at $ y = 0 $. Together, these diagnostics suggest that Model 5 is both well-calibrated and well-aligned with the observed data, representing the most accurate and epistemically sound predictive model of the three.

\begin{figure}[H]
\centering
\includegraphics[width=0.99\textwidth]{figures/model5_evaluation.pdf}
\caption{Model 5: Prior/posterior predictive checks (top), LOO-PIT diagnostics (bottom).}
\label{fig:model5-eval}
\end{figure}

\subsection{Predictive Coverage}

For each model, I computed the proportion of observed values falling within the 94\% highest density intervals (HDIs) of the posterior predictive distribution. Coverage was assessed separately for in-sample (NOMAD) and out-of-sample (SeaBASS) datasets.

\subsubsection{Model 1}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{figures/model1_predictive_coverage.pdf}
\caption{Model 1: Predictive coverage plots. Left: in-sample (NOMAD). Right: out-of-sample (SeaBASS).}
\label{fig:model1-coverage}
\end{figure}

\subsubsection{Model 2}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{figures/model2_predictive_coverage.pdf}
\caption{Model 2: Predictive coverage plots. Left: in-sample. Right: out-of-sample.}
\label{fig:model2-coverage}
\end{figure}

\subsubsection{Model 5}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{figures/model5_predictive_coverage.pdf}
\caption{Model 5: Predictive coverage plots. Left: in-sample. Right: out-of-sample.}
\label{fig:model5-coverage}
\end{figure}

Overall, each increase in model complexity yielded improvements in calibration and generalizability, as indicated by posterior predictive checks, leave-one-out probability integral transform (LOO-PIT) diagnostics, and predictive interval coverage. Model 5 exhibited the best out-of-sample calibration and predictive accuracy.

\subsection{Model Comparison}

See Table~\ref{tab:model-comparison} for ELPD-based comparison and Figure~\ref{fig:model-comparison-plot} for a visual summary.

\begin{table}[H]
\centering
\caption{Model comparison using PSIS-LOO. Higher ELPD indicates better expected out-of-sample predictive performance. $\Delta$ELPD shows differences relative to best model. $p_{\text{loo}}$ is the effective number of parameters.}
\label{tab:model-comparison}
\begin{tabular}{lrrr}
\toprule
Model & ELPD & $\Delta$ELPD & $p_{\text{loo}}$ \\
\midrule
Model 5 & -- & -- & -- \\
Model 2 & -- & -- & -- \\
Model 1 & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=0.75\textwidth]{figures/elpd_comparison_plot.pdf}
\caption{Model comparison plot showing $\Delta$ELPD and $p_{\text{loo}}$ across models.}
\label{fig:model-comparison-plot}
\end{figure}

\section{Discussion}
The posterior predictive distribution is an information-rich construct that reflects both model structure and epistemic (and aleatoric) uncertainty. Rather than measuring residual error around means, I evaluate models based on their ability to generate data consistent with observations, and on how well-calibrated their uncertainty estimates are. This is done through prior and posterior predictive checks for in-sample fit, and LOO-PIT diagnostics for out-of-sample calibration. These tools provide a more appropriate basis for evaluating Bayesian models than traditional metrics designed for frequentist or deterministic frameworks.

\section*{Declaration of Generative AI and AI-Assisted Technologies in the Writing Process}
During the preparation of this work, the author used ChatGPT to help polish the language of the Methods and Results sections. After using this tool, the authors reviewed and edited the content as needed and assume full responsibility for the content of the publication.

\bibliographystyle{elsarticle-harv}
\bibliography{references}  % Create refs.bib file or replace with manual entries

\end{document}
